---
title: "Drunk Driving Surveillance Dashboard"
format:
  dashboard:
    theme: default
    css: styles3.css
    orientation: rows
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
---

Welcome to the Drunk Driving Surveillance Dashboard!

This dashboard displays data on alcohol-related vehicle crashes in El Paso, Texas and is updated every month. 


```{r}

rm(list = ls())

suppressMessages(library(reactable))
suppressMessages(library(DT))
suppressMessages(library(gt))
suppressMessages(library(quarto))
suppressMessages(library(rsconnect))
suppressMessages(library(shiny))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
suppressMessages(library(stringr))
suppressMessages(library(readr))
suppressMessages(library(tidyverse))
suppressMessages(library(leaflet))
suppressMessages(library(leaflet.extras))


data <- read.csv("C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/elpaso_workingdata.csv", header = TRUE)


data$City <- recode(data$City,
  "EL PASO" = "El Paso"
)

# drop current month, since its incomplete
data <- data %>%
  mutate(
    Crash.Date = mdy(Crash.Date)
  ) %>%
  filter(
    !(year(Crash.Date) == year(Sys.Date()) &
      month(Crash.Date) == month(Sys.Date()))
  )

person_level <- data

vehicle_level <- data %>%
  distinct(Crash.ID, Unit.Number, .keep_all = TRUE)

crash_level <- data %>%
  distinct(Crash.ID, .keep_all = TRUE) 
```

# Monthly Brief

```{r}
# ----------------------------
# REPORTING MONTH (DATA LAGS 1 MONTH)
# ----------------------------
report_date <- Sys.Date() %m-% months(1)
report_month <- format(report_date, "%B %Y")
report_year <- year(report_date)
prev_year <- report_year - 1

# ----------------------------
# MONTHLY DATA
# ----------------------------
this_month <- crash_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == report_year,
    month(Crash.Date) == month(report_date)
  )

last_month <- crash_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == prev_year,
    month(Crash.Date) == month(report_date)
  )

# ----------------------------
# METRICS
# ----------------------------
crashes_now <- n_distinct(this_month$Crash.ID)
crashes_prev <- n_distinct(last_month$Crash.ID)

# ----------------------------
# INJURIES (MATCH YOUR CLEAN LOGIC)
# ----------------------------
injuries_now <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == report_year,
    month(Crash.Date) == month(report_date),
    !Person.Injury.Severity %in% c(
      "99 - UNKNOWN",
      "95 - AUTONOMOUS",
      "N - NOT INJURED",
      "K - FATAL INJURY"
    ),
    !is.na(Person.Injury.Severity)
  ) %>%
  summarise(n = n()) %>%
  pull(n)

injuries_prev <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == prev_year,
    month(Crash.Date) == month(report_date),
    !Person.Injury.Severity %in% c(
      "99 - UNKNOWN",
      "95 - AUTONOMOUS",
      "N - NOT INJURED",
      "K - FATAL INJURY"
    ),
    !is.na(Person.Injury.Severity)
  ) %>%
  summarise(n = n()) %>%
  pull(n)

# ----------------------------
# FATALITIES (KEEP CONSISTENT)
# ----------------------------
fatal_now <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == report_year,
    month(Crash.Date) == month(report_date),
    Person.Injury.Severity == "K - FATAL INJURY"
  ) %>%
  summarise(n = n()) %>%
  pull(n)

fatal_prev <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == prev_year,
    month(Crash.Date) == month(report_date),
    Person.Injury.Severity == "K - FATAL INJURY"
  ) %>%
  summarise(n = n()) %>%
  pull(n)

# % changes
pct_change_crash <- ifelse(crashes_prev == 0, NA, (crashes_now - crashes_prev)/crashes_prev*100)
pct_change_injury <- ifelse(injuries_prev == 0, NA, (injuries_now - injuries_prev)/injuries_prev*100)
pct_change_fatal <- ifelse(fatal_prev == 0, NA, (fatal_now - fatal_prev)/fatal_prev*100)

fmt <- function(x) ifelse(is.na(x), "N/A", paste0(round(x,1), "%"))

# ----------------------------
# MONTHLY SERIES (CURRENT VS PRIOR YEAR)
# ----------------------------
build_monthly <- function(df, value){
  df %>%
    mutate(
      year = year(Crash.Date),
      month = month(Crash.Date)
    ) %>%
    filter(year %in% c(report_year, prev_year)) %>%
    group_by(year, month) %>%
    summarise(val = {{value}}, .groups="drop") %>%
    complete(year, month = 1:12, fill = list(val = 0)) %>%
    mutate(
      month_lab = factor(month.abb[month], levels = month.abb),
      year = factor(year)
    )
}

# crashes
crash_plot_data <- build_monthly(crash_level %>% filter(City=="El Paso"),
                                n_distinct(Crash.ID))

# fatalities
fatal_plot_data <- build_monthly(
  person_level %>%
    filter(City=="El Paso",
           Person.Injury.Severity == "K - FATAL INJURY"),
  n()
)

# injuries
injury_plot_data <- build_monthly(
  person_level %>%
    filter(
      City=="El Paso",
      !Person.Injury.Severity %in% c(
        "99 - UNKNOWN",
        "95 - AUTONOMOUS",
        "N - NOT INJURED",
        "K - FATAL INJURY"
      ),
      !is.na(Person.Injury.Severity)
    ),
  n()
)

# ----------------------------
# DIRECTION LABELS
# ----------------------------
dir_crash <- case_when(
  is.na(pct_change_crash) ~ "",
  pct_change_crash > 0 ~ "increase",
  pct_change_crash < 0 ~ "decrease",
  TRUE ~ "no change"
)

dir_fatal <- case_when(
  is.na(pct_change_fatal) ~ "",
  pct_change_fatal > 0 ~ "increase",
  pct_change_fatal < 0 ~ "decrease",
  TRUE ~ "no change"
)

dir_injury <- case_when(
  is.na(pct_change_injury) ~ "",
  pct_change_injury > 0 ~ "increase",
  pct_change_injury < 0 ~ "decrease",
  TRUE ~ "no change"
)

```

## Row {.tabset height=35%}

``` {r}
# remove future months for current year only
filter_future <- function(df){
  df %>%
    filter(
      !(year == report_year & month > month(report_date))
    )
}

crash_plot_data  <- filter_future(crash_plot_data)
fatal_plot_data  <- filter_future(fatal_plot_data)
injury_plot_data <- filter_future(injury_plot_data)

# comparison months for brief
comparison_month <- format(
  as.Date(paste0(prev_year, "-", month(report_date), "-01")),
  "%B %Y"
)
```

``` {r}
#| title: Crashes

p1 <- ggplot(crash_plot_data,
             aes(x = month_lab, y = val, fill = year)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = val, y = val + 1),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, max(crash_plot_data$val, na.rm = TRUE) * 1.1)
  ) +
  labs(title = "Total Alcohol-Related Crashes",
       x = "", y = "Total crashes", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p1) 
```

``` {r}
#| title: Injuries

p3 <- ggplot(injury_plot_data,
             aes(x = month_lab, y = val, fill = year)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = val, y = val + 1),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, max(injury_plot_data$val, na.rm = TRUE) * 1.1)
  ) +
  labs(title = "Total People Injured in Alcohol-Related Crashes",
       x = "", y = "Total people injured", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p3) 
```

``` {r}
#| title: Fatalities

p2 <- ggplot(fatal_plot_data,
             aes(x = month_lab, y = val, fill = year)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = val, y = val * 1.05),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, max(fatal_plot_data$val, na.rm = TRUE) * 1.1)
  ) +
  labs(title = "Total Alcohol-Related Traffic Fatalities",
       x = "", y = "Total fatalities", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p2) 

```


## Row {height=30%}

### Column 

::: {.card title="**Monthly Summary**"}
In **`r report_month`**, there were:

- **`r crashes_now`** alcohol-related crashes (**`r fmt(pct_change_crash)`** `r dir_crash` vs. `r comparison_month`)  
- **`r injuries_now`** people injured in an alcohol-related crash (**`r fmt(pct_change_injury)`** `r dir_injury` vs. `r comparison_month`)

- **`r fatal_now`** fatalities associated with alcohol-related crashes (**`r fmt(pct_change_fatal)`** `r dir_fatal` vs. `r comparison_month`)

:::

### Column

::: {.card title="**Navigation Links**"}

- [Monthly Brief](#monthly-brief) — Latest monthly summary of crashes, injuries, and fatalities
- [About this Dashboard](#about-this-dashboard) — Data sources, definitions, and technical details 
- [Trends](#trends) — Yearly trend patterns and year-to-date totals
- [Holidays](#holidays) — Crash patterns around major holidays and event periods  
- [Drivers](#drivers) — Driver-level characteristics, risk indicators, and year-to-date totals 
- [Map](#map) — Geographic distribution of alcohol-related crashes  
- [Quiz](#quiz) — How well do you know the data in El Paso? Take the quiz and find out!

:::



# About this Dashboard

## Row {height=50%}

::: {.card title="**About This Dashboard**"}

Last Update: April 2026

The Drunk Driving Surveillance Dashboard provides monthly, city-level data on alcohol-related motor vehicle crashes in El Paso, Texas. It is designed as a public-facing resource for journalists, policymakers, and researchers seeking timely and transparent information on impaired driving trends.

Crash data is downloaded from the Texas Department of Transportation (TXDOT) website using the 
[Crash Records Information System (CRIS) Query Tool](https://cris.dot.state.tx.us/public/Query/app/home). All crash data in this dashboard represent those reported by law enforcement, were classified as alcohol-related, and occurred in the city of El Paso.

<u>Definition of Alcohol-Related Crash</u>

A crash is defined as alcohol-related if involved drivers or individuals were reported as "Under the Influence of Alcohol", "Had Been Drinking", or were reported to have a blood alcohol content (BAC) greater than 0.

<u>Definition of Alcohol-Related Injury/Fatality</u>

Injuries include people who were classified as having a suspected serious injury, suspected minor injury, or possible injury. Fatalities are defined as any motor vehicle traffic injury crash that results in the death of one or more persons within 30 days of the incident.

<u>Definition of Intoxicated Driver</u>

Driver and vehicle characteristics in these graphs represent intoxicated drivers. An intoxicated driver is defined as a person who's BAC test result returned a value of 0.08g/dL or greater.

<u>Acronyms</u>

BAC = Blood alcohol content; YTD = Year to date; TXDOT = Texas Department of Transportation; CRIS = Crash Records Information System

:::

::: {.card title="**For Media**"}

<u>Key Notes for Reporting</u>

- Crashes, injuries, and fatalities in the dashboard represent those that were alcohol-related  
- Trends are based on official crash records from TXDOT
- Monthly updates may revise prior counts as data are finalized  

<u>Using Figures and Charts</u>

All figures in this dashboard are interactive and downloadable.

- To expand a figure, click the “Expand” button located at the bottom-right corner of each chart. Charts can easily be resized by adjusting the size of your internet browsing window.  
- To download a figure, hover over the chart and click the camera icon to save it as an image  
- Charts may be used in media coverage with attribution. *Data Source: TXDOT CRIS Query Tool | Figure created by El Paso Crime Lab*  

<u>How to Cite This Dashboard</u>

When referencing this dashboard, please use the following citation:

El Paso Crime Lab. *Drunk Driving Surveillance Dashboard: El Paso, Texas* (2026). [www.epcrimelab.org/crash_dashboard.html]

<u>Suggested Language</u>

- “Crash data on alcohol-related crashes in El Paso show…”
- “Local data on alcohol-related traffic fatalities indicate…”

<u>Contact</u>

For questions or data clarification, please contact:  
mduran29@asu.edu

:::

::: {.card title="**Technical Details**"}

<u>Calculation of Moving Average</u>

The line graphs on the [Trends](#trends) page show 12-month moving averages for crashes, fatalities, and injuries between 2016 and *Present*. This technique results in a smooth line, allowing the viewer to see if a trend is increasing or decreasing.

<u>Holiday Analyses</u>

[Holidays](#holidays) statistics were calculated by aligning crash data to each holiday. For fixed-date holidays (e.g., New Year’s Day, Independence Day, Halloween, Christmas, and St. Patrick’s Day), each crash was assigned an event time based on the number of days before or after the calendar date of the holiday within each year. For shifting holidays (e.g., Super Bowl and Memorial Day), the holiday date was first computed for each year and then used as the reference point for event-time calculation.

In all cases, crashes were restricted to a ±7-day window around the holiday and then aggregated across years (2016–2025), excluding 2020 to avoid COVID-19-related disruptions. Average daily crash counts were then computed for each relative day to identify typical patterns before and after each holiday.

<u>Missing Data</u>

A small amount of data for [Drivers](#drivers) characteristics (e.g., gender, race/ethnicity, vehicle color) is missing. Observations with missing or unknown values in these fields were excluded.

<u>Limitations</u>

Not all alcohol-involved crashes are identified in official records. In many cases, alcohol use may go undetected or unreported, particularly when testing is not conducted or impairment is not clearly observed. As a result, these data likely underestimate the true number of alcohol-related crashes—a gap often referred to as the “dark figure” of impaired driving.

Despite this limitation, crash data remain one of the most reliable and policy-relevant indicators of drunk driving. Unlike arrest data, which are influenced by enforcement practices, crash records capture real-world harm—injuries and fatalities—providing a consistent basis for tracking trends over time and understanding the public safety impact of impaired driving.

:::

# Trends

## Row {height="200px"}

```{r}
#| content: valuebox
#| title: "Total Crashes (YTD)"
ytd_alcohol_crashes <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  summarise(
    ytd_alcohol_crashes = n_distinct(Crash.ID)
  ) %>%
  pull(ytd_alcohol_crashes)

list(
  icon = "car-front-fill",
  color = "primary",
  value = ytd_alcohol_crashes )
```

```{r}
#| content: valuebox
#| title: "Total People Injured (YTD)"

ytd_people_injury <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0
    )
  )

ytd_people_injury <- ytd_people_injury %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    injured == 1) %>%
    nrow()


list(
  icon = "bandaid",
  color = "primary",
  value = ytd_people_injury)

```

```{r}
#| content: valuebox
#| title: "Total Fatalities (YTD)"
ytd_people_fatal <- person_level %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    )
  )
ytd_alcohol_fatalities <- ytd_people_fatal %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    fatal == 1) %>%
    nrow()

list(
  icon = "hospital",
  color = "danger",
  value = ytd_alcohol_fatalities
)
```

## Row {.tabset height="400px"}

``` {r}
#| title: Crash Trends
monthly_counts <- crash_level %>%
  filter(City == "El Paso") %>%   # 🔴 THIS IS THE KEY FIX
  mutate(
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    crashes = n_distinct(Crash.ID),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    ma_12 = zoo::rollmean(crashes, k = 12, fill = NA, align = "right")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed
  geom_line(
    aes(y = crashes, color = "Total alcohol-involved crashes"),
    linewidth = 0.8
  ) +

  # moving average
  geom_line(
    aes(y = ma_12, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total alcohol-involved crashes" = "grey70",
      "Average" = "black"
    )
  ) +

  labs(
    x = "",
    y = "Total crashes",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p) 
```


``` {r}
#| title: Injury Trends

injury_monthly <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0
    ),
    month = floor_date(Crash.Date, "month")
  ) %>%
  filter(
    City == "El Paso",
    injured == 1   # uses your indicator
  ) %>%
  group_by(month) %>%
  summarise(
    injuries = n(),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    ma_12 = zoo::rollmean(injuries, k = 12, fill = NA, align = "right")
  )

# ----------------------------
# Step 2: plot (MA only)
# ----------------------------
p <- ggplot(injury_monthly, aes(x = month)) +

  # observed
  geom_line(
    aes(y = injuries, color = "Total people injured in alcohol-involved crashes"),
    linewidth = 0.8
  ) +

  # moving average
  geom_line(
    aes(y = ma_12, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total people injured in alcohol-involved crashes" = "grey70",  # yellow
      "Average"        = "#F77F00"   # orange
    )
  ) +

  labs(
    x = "",
    y = "Total people injured",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)

```


``` {r}
#| title: Fatality Trends
monthly_counts <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    ),
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    alcohol_fatal_crashes = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    alcohol_fatal_ma = zoo::rollmean(alcohol_fatal_crashes,
                                     k = 12,
                                     fill = NA,
                                     align = "right")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed total fatalities (grey)
  geom_line(
    aes(y = alcohol_fatal_crashes, color = "Total alcohol-involved fatalities"),
    linewidth = 1
  ) +

  # alcohol fatalities 12-mo moving average (maroon)
  geom_line(
    aes(y = alcohol_fatal_ma, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total alcohol-involved fatalities" = "grey70",
      "Average" = "maroon"
    )
  ) +

  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, 2)
  ) +

  labs(
    x = "",
    y = "Total fatalities",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p) 
```


``` {r}
#| title: Crashes by Month
monthly_year <- crash_level %>%
  filter(City == "El Paso") %>%
  mutate(
    date = as.Date(Crash.Date),
    year = year(date),
    month_num = month(date)
  ) %>%
  group_by(year, month_num) %>%
  summarise(
    monthly_crashes = n_distinct(Crash.ID),
    .groups = "drop"
  )


monthly_avg <- monthly_year %>%
  group_by(month_num) %>%
  summarise(
    avg_crashes = mean(monthly_crashes),
    .groups = "drop"
  ) %>%
  
  mutate(
    pct = avg_crashes / sum(avg_crashes),
    month_name = factor(month.abb[month_num], levels = month.abb)
  )

overall_avg <- mean(monthly_avg$avg_crashes)
# ----------------------------
# Step 2: Average by month-of-year
# ----------------------------
monthly_avg <- monthly_avg %>%
  mutate(
    label = paste0(
      round(avg_crashes, 0),
      "\n(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )

# ----------------------------
# Step 3: overall mean line
# ----------------------------


# ----------------------------
# Step 4: ggplot base
# ----------------------------
p <- ggplot(monthly_avg, aes(
  x = month_name,
  y = avg_crashes,
  text = paste0(
    "Month: ", month_name,
    "<br>Crashes: ", round(avg_crashes, 0),
    "<br>Percent: ", sprintf("%.1f%%", pct * 100)
  )
)) +

  geom_col(fill = "steelblue") +

geom_text(
  aes(y = avg_crashes + 7, label = label),
  size = 3.5
) +
geom_col(fill = "steelblue", width = 0.5) +

  geom_hline(
    aes(yintercept = overall_avg, color = "Average"),
    linetype = "dashed",
    linewidth = 1
  ) +

  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  labs(
    x = "",
    y = "Average crashes"
  ) +

  theme_minimal() +

  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ----------------------------
# Step 5: Plotly conversion
# ----------------------------
ggplotly(p) 
```

``` {r}
#| title: Fatalities by Month
monthly_fatal_year <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    date = Crash.Date,
    year = year(date),
    month_num = month(date),
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    ),
  ) %>%
  group_by(year, month_num) %>%
  summarise(
    monthly_fatalities = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------
# Step 2: Average fatalities by month-of-year
# ----------------------------
monthly_fatal_avg <- monthly_fatal_year %>%
  group_by(month_num) %>%
  summarise(
    avg_fatalities = mean(monthly_fatalities),
    .groups = "drop"
  ) %>%
  mutate(
    month_name = factor(month.abb[month_num], levels = month.abb),
    pct = avg_fatalities / sum(avg_fatalities),
    label = paste0(
      round(avg_fatalities, 1),
      "<br>(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )

# ----------------------------
# Step 3: overall mean line
# ----------------------------
overall_avg <- mean(monthly_fatal_avg$avg_fatalities)

# ----------------------------
# Step 4: ggplot
# ----------------------------
p <- ggplot(monthly_fatal_avg, aes(
  x = month_name,
  y = avg_fatalities,
  text = label
)) +

  geom_col(fill = "steelblue", width = 0.5) +

  geom_text(
    aes(y = avg_fatalities + max(avg_fatalities) * 0.1, label = label),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, max(monthly_fatal_avg$avg_fatalities, na.rm = TRUE) * 1.3)
) +  
  geom_hline(
    aes(yintercept = overall_avg, color = "Average"),
    linetype = "dashed",
    linewidth = 1
  ) +

  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  labs(
    x = "",
    y = "Average fatalities"
  ) +

  theme_minimal(base_size = 12) +

  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ----------------------------
# Step 5: Plotly
# ----------------------------
ggplotly(p) 
```

## Row {height="400px"}

``` {r}
#| title: Crashes by Day of Week (YTD)
dow_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  count(Day.of.Week) %>%
  mutate(
    Day.of.Week = factor(
      tools::toTitleCase(tolower(Day.of.Week)),
      levels = c("Sunday", "Monday", "Tuesday", "Wednesday",
                 "Thursday", "Friday", "Saturday")
    )
  ) %>%
  mutate(
    pct = n / sum(n),
    label = paste0(
      n,
      "<br>(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )

avg_val <- mean(dow_data$n, na.rm = TRUE)

p3 <- ggplot(dow_data, aes(
  x = Day.of.Week,
  y = n,
  text = label
)) +

  geom_col(fill = "steelblue", width = 0.5) +

  geom_text(
    aes(y = n + max(n) * 0.1, label = label),
    size = 3.5
  ) +

  geom_hline(
    aes(yintercept = avg_val, color = "Average"),
    linewidth = 1.2,
    linetype = "dashed"
  ) +

  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +
scale_y_continuous(
  limits = c(0, max(dow_data$n, na.rm = TRUE) * 1.3)
) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +

  labs(
    x = "",
    y = "Total crashes"
  )

ggplotly(p3) 

```

``` {r}
#| title: Crashes by Hour (YTD)
# ----------------------------
# Step 1: Define correct order
# ----------------------------
time_levels <- c(
  "12am – 3am", "3am – 6am", "6am – 9am", "9am – 12pm",
  "12pm – 3pm", "3pm – 6pm", "6pm – 9pm", "9pm – 12am"
)


# ----------------------------
# Step 2: Build hourly data
# ----------------------------
hour_data <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  mutate(
    hour = as.numeric(substr(Hour.of.Day, 1, 2)),

    time_block = case_when(
      hour >= 0  & hour <= 2  ~ "12am – 3am",
      hour >= 3  & hour <= 5  ~ "3am – 6am",
      hour >= 6  & hour <= 8  ~ "6am – 9am",
      hour >= 9  & hour <= 11 ~ "9am – 12pm",
      hour >= 12 & hour <= 14 ~ "12pm – 3pm",
      hour >= 15 & hour <= 17 ~ "3pm – 6pm",
      hour >= 18 & hour <= 20 ~ "6pm – 9pm",
      hour >= 21 & hour <= 23 ~ "9pm – 12am",
      TRUE ~ NA_character_
    )
  ) %>%

  # count crashes per block
  count(time_block) %>%

  # force full structure (adds zeros)
  complete(time_block = time_levels, fill = list(n = 0)) %>%

  # enforce correct order
  mutate(
    time_block = factor(time_block, levels = time_levels),
    pct = n / sum(n),
    label = paste0(
      n,
      "<br>(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )

# ----------------------------
# Step 3: overall mean line
# ----------------------------
avg_block <- mean(hour_data$n, na.rm = TRUE)

# ----------------------------
# Step 4: ggplot
# ----------------------------
p4 <- ggplot(hour_data, aes(
  x = time_block,
  y = n,
  text = label
)) +

  geom_col(fill = "steelblue", width = 0.5) +

  geom_text(
    aes(y = n + max(n) * 0.1, label = label),
    size = 3.5
  ) +

  geom_hline(
    aes(yintercept = avg_block, color = "Average"),
    linetype = "dashed",
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +
scale_y_continuous(
  limits = c(0, max(hour_data$n, na.rm = TRUE) * 1.3)
) +
  labs(
    x = "",
    y = "Total crashes"
  ) +

  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

# ----------------------------
# Step 5: Plotly
# ----------------------------
ggplotly(p4, tooltip = "text") 

```

# Holidays

## Row {height="250px"}

```{r}
#| title: New Year's Day

base <- crash_level %>%
  filter(City == "El Paso", Crash.Year != 2020) %>%
  mutate(
    date = as.Date(Crash.Date),
    year = Crash.Year   # 🔥 FIX: required
  )

# ----------------------------
# 1. Build explicit event window per year
# ----------------------------
ny_window <- base %>%
  distinct(year) %>%
  mutate(
    holiday = as.Date(paste0(year, "-01-01"))
  ) %>%
  tidyr::expand_grid(event_time = -7:7) %>%
  mutate(date = holiday + event_time)

# ----------------------------
# 2. Join crashes to window
# ----------------------------
data <- ny_window %>%
  left_join(
    base %>% select(date, Crash.ID),
    by = "date"
  )

# ----------------------------
# 3. Aggregate
# ----------------------------
plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups = "drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes, na.rm = TRUE), .groups = "drop")

# ----------------------------
# 4. Plot
# ----------------------------
p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.3,
              color = "#2C3E50", size = 1) +
  geom_point(size = 3, color = "#34495E") +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#C0392B", size = 1) +
  scale_x_continuous(breaks = -7:7, limits = c(-7, 7)) +
  labs(x = "Days from New Year's Day", y = "Average Crashes") +
  theme_minimal(base_size = 10)

y_min <- min(plot_data$avg_crashes, na.rm = TRUE)
y_max <- max(plot_data$avg_crashes, na.rm = TRUE)

ggplotly(p) %>%
  layout(xaxis = list(range = c(-7, 7))) %>%
  add_ribbons(
    x = seq(-0.5, 0.5, length.out = 50),
    ymin = y_min, ymax = y_max,
    fillcolor = "rgba(255,0,0,0.08)",
    line = list(width = 0),
    inherit = FALSE,
    showlegend = FALSE
  ) 
```

``` {r}
#| title: Superbowl 

superbowl <- base %>%
  distinct(year) %>%
  mutate(
    feb1 = as.Date(paste0(year, "-02-01")),
    holiday = feb1 + (7 - wday(feb1)) + 7
  )

data <- base %>%
  left_join(superbowl, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Superbowl", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(0 - 0.5, 0 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(-7 - 0.5, -6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(7 - 0.5, 7 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))  
```

## Row {height="250px"}

```{r}
#| title: St. Patrick's Day

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-03-17")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Saint Patrick's Day", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-4 - 0.5, -4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(-1 - 0.5, -1 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(1 - 0.5, 1 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(4 - 0.5, 4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))  
```

``` {r}
#| title: Memorial Day

memorial <- base %>%
  distinct(year) %>%
  mutate(
    memorial_day = as.Date(paste0(year, "-05-31")) -
      (wday(as.Date(paste0(year, "-05-31")), week_start = 1) - 1)
  )

data <- base %>%
  left_join(memorial, by="year") %>%
  mutate(event_time = as.numeric(date - memorial_day)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Memorial Day", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-2 - 0.5, -2 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(5 - 0.5, 5 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))  
```

## Row {height="250px"}


``` {r}
#| title: Cinco de Mayo

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-05-05")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from May 5", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-6 - 0.5, -6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(1 - 0.5, 3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 

```

``` {r}
#| title: Independence Day

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-07-04")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from July 4", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-4 - 0.5, -4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(0 - 0.5, 1 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(3 - 0.5, 3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

## Row {height="250px"}

``` {r}
#| title: Labor Day

labor <- base %>%
  distinct(year) %>%
  mutate(
    sept1 = as.Date(paste0(year, "-09-01")),
    holiday = sept1 + (8 - wday(sept1, week_start=1)) %% 7
  )

data <- base %>%
  left_join(labor, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Labor Day", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-3 - 0.5, 0 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(5 - 0.5, 6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

``` {r}
#| title: Halloween

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-10-31")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from October 31", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-4 - 0.5, -4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(-2 - 0.5, -2 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(4 - 0.5, 4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(6 - 0.5, 6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

## Row {height="250px"}

``` {r}
#| title: Thanksgiving

thanksgiving <- base %>%
  distinct(year) %>%
  mutate(
    nov1 = as.Date(paste0(year, "-11-01")),
    holiday = nov1 + (4 - wday(nov1) + 7) %% 7 + 21
  )

data <- base %>%
  left_join(thanksgiving, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Thanksgiving", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-3 - 0.5, -3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(2 - 0.5, 3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

``` {r}
#| title: Christmas

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-12-25")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from December 25", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(4 - 0.5, 4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

# Drivers

``` {r}
# drivers object with BACs only
drivers <- person_level %>%
  mutate(
    driver_flag = case_when(
      Person.Type %in% c(
        "1 - DRIVER",
        "5 - DRIVER OF MOTORCYCLE TYPE VEHICLE"
      ) ~ 1,
      TRUE ~ 0
    ),

    bac_value = as.numeric(Person.Blood.Alcohol.Content.Test.Result),

    bac_result = case_when(
      !is.na(bac_value) & bac_value > 0 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    driver_flag == 1,
  )
```

## Row {height="200px"}

``` {r}
#| content: valuebox
#| title: "Total Intoxicated Drivers (YTD)"

total_intox_drivers <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  summarise(n = n()) %>%
  pull(n)

list(
  icon = "emoji-dizzy",
  color = "primary",
  value = format(total_intox_drivers, big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Average BAC Result (YTD)"
bac_stats <- drivers %>%
  filter(bac_result == 1) %>%
  summarise(
    mean_bac = mean(bac_value, na.rm = TRUE)
  )
list(
  icon = "moisture",
  color = "danger",
  value = paste0(
    round(bac_stats$mean_bac, 2)
  )
)
```

```{r}
#| content: valuebox
#| title: "Average Age (YTD)"

age_stats <- drivers %>%
  filter(bac_value>=0.08 & bac_result==1) %>%
  mutate(
    age_num = as.numeric(Person.Age)
  ) %>%
  summarise(
    mean_age = mean(age_num, na.rm = TRUE),
    sd_age = sd(age_num, na.rm = TRUE)
  )

list(
  icon = "person-fill",
  color = "primary",
  value = paste0(
    round(age_stats$mean_age, 0)
  )
)
```

## Row { height=70%}

### Column {.tabset}

``` {r}
#| title: Driver's Zipcode (YTD)
zip_top10 <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1,
         !is.na(Driver.Zip.Code)) %>%
  group_by(Driver.Zip.Code) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  slice_head(n = 10)

zip_top10 <- zip_top10 %>%
  mutate(
    pct = n / sum(n),
    label_full = paste0(Driver.Zip.Code, "\n", n, " (", scales::percent(pct), ")")
  )


plot_ly(
  data = zip_top10,
  type = "treemap",
  labels = ~label_full,
  parents = rep("", nrow(zip_top10)),
  values = ~n,
  textinfo = "label",
  marker = list(
    colors = rep("#f4b942", nrow(zip_top10))
  ),
  hovertemplate = paste(
    "ZIP: %{label}<br>",
    "Drivers: %{value}<br>",
    "<extra></extra>"
  )
) 

```

```{r}
#| title: Driver's Demographics (YTD)

drivers_clean <- drivers %>%
  mutate(
    Person.Ethnicity = case_when(
      
      # Missing / unknown categories
      Person.Ethnicity %in% c("No Data", "99 - UNKNOWN", "UNKNOWN", "", NA) ~ NA_character_,
      
      # Special cases
      Person.Ethnicity == "95 - AUTONOMOUS" ~ "Autonomous",
      Person.Ethnicity == "98 - OTHER" ~ "Other",
      
      # Race/Ethnicity categories
      grepl("A - ASIAN", Person.Ethnicity) ~ "Asian",
      grepl("B - BLACK", Person.Ethnicity) ~ "Black",
      grepl("H - HISPANIC", Person.Ethnicity) ~ "Hispanic",
      grepl("I - AMER", Person.Ethnicity) ~ "American Indian/Alaska Native",
      grepl("W - WHITE", Person.Ethnicity) ~ "White",
      
      # Fallback
      TRUE ~ NA_character_
    ),
    
    Person.Gender = case_when(
      Person.Gender %in% c("99 - UNKNOWN", "UNKNOWN", "No Data", "") ~ NA_character_,
      Person.Gender == "1 - MALE" ~ "Male",
      Person.Gender == "2 - FEMALE" ~ "Female",
      TRUE ~ NA_character_
    )
  )

demo_table <- drivers_clean %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  filter(!is.na(Person.Ethnicity), !is.na(Person.Gender)) %>%
  
  mutate(
    Person.Gender = factor(Person.Gender, levels = c("Female", "Male")),
    Person.Age = as.numeric(Person.Age)
  ) %>%
  
  group_by(Person.Ethnicity, Person.Gender) %>%
  summarise(
    avg_bac = mean(bac_value, na.rm = TRUE),
    avg_age = mean(Person.Age, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    pct = scales::percent(n / sum(n), accuracy = 0.1),
    
    avg_bac = round(avg_bac, 2),
    avg_age = round(avg_age, 0)
  ) %>%
  
  arrange(desc(n)) %>%
  
  rename(
    "Race/Ethnicity" = Person.Ethnicity,
    "Sex" = Person.Gender,
    "Average Age" = avg_age,
    "Average BAC" = avg_bac,
    "Count" = n,
    "Percent" = pct
  )


reactable(
  demo_table,

  defaultPageSize = 10,
  searchable = TRUE,

  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,

  columns = list(
    `Average BAC` = colDef(format = colFormat(digits = 2)),
    `Average Age` = colDef(format = colFormat(digits = 0)),

    Count = colDef(
      align = "center"
    ),

    Percent = colDef(
      align = "center"
    )
  ),

  defaultSorted = list(Count = "desc")
)

```


### Column {.tabset}

``` {r}
#| title: Driver's Vehicle Information (YTD)

vehicle_table <- drivers %>%
  
  # keep BAC filter
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  
  # only remove No Data for Make + Model
  filter(
    !Vehicle.Make %in% c("No Data", "No data", "", NA),
    !Vehicle.Model.Name %in% c("No Data", "No data", "", NA),
    !is.na(Vehicle.Make),
    !is.na(Vehicle.Model.Name)
  ) %>%
  
  mutate(

    # Make cleanup
    make = case_when(
      toupper(Vehicle.Make) %in% c("GMC", "BMW") ~ toupper(Vehicle.Make),
      TRUE ~ str_to_title(tolower(Vehicle.Make))
    ),

    # Model cleanup
    model = Vehicle.Model.Name %>%
      str_remove("\\s*\\(.*\\)") %>%
      str_to_sentence()
  ) %>%
  
  # ----------------------------
  # Group only Make + Model
  # ----------------------------
  group_by(make, model) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  
  # Percent of total
  mutate(
    pct = scales::percent(n / sum(n), accuracy = 0.1)
  ) %>%
  
  arrange(desc(n))

reactable(
  vehicle_table,

  searchable = TRUE,
  sortable = TRUE,
  pagination = TRUE,
  defaultPageSize = 10,

  columns = list(

    make = colDef(name = "Make"),
    model = colDef(name = "Model"),

    n = colDef(name = "Count"),

    pct = colDef(name = "Percent")
  ),

  defaultSorted = list(n = "desc")
)

```

```{r}
#| title: Driver's Vehicle Color (YTD)
plot_data <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  mutate(
    color = case_when(
      grepl("BLK", Vehicle.Color) ~ "Black",
      grepl("WHI", Vehicle.Color) ~ "White",
      grepl("GRY", Vehicle.Color) ~ "Gray",
      grepl("RED", Vehicle.Color) ~ "Red",
      grepl("BLU", Vehicle.Color) ~ "Blue",
      grepl("GRN", Vehicle.Color) ~ "Green",
      grepl("SIL", Vehicle.Color) ~ "Silver",
      grepl("TAN", Vehicle.Color) ~ "Tan",
      grepl("MAR", Vehicle.Color) ~ "Maroon",
      grepl("YEL", Vehicle.Color) ~ "Yellow",
      grepl("GLD", Vehicle.Color) ~ "Gold",
      grepl("PLE", Vehicle.Color) ~ "Purple",
      grepl("BGE", Vehicle.Color) ~ "Beige",
      TRUE ~ "Other"
    )
  ) %>%
  count(color) %>%
  mutate(
    pct = n / sum(n),
    label = paste0(n, " (", scales::percent(pct), ")")
  )
plot_ly(
  data = plot_data,
  x = ~n,
  y = ~fct_reorder(color, n),
  type = "bar",
  orientation = "h",
  marker = list(color = "steelblue"),
  text = ~label,
  textposition = "outside"
) %>%
layout(
  xaxis = list(
    title = "Total intoxicated drivers",
    range = c(0, max(plot_data$n) * 1.15)  # 👈 key fix
  ),
  yaxis = list(title = ""),
  margin = list(r = 120)  # optional extra breathing room
) 

```




# Map

## Row {height=100%}

### Column {.tabset width=70%}

```{r}
#| title: Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "black"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

```{r}
#| title: Fatal Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026,
         Fatal.Crash.Flag == 1) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "maroon"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

### Column {.tabset width=30%}

``` {r}
#| title: Crash Locations (YTD)
top_streets <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    Street.Name = case_when(
      str_detect(Street.Name, " ") ~ str_to_title(str_to_lower(Street.Name)),
      TRUE ~ Street.Name
    ),
    
    # Optional: clean common suffixes to standard casing
    Street.Name = str_replace_all(Street.Name, "\\bDr\\b", "Dr"),
    Street.Name = str_replace_all(Street.Name, "\\bBlvd\\b", "Blvd"),
    Street.Name = str_replace_all(Street.Name, "\\bCt\\b", "Ct"),
    Street.Name = str_replace_all(Street.Name, "\\bRd\\b", "Rd"),
    Street.Name = str_replace_all(Street.Name, "\\bSt\\b", "St"),
    Street.Name = str_replace_all(Street.Name, "\\bAve\\b", "Ave"),
    
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Street Name" = Street.Name,
    "Crash Count" = n,
    "Percent (%)" = percent
  )

reactable(
  top_streets,
  defaultPageSize = 15,
  striped = TRUE,
  highlight = TRUE,
  searchable = TRUE,
  defaultSorted = list(`Crash Count` = "desc"),
  columns = list(
    `Street Name` = colDef(minWidth = 200),
    `Crash Count` = colDef(align = "center"),
    `Percent (%)` = colDef(
      format = colFormat(digits = 2),
      align = "center"
    )
  ),
  compact = TRUE,
  bordered = FALSE
)
```

``` {r}
#| title: Fatal Crash Locations (YTD)
top_fatal_streets <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    Fatal.Crash.Flag == 1
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    Street.Name = case_when(
      str_detect(Street.Name, " ") ~ str_to_title(str_to_lower(Street.Name)),
      TRUE ~ Street.Name
    ),
    
    # Preserve directional abbreviations
    Street.Name = str_replace_all(Street.Name, "\\b(N|S|E|W|NE|NW|SE|SW)\\b", toupper),
    
    # Preserve street suffix casing
    Street.Name = str_replace_all(Street.Name, "\\bDr\\b", "Dr"),
    Street.Name = str_replace_all(Street.Name, "\\bBlvd\\b", "Blvd"),
    Street.Name = str_replace_all(Street.Name, "\\bCt\\b", "Ct"),
    Street.Name = str_replace_all(Street.Name, "\\bRd\\b", "Rd"),
    Street.Name = str_replace_all(Street.Name, "\\bSt\\b", "St"),
    Street.Name = str_replace_all(Street.Name, "\\bAve\\b", "Ave"),
    
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Street Name" = Street.Name,
    "Fatal Crashes" = n,
    "Percent (%)" = percent
  )

reactable(
  top_fatal_streets,
  
  defaultPageSize = 10,
  striped = TRUE,
  highlight = TRUE,
  searchable = TRUE,
  
  defaultSorted = list(`Fatal Crashes` = "desc"),
  
  columns = list(
    `Street Name` = colDef(minWidth = 220),
    `Fatal Crashes` = colDef(align = "center"),
    `Percent (%)` = colDef(
      format = colFormat(digits = 2),
      align = "center"
    )
  ),
  
  compact = TRUE,
  bordered = FALSE
)
```

# Quiz

::: {.column width="100%"}

::: {.card title="Drunk Driving Quiz" style="min-height:780px; margin:0 auto;"}

<div id="quizApp">

  <h4 id="introStats">Loading stats...</h4>
  <h4 id="avgStats"></h4>

  <button onclick="startQuiz()" id="startBtn">Start Quiz</button>

  <p id="tagline" style="font-style: italic;"></p>

  <hr>

  <div id="quizBox" style="display:none"></div>

</div>

<script>
const SCRIPT_URL = "https://script.google.com/macros/s/AKfycbw4TekPtUfV08eUUU9TbRu0QdeUl9_mondBZS0c179w-eQQb4YliAQwwUHyODuErsYA/exec";

let qIndex = 0;
let score = 0;

/* ---------------- QUIZ ---------------- */
const quiz = [

  {
    q: "What day(s) of the week do alcohol-related crashes most often occur?",
    a: ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Weekends (Saturday and Sunday)"],
    correct: "Weekends (Saturday and Sunday)",
    explanation: "About 50% of all alcohol-related crashes in El Paso occur on the weekend."
  },

  {
    q: "What time of day do alcohol-related crashes most often occur?",
    a: ["10:00 AM to 5:00 PM", "12:00 PM to 6:00 PM", "6:00 PM to 11:00 PM", "9:00 PM to 3:00 AM"],
    correct: "9:00 PM to 3:00 AM",
    explanation: "More than half of all alcohol-related crashes in El Paso occur during late-night and early morning hours."
  },

  {
    q: "Since 2016, the total number of alcohol-related crashes in El Paso has decreased.",
    a: ["True", "False"],
    correct: "True",
    explanation: "Overall trends show a decline in alcohol-related crashes since 2016. The highest number of alcohol-related crashes recorded was in September 2017."
  },

  {
    q: "What is the legal blood alcohol concentration (BAC) limit for driving in Texas?",
    a: ["0.05%", "0.08%", "0.10%"],
    correct: "0.08%",
    explanation: "In Texas, it is illegal to drive with a BAC of 0.08% or higher. This level indicates significant impairment in reaction time and judgment."
  },

  {
    q: "On average, what blood alcohol concentration (BAC) level do drivers involved in alcohol-related crashes in El Paso have?",
    a: ["Less than the legal limit (<0.08%)", "Around the legal limit (~0.08%)", "More than twice the legal limit (>0.16%)"],
    correct: "More than twice the legal limit (>0.16%)",
    explanation: "The data shows that the average BAC among drivers who crashed and were tested for alcohol is greater than 0.16, which is more than twice the legal limit and associated with severe impairment."
  },

  {
    q: "On which holiday do crashes sharply increase in El Paso?",
    a: ["Memorial Day", "Christmas", "New Years Day"],
    correct: "New Years Day",
    explanation: "The data shows the average number of alcohol-involved crashes on New Years Day doubles compared to the previous day."
  },

  {
    q: "Which roadway has the most alcohol-related traffic fatalities in El Paso?",
    a: ["Mesa Street", "Zaragoza Road", "I-10"],
    correct: "I-10",
    explanation: "The majority of crashes and fatalities occur along I-10 in El Paso, likely due to higher speeds and increased crash severity when alcohol is involved."
  }

];

/* ---------------- STATS ---------------- */
function loadStats() {

  fetch(SCRIPT_URL)
    .then(r => r.text())
    .then(t => {

      const data = JSON.parse(t);
      data.shift();

      const scores = data
        .map(r => Number(r[3]))
        .filter(v => !isNaN(v));

      const count = scores.length;

      const avg = count
        ? (scores.reduce((a,b)=>a+b,0)/count).toFixed(1)
        : 0;

      document.getElementById("introStats").innerHTML =
        `${count} people have taken this quiz.`;

      document.getElementById("avgStats").innerHTML =
        `Average score: ${avg}%`;
    });
}

/* ---------------- START ---------------- */
function startQuiz() {
  document.getElementById("startBtn").style.display = "none";
  document.getElementById("quizBox").style.display = "block";
  loadQuestion();
}

/* ---------------- LOAD QUESTION ---------------- */
function loadQuestion() {

  const q = quiz[qIndex];
  const box = document.getElementById("quizBox");

  box.innerHTML = `
    <h4>Question ${qIndex + 1} of ${quiz.length}</h4>
    <p>${q.q}</p>
    <div id="answers"></div>
    <p id="feedback"></p>
    <button id="nextBtn" style="display:none" onclick="nextQuestion()">Next</button>
  `;

  const answersDiv = document.getElementById("answers");

  q.a.forEach(choice => {

    const btn = document.createElement("button");
    btn.innerText = choice;

    btn.onclick = () => {

      const buttons = answersDiv.querySelectorAll("button");
      buttons.forEach(b => b.disabled = true);

      if (choice === q.correct) {
        score++;
        btn.style.backgroundColor = "#c8f7c5";

        document.getElementById("feedback").innerHTML =
          `<span style="color:green;"><b>Correct</b></span><br><br>${q.explanation}`;
      } else {
        btn.style.backgroundColor = "#f7c5c5";

        buttons.forEach(b => {
          if (b.innerText === q.correct) {
            b.style.backgroundColor = "#c8f7c5";
          }
        });

        document.getElementById("feedback").innerHTML =
          `<b>Correct answer:</b> ${q.correct}<br><br>${q.explanation}`;
      }

      document.getElementById("nextBtn").style.display = "inline-block";
    };

    answersDiv.appendChild(btn);
  });
}

/* ---------------- NEXT ---------------- */
function nextQuestion() {
  qIndex++;

  if (qIndex < quiz.length) {
    loadQuestion();
  } else {
    finishQuiz();
  }
}

/* ---------------- FINISH ---------------- */
function finishQuiz() {

  const percent = Math.round((score / quiz.length) * 100);

  fetch(SCRIPT_URL, {
    method: "POST",
    body: JSON.stringify({
      timestamp: new Date().toISOString(),
      score: percent
    })
  });

  document.getElementById("quizBox").innerHTML =
    `<h3>Finished</h3>
     <p>Your score: <b>${percent}%</b></p>`;
}

/* ---------------- INIT ---------------- */
document.addEventListener("DOMContentLoaded", loadStats);
</script>
