---
title: "El Paso Alcohol-Related Crash Dashboard"
format:
  dashboard:
    theme: default
    css: styles2.css
    orientation: rows
execute:
  echo: false
  warning: false
  message: false
---

Welcome to the El Paso Alcohol-Related Crash Dashboard!

This dashboard displays data on alcohol-related vehicle crashes in El Paso, Texas and is updated on a monthly basis.

```{r}

rm(list = ls())

suppressMessages(library(gt))
suppressMessages(library(quarto))
suppressMessages(library(rsconnect))
suppressMessages(library(shiny))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
suppressMessages(library(stringr))
suppressMessages(library(readr))
suppressMessages(library(tidyverse))
suppressMessages(library(leaflet))
suppressMessages(library(leaflet.extras))


files <- c(
  austin = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/austin_workingdata.csv",
  brownsville = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/brownsville_workingdata.csv",
  dallas = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/dallas_workingdata.csv",
  elpaso = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/elpaso_workingdata.csv",
  fortworth = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/fortworth_workingdata.csv",
  houston = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/houston_workingdata.csv",
  laredo = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/laredo_workingdata.csv",
  mcallen = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/mcallen_workingdata.csv",
  sanantonio = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/sanantonio_workingdata.csv"
)

all_cities <- bind_rows(lapply(names(files), function(city) {
  df <- read.csv(files[[city]], header = TRUE)
  df$city <- city
  df
}))

all_cities$City <- recode(all_cities$City,
  "AUSTIN" = "Austin",
  "BROWNSVILLE" = "Brownsville",
  "DALLAS" = "Dallas",
  "EL PASO" = "El Paso",
  "FORT WORTH" = "Fort Worth",
  "HOUSTON" = "Houston",
  "MCALLEN" = "McAllen",
  "SAN ANTONIO" = "San Antonio",
  "LAREDO" = "Laredo"
)

# drop current month, since its incomplete
all_cities <- all_cities %>%
  mutate(
    Crash.Date = mdy(Crash.Date)
  ) %>%
  filter(
    !(year(Crash.Date) == year(Sys.Date()) &
      month(Crash.Date) == month(Sys.Date()))
  )

person_level <- all_cities

vehicle_level <- all_cities %>%
  distinct(Crash.ID, Unit.Number, .keep_all = TRUE)

crash_level <- all_cities %>%
  distinct(Crash.ID, .keep_all = TRUE) 
```

# Trends


## Row {height=30%}

```{r}
#| content: valuebox
#| title: "Total Crashes (YTD)"
ytd_alcohol_crashes <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  summarise(
    ytd_alcohol_crashes = n_distinct(Crash.ID)
  ) %>%
  pull(ytd_alcohol_crashes)

list(
  icon = "car-front-fill",
  color = "primary",
  value = ytd_alcohol_crashes )
```

```{r}
#| content: valuebox
#| title: "Total People Injured (YTD)"

person_level <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0
    )
  )

ytd_people_alcohol <- person_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    injured == 1) %>%
    nrow()


list(
  icon = "bandaid",
  color = "primary",
  value = ytd_people_alcohol)

```

```{r}
#| content: valuebox
#| title: "Total Fatalities (YTD)"
person_level <- person_level %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    )
  )
ytd_alcohol_fatalities <- person_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    fatal == 1) %>%
    nrow()

list(
  icon = "hospital",
  color = "danger",
  value = ytd_alcohol_fatalities
)
```


## Row {height=70%}

### Column {.tabset}

``` {r}
#| title: Crashes
monthly_counts <- crash_level %>%
  mutate(
    month = floor_date(Crash.Date, "month"),
  ) %>%
  group_by(month) %>%
  summarise(
    alcohol_crashes = n_distinct(Crash.ID),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    alcohol_ma = zoo::rollmean(alcohol_crashes, k = 12, fill = NA, align = "center")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed
  geom_line(
    aes(y = alcohol_crashes, color = "Total crashes"),
    linewidth = 0.8
  ) +

  # moving average
  geom_line(
    aes(y = alcohol_ma, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total crashes" = "grey70",
      "Average" = "black"
    )
  ) +

  labs(
    x = "",
    y = "Total crashes",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)
```

``` {r}
#| title: Fatalities
monthly_counts <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    alcohol_fatal_crashes = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    alcohol_fatal_ma = zoo::rollmean(alcohol_fatal_crashes,
                                     k = 12,
                                     fill = NA,
                                     align = "right")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed total fatalities (grey)
  geom_line(
    aes(y = alcohol_fatal_crashes, color = "Total fatalities"),
    linewidth = 1
  ) +

  # alcohol fatalities 12-mo moving average (maroon)
  geom_line(
    aes(y = alcohol_fatal_ma, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total fatalities" = "grey70",
      "Average" = "maroon"
    )
  ) +

  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, 2)
  ) +

  labs(
    x = "",
    y = "Total fatalities",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)
```

### Column {.tabset}

``` {r}
#| title: Day of Crash (YTD)
dow_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  count(Day.of.Week) %>%
  mutate(
    Day.of.Week = factor(
      tools::toTitleCase(tolower(Day.of.Week)),
      levels = c("Sunday", "Monday", "Tuesday", "Wednesday",
                 "Thursday", "Friday", "Saturday")
    )
  )

avg_val <- mean(dow_data$n, na.rm = TRUE)

p3 <- ggplot(dow_data, aes(x = Day.of.Week, y = n)) +

  # bars
  geom_col(fill = "steelblue") +

  # average line
  geom_hline(
    aes(yintercept = avg_val, color = "Average"),
    linewidth = 1.2,
    linetype = "dashed"
  ) +

  # fake mapping for legend consistency
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "",
    y = "Total crashes"
  )

ggplotly(p3)

```

``` {r}
#| title: Hour of Crash (YTD)
hour_data <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  mutate(
    hour = as.numeric(substr(Hour.of.Day, 1, 2)),
    time_block = case_when(
      hour >= 0 & hour <= 5 ~ "00:00–05:59",
      hour >= 6 & hour <= 11 ~ "06:00–11:59",
      hour >= 12 & hour <= 17 ~ "12:00–17:59",
      hour >= 18 & hour <= 23 ~ "18:00–23:59",
      TRUE ~ "Unknown"
    )
  ) %>%
  count(time_block)

avg_block <- mean(hour_data$n, na.rm = TRUE)

p4 <- ggplot(hour_data, aes(x = time_block, y = n)) +

  geom_col(fill = "steelblue") +

  geom_hline(
    yintercept = avg_block,
    color = "orange",
    linetype = "dashed",
    linewidth = 1.2
  ) +

  theme_minimal() +
  theme(
    legend.position = "none"
  ) +
  labs(
    x = "",
    y = "Total crashes"
  )

ggplotly(p4)
```


# Drivers

``` {r}
# drivers object with BACs
drivers <- person_level %>%
  mutate(
    driver_flag = case_when(
      Person.Type %in% c(
        "1 - DRIVER",
        "5 - DRIVER OF MOTORCYCLE TYPE VEHICLE"
      ) ~ 1,
      TRUE ~ 0
    ),

    bac_value = as.numeric(Person.Blood.Alcohol.Content.Test.Result),

    bac_result = case_when(
      !is.na(bac_value) & bac_value > 0 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    driver_flag == 1,
    bac_result == 1
  )
```

## Row {height=30%}

```{r}
#| content: valuebox
#| title: "Average BAC (YTD)"
bac_stats <- drivers %>%
  summarise(
    mean_bac = mean(bac_value, na.rm = TRUE),
    sd_bac = sd(bac_value, na.rm = TRUE)
  )
list(
  icon = "moisture",
  color = "danger",
  value = paste0(
    round(bac_stats$mean_bac, 2)
  )
)
```

```{r}
#| content: valuebox
#| title: "Male Drivers (YTD)"
male_pct <- drivers %>%
  mutate(
    male = if_else(Person.Gender == "1 - MALE", 1, 0)
  ) %>%
  summarise(
    pct_male = mean(male, na.rm = TRUE) * 100
  ) %>%
  pull(pct_male)

list(
  icon = "gender-male",
  color = "primary",
  value = paste0(round(male_pct, 1), "%")
)
```

```{r}
#| content: valuebox
#| title: "Average Age (YTD)"

age_stats <- drivers %>%
  mutate(
    age_num = as.numeric(Person.Age)
  ) %>%
  summarise(
    mean_age = mean(age_num, na.rm = TRUE),
    sd_age = sd(age_num, na.rm = TRUE)
  )

list(
  icon = "person-fill",
  color = "primary",
  value = paste0(
    round(age_stats$mean_age, 1)
  )
)
```

## Row { height=70%}

### Column

```{r}
#| title: Drivers Over the Legal Limit (YTD)


drivers %>%
  mutate(
    bac_group = if_else(
      bac_value >= 0.08,
      "0.08+ (Over legal limit)",
      "<0.08"
    )
  ) %>%
  count(bac_group) %>%
  plot_ly(
    labels = ~bac_group,
    values = ~n,
    type = "pie",
    textinfo = "label+percent",
    marker = list(
      colors = c("maroon", "lightgray")
    )
  )

```


### Column {.tabset}

```{r}
#| title: Vehicle Color
drivers %>%
  mutate(
    color = case_when(
      grepl("BLK", Vehicle.Color) ~ "Black",
      grepl("WHI", Vehicle.Color) ~ "White",
      grepl("GRY", Vehicle.Color) ~ "Gray",
      grepl("RED", Vehicle.Color) ~ "Red",
      grepl("BLU", Vehicle.Color) ~ "Blue",
      grepl("GRN", Vehicle.Color) ~ "Green",
      grepl("SIL", Vehicle.Color) ~ "Silver",
      grepl("TAN", Vehicle.Color) ~ "Tan",
      grepl("MAR", Vehicle.Color) ~ "Maroon",
      grepl("YEL", Vehicle.Color) ~ "Yellow",
      grepl("GLD", Vehicle.Color) ~ "Gold",
      grepl("PLE", Vehicle.Color) ~ "Purple",
      grepl("BGE", Vehicle.Color) ~ "Beige",
      TRUE ~ "Other"
    )
  ) %>%
  count(color) %>%
  plot_ly(
    x = ~n,
    y = ~fct_reorder(color, n),
    type = "bar",
    orientation = "h",
    marker = list(color = "steelblue")
  ) %>%
  layout(xaxis = list(title = "Total Drivers"), yaxis = list(title = ""))

```

``` {r}
#| title: Vehicle Body Style
drivers %>%
  mutate(
    body = case_when(
      grepl("MC", Vehicle.Body.Style) ~ "Motorcycle",
      grepl("P2", Vehicle.Body.Style) ~ "Passenger Car (2-door)",
      grepl("P4", Vehicle.Body.Style) ~ "Passenger Car (4-door)",
      grepl("PK", Vehicle.Body.Style) ~ "Pickup",
      grepl("SV", Vehicle.Body.Style) ~ "SUV",
      grepl("TR", Vehicle.Body.Style) ~ "Truck",
      TRUE ~ "Other"
    )
  ) %>%
  count(body) %>%
  plot_ly(
    x = ~n,
    y = ~fct_reorder(body, n),
    type = "bar",
    orientation = "h",
    marker = list(color = "steelblue")
  ) %>%
  layout(xaxis = list(title = "Total Drivers"), yaxis = list(title = ""))
```

``` {r}
#| title: Vehicle Make
drivers %>%
  mutate(
    make = case_when(
      Vehicle.Make %in% c("FORD","CHEVROLET","DODGE","GMC") ~ Vehicle.Make,
      Vehicle.Make %in% c("TOYOTA","HONDA","NISSAN","HYUNDAI","KIA","MAZDA","SUBARU","VOLKSWAGEN","LEXUS") ~ Vehicle.Make,
      TRUE ~ NA_character_
    ),
    make = str_to_title(make)
  ) %>%
  filter(!is.na(make)) %>%
  count(make) %>%
  plot_ly(
    x = ~n,
    y = ~fct_reorder(make, n),
    type = "bar",
    orientation = "h",
    marker = list(color = "steelblue")
  ) %>%
  layout(
    xaxis = list(title = "Total drivers"),
    yaxis = list(title = "")
  )
```

# Map

## Row {height=100%}

### Column {.tabset width=75%}

```{r}
#| title: Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "black"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

```{r}
#| title: Fatal Crashes (Since 2016)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Fatal.Crash.Flag == 1) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "maroon"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

### Column {width=25%}

``` {r}
#| title: Crash Locations (YTD)
top_10_streets <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  slice_head(n = 15) %>%
  rename(
    "Street Name" = Street.Name,
    "Crash Count" = n,
    "Percent (%)" = percent
  )

top_10_streets %>%
  gt()
```


# About

## Row {height=50%}

::: {.card title="About This Dashboard"}

Last updated: March 2026

All data can be downloaded at the Texas Department of Transportation's (TXDOT) website, using the TXDOT-authored filter in the
[Crash Query Tool](https://cris.dot.state.tx.us/public/Query/app/home). This dashboard specifically shows data on crashes classified as alcohol-related and that occurred in the city of El Paso.

:::

::: {.card title="Technical Details"}

#### Definition of Alcohol-Related Crash

A crash is defined as alcohol-related crashes if involved drivers or individuals were reported as:
- "Under the Influence of Alcohol"
- "Had Been Drinking"
- had a BAC greater than 0

#### Trends Page

- Averages for crashes and fatal crashes were calculated using a 12-month moving average. This technique results in a smooth line, allowing the viewer to see if a trend is increasing or decreasing over a period of time.
- Injuries include people who were classified as having a suspected serious injury, suspected minor injury, or possible injury.

:::

# Quiz

::: {.column width="100%"}

::: {.card title="Drunk Driving Quiz" style="min-height:780px; margin:0 auto;"}

<div id="quizApp">

  <h4 id="introStats">Loading stats...</h4>
  <h4 id="avgStats"></h4>

  <button onclick="startQuiz()" id="startBtn">Start Quiz</button>

  <p id="tagline" style="font-style: italic;"></p>

  <hr>

  <div id="quizBox" style="display:none"></div>

</div>


<script>
const SCRIPT_URL = "https://script.google.com/macros/s/AKfycbw4TekPtUfV08eUUU9TbRu0QdeUl9_mondBZS0c179w-eQQb4YliAQwwUHyODuErsYA/exec";

let qIndex = 0;
let score = 0;

/* ---------------- QUIZ ---------------- */
const quiz = [

  {
    q: "When are alcohol-related crashes most likely to happen in El Paso?",
    a: ["Weekday mornings", "Weekends (Saturday and Sunday)", "Midday during the week"],
    correct: "Weekends (Saturday and Sunday)",
    explanation: "Crash data shows that alcohol-related crashes in El Paso are most common on weekends."
  },

  {
    q: "What time of day do alcohol-related crashes most often occur?",
    a: ["6:00 AM to 12:00 PM", "12:00 PM to 6:00 PM", "6:00 PM to 6:00 AM"],
    correct: "6:00 PM to 6:00 AM",
    explanation: "Most alcohol-related crashes happen at night and early morning hours, when visibility is lower and alcohol consumption is more common."
  },

  {
    q: "Since 2016, what has happened to alcohol-related crashes in El Paso?",
    a: ["They have increased steadily", "They have decreased overall", "They have stayed exactly the same"],
    correct: "They have decreased overall",
    explanation: "Overall trends show a decline in alcohol-related crashes since 2016. There was a sharp drop during COVID-19, followed by a return toward normal levels, and then a continued downward trend."
  },

  {
    q: "What is the legal blood alcohol concentration (BAC) limit for driving in Texas?",
    a: ["0.05%", "0.08%", "0.10%"],
    correct: "0.08%",
    explanation: "In Texas, it is illegal to drive with a BAC of 0.08% or higher. This level indicates significant impairment in reaction time and judgment."
  },

  {
    q: "On average, what BAC level do drivers involved in alcohol-related crashes in El Paso have?",
    a: ["About 0.08% (legal limit)", "About 0.10%", "About 0.16% (more than twice the legal limit)"],
    correct: "About 0.16% (more than twice the legal limit)",
    explanation: "The data shows that many drivers involved in alcohol-related crashes have very high BAC levels—around 0.16%—which is more than twice the legal limit and associated with severe impairment."
  }

];

/* ---------------- STATS ---------------- */
function loadStats() {

  fetch(SCRIPT_URL)
    .then(r => r.text())
    .then(t => {

      const data = JSON.parse(t);
      data.shift();

      const scores = data
        .map(r => Number(r[3]))   // COLUMN D
        .filter(v => !isNaN(v));

      const count = scores.length;

      const avg = count
        ? (scores.reduce((a,b)=>a+b,0)/count).toFixed(1)
        : 0;

      document.getElementById("introStats").innerHTML =
        `${count} people have taken this quiz.`;

      document.getElementById("avgStats").innerHTML =
        `Average score: ${avg}%`;

      document.getElementById("tagline").innerHTML =
        "";
    });
}

/* ---------------- START ---------------- */
function startQuiz() {
  document.getElementById("startBtn").style.display = "none";
  document.getElementById("quizBox").style.display = "block";
  loadQuestion();
}

/* ---------------- LOAD QUESTION ---------------- */
function loadQuestion() {

  const q = quiz[qIndex];
  const box = document.getElementById("quizBox");

  box.innerHTML = `
    <h4>Question ${qIndex + 1} of ${quiz.length}</h4>
    <p>${q.q}</p>
    <div id="answers"></div>
    <p id="feedback"></p>
    <button id="nextBtn" style="display:none" onclick="nextQuestion()">Next</button>
  `;

  const answersDiv = document.getElementById("answers");

  q.a.forEach(choice => {

    const btn = document.createElement("button");
    btn.innerText = choice;

    btn.onclick = () => {

      const buttons = answersDiv.querySelectorAll("button");
      buttons.forEach(b => b.disabled = true);

      if (choice === q.correct) {
        score++;
        btn.style.backgroundColor = "#c8f7c5";

        document.getElementById("feedback").innerHTML =
          `
          <span style="color:green;"><b>Correct</b></span><br><br>
          ${q.explanation}
          `;
      } else {
        btn.style.backgroundColor = "#f7c5c5";

        buttons.forEach(b => {
          if (b.innerText === q.correct) {
            b.style.backgroundColor = "#c8f7c5";
          }
        });

          document.getElementById("feedback").innerHTML =
          `
          <b>Correct answer:</b> ${q.correct}<br><br>
          ${q.explanation}
          `;
      }

      document.getElementById("nextBtn").style.display = "inline-block";
    };

    answersDiv.appendChild(btn);
  });
}

/* ---------------- NEXT ---------------- */
function nextQuestion() {

  qIndex++;

  if (qIndex < quiz.length) {
    loadQuestion();
  } else {
    finishQuiz();
  }
}

/* ---------------- FINISH ---------------- */
function finishQuiz() {

  const percent = Math.round((score / quiz.length) * 100);

  fetch(SCRIPT_URL, {
    method: "POST",
    body: JSON.stringify({
      timestamp: new Date().toISOString(),
      score: percent
    })
  });

  document.getElementById("quizBox").innerHTML =
    `<h3>Finished</h3>
     <p>Your score: <b>${percent}%</b></p>`;
}

/* ---------------- INIT ---------------- */
document.addEventListener("DOMContentLoaded", loadStats);
</script>