---
title: "Alcohol-Related Crash Dashboard"
format:
  dashboard:
    theme: default
    css: styles2.css
    orientation: rows
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
editor: source

---

::: {style="padding:12px 0 18px 0; line-height:1.4;"}
<p style="margin:0; font-size:14px; color:#444;">

This dashboard tracks all reported vehicle crashes in El Paso, Texas and
is updated monthly with the most recent available data.

</p>

<p style="margin:8px 0 0 0; font-size:13px; color:#666;">

<b>How to use:</b><br> • Click the <b>bottom-right expand icon</b> to
enlarge any chart<br> • Hover over plots for detailed values<br> • Click
the <b>camera icon</b> (if available) to download charts as images<br>

</p>

<p style="margin:8px 0 0 0; font-size:13px; color:#666;">

<b>Mobile users:</b><br> • Double-tap the chart to reset the scale<br> •
Use the expand button for full-screen view<br> • Rotate your device for
better readability<br>

</p>

<p style="margin:8px 0 0 0; font-size:12px; color:#888;">

Tip: For best experience, view on desktop for full interactivity.

</p>
:::

```{r}

rm(list = ls())

suppressMessages(library(reactable))
suppressMessages(library(DT))
suppressMessages(library(gt))
suppressMessages(library(quarto))
suppressMessages(library(rsconnect))
suppressMessages(library(shiny))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
suppressMessages(library(stringr))
suppressMessages(library(readr))
suppressMessages(library(tidyverse))
suppressMessages(library(leaflet))
suppressMessages(library(leaflet.extras))
suppressMessages(library(sf))


data <- read.csv("C:/Users/Matt/Documents/data/ep_data_(01-01-2016_05-02-2026).csv", header = TRUE)

data$City <- recode(data$City,
  "EL PASO" = "El Paso"
)

# drop current month, since its incomplete
data <- data %>%
  mutate(
    Crash.Date = mdy(Crash.Date)
  ) %>%
  filter(
    !(year(Crash.Date) == year(Sys.Date()) &
      month(Crash.Date) == month(Sys.Date()))
  )

person_level <- data

vehicle_level <- data %>%
  distinct(Crash.ID, Unit.Number, .keep_all = TRUE)

crash_level <- data %>%
  distinct(Crash.ID, .keep_all = TRUE) 
```

# Monthly Brief {#monthly-brief}

```{r}
# ----------------------------
# REPORTING MONTH (DATA LAGS 1 MONTH)
# ----------------------------
report_date <- Sys.Date() %m-% months(1)
report_month <- format(report_date, "%B %Y")
report_year <- year(report_date)
prev_year <- report_year - 1

# ----------------------------
# MONTHLY DATA
# ----------------------------
this_month <- crash_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == report_year,
    month(Crash.Date) == month(report_date)
  )

last_month <- crash_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == prev_year,
    month(Crash.Date) == month(report_date)
  )

# ----------------------------
# METRICS
# ----------------------------
crashes_now <- n_distinct(this_month$Crash.ID)
crashes_prev <- n_distinct(last_month$Crash.ID)

# ----------------------------
# INJURIES (MATCH YOUR CLEAN LOGIC)
# ----------------------------
injuries_now <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == report_year,
    month(Crash.Date) == month(report_date),
    !Person.Injury.Severity %in% c(
      "99 - UNKNOWN",
      "95 - AUTONOMOUS",
      "N - NOT INJURED",
      "K - FATAL INJURY"
    ),
    !is.na(Person.Injury.Severity)
  ) %>%
  summarise(n = n()) %>%
  pull(n)

injuries_prev <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == prev_year,
    month(Crash.Date) == month(report_date),
    !Person.Injury.Severity %in% c(
      "99 - UNKNOWN",
      "95 - AUTONOMOUS",
      "N - NOT INJURED",
      "K - FATAL INJURY"
    ),
    !is.na(Person.Injury.Severity)
  ) %>%
  summarise(n = n()) %>%
  pull(n)

# ----------------------------
# FATALITIES (KEEP CONSISTENT)
# ----------------------------
fatal_now <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == report_year,
    month(Crash.Date) == month(report_date),
    Person.Injury.Severity == "K - FATAL INJURY"
  ) %>%
  summarise(n = n()) %>%
  pull(n)

fatal_prev <- person_level %>%
  filter(
    City == "El Paso",
    year(Crash.Date) == prev_year,
    month(Crash.Date) == month(report_date),
    Person.Injury.Severity == "K - FATAL INJURY"
  ) %>%
  summarise(n = n()) %>%
  pull(n)



# ----------------------------
# MONTHLY SERIES (CURRENT VS PRIOR YEAR)
# ----------------------------
build_monthly <- function(df, value){
  df %>%
    mutate(
      year = year(Crash.Date),
      month = month(Crash.Date)
    ) %>%
    filter(year %in% c(report_year, prev_year)) %>%
    group_by(year, month) %>%
    summarise(val = {{value}}, .groups="drop") %>%
    complete(year, month = 1:12, fill = list(val = 0)) %>%
    mutate(
      month_lab = factor(month.abb[month], levels = month.abb),
      year = factor(year)
    )
}

# crashes
crash_plot_data <- build_monthly(crash_level %>% filter(City=="El Paso"),
                                n_distinct(Crash.ID))

# fatalities
fatal_plot_data <- build_monthly(
  person_level %>%
    filter(City=="El Paso",
           Person.Injury.Severity == "K - FATAL INJURY"),
  n()
)

# injuries
injury_plot_data <- build_monthly(
  person_level %>%
    filter(
      City=="El Paso",
      !Person.Injury.Severity %in% c(
        "99 - UNKNOWN",
        "95 - AUTONOMOUS",
        "N - NOT INJURED",
        "K - FATAL INJURY"
      ),
      !is.na(Person.Injury.Severity)
    ),
  n()
)


```

## Row {height="250px"}

### Column

::: {.card title="**Monthly Summary**"}
In **`r report_month`**, there were:

- **`r crashes_now`** alcohol-related crashes

- **`r injuries_now`** people injured in an alcohol-related crash

- **`r fatal_now`** fatalities associated with alcohol-related crashes
:::

### Column

::: {.card title="**Navigation Links**"}
- [Monthly Brief](#monthly-brief) — Latest monthly summary of crashes,
  injuries, and fatalities
- [About this Dashboard](#about-this-dashboard) — Data sources,
  definitions, and technical details
- [Trends](#trends) — Yearly trend patterns and year-to-date totals
- [Holidays](#holidays) — Crash patterns around major holidays and event
  periods\
- [Drivers](#drivers) — Driver-level characteristics, risk indicators,
  and year-to-date totals
- [Maps](#maps) — Locations of crashes, peak times, police regions, city districts\
- [Quiz](#quiz) — How well do you know the data in El Paso? Take the
  quiz and find out!
:::


## Row {.tabset height="400px"}

```{r}
# remove future months for current year only
filter_future <- function(df){
  df %>%
    filter(
      !(year == report_year & month > month(report_date))
    )
}

crash_plot_data  <- filter_future(crash_plot_data)
fatal_plot_data  <- filter_future(fatal_plot_data)
injury_plot_data <- filter_future(injury_plot_data)

# comparison months for brief
comparison_month <- format(
  as.Date(paste0(prev_year, "-", month(report_date), "-01")),
  "%B %Y"
)


```

```{r}
#| title: Crashes

## Color pallete
ytd_colors <- c(
  "grey", 
  "steelblue"   
)

p1 <- ggplot(crash_plot_data,
             aes(x = month_lab, y = val, fill = year)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = val, y = val + 1),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  scale_fill_manual(values = ytd_colors) +   # 🔥 FIX
  scale_y_continuous(
    limits = c(0, max(crash_plot_data$val, na.rm = TRUE) * 1.1)
  ) +
  labs(title = "Alcohol-Related Crashes in El Paso",
       x = "", y = "Total crashes", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p1) %>%
  layout(
    yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes in El Paso</b>",
      "<br><sup>2025–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
    xaxis = list(
      showgrid = FALSE,
      zeroline = FALSE
    ),
    annotations = list(
      list(
        x = 0, y = -0.18,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )
```

```{r}
#| title: Injuries

## Color pallete
ytd_colors <- c(
  "grey", 
  "orange"   
)

p1 <- ggplot(injury_plot_data,
             aes(x = month_lab, y = val, fill = year)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = val, y = val + 1),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  scale_fill_manual(values = ytd_colors) +   # 🔥 FIX
  scale_y_continuous(
    limits = c(0, max(injury_plot_data$val, na.rm = TRUE) * 1.1)
  ) +
  labs(title = "Injuries from Alcohol-Related Crashes in El Paso",
       x = "", y = "Total people injured", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p1) %>%
  layout(
    yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE ),
  title = list(
    text = paste0(
      "<b>Injuries from Alcohol-Related Crashes in El Paso</b>",
      "<br><sup>2025–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
    xaxis = list(
      showgrid = FALSE,
      zeroline = FALSE
    ),
    annotations = list(
      list(
        x = 0, y = -0.18,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )

```

```{r}
#| title: Fatalities

## Color pallete
ytd_colors <- c(
  "grey", 
  "maroon"   
)


p1 <- ggplot(fatal_plot_data,
             aes(x = month_lab, y = val, fill = year)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = val, y = val * 1.03),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  scale_fill_manual(values = ytd_colors) +   # 🔥 FIX
  scale_y_continuous(
    limits = c(0, max(fatal_plot_data$val, na.rm = TRUE) * 1.1)
  ) +
  labs(title = "Alcohol-Related Traffic Fatalities in El Paso",
       x = "", y = "Total fatalities", fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p1) %>%
  layout(
    yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE ),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Traffic Fatalities in El Paso</b>",
      "<br><sup>2025–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
    annotations = list(
      list(
        x = 0, y = -0.18,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )


```



# About this Dashboard {#about-this-dashboard}

## Row {height="50%"}

::: {.card title="**About This Dashboard**"}
Last Update: May 2, 2026

The Alcohol-Related Crash Dashboard provides monthly, city-level data on
alcohol-related motor vehicle crashes in El Paso, Texas. It is designed
as a public-facing resource for journalists, policymakers, and
researchers seeking timely and transparent information on impaired
driving trends.

Crash data is downloaded from the Texas Department of Transportation
(TXDOT) website using the [Crash Records Information System (CRIS) Query
Tool](https://cris.dot.state.tx.us/public/Query/app/home). All crash
data in this dashboard represent those reported by law enforcement, were
classified as alcohol-related, and occurred in the city of El Paso.

<u>Definition of Alcohol-Related Crash</u>

A crash is defined as alcohol-related if item 38 *Contributing
Factors* in the [2023 Texas Peace Officer's Crash
Report](https://www.txdot.gov/data-maps/crash-reports-records/forms-law-enforcement.html)
contains code 45 (Had Been Drinking) or 67 (Intoxicated-Alcohol), or if
a driver's blood alcohol content (BAC) is reported as greater than 0.
This coding scheme aligns with the Alcohol-Related crash filter in the
TXDOT CRIS Query Tool.

<u>Definition of Injury and Fatality</u>

A person is defined as injured if item 16 in the [2023 Texas Peace Officer's
Crash
Report](https://www.txdot.gov/data-maps/crash-reports-records/forms-law-enforcement.html)
contains code A (Suspected Serious Injury), B (Suspected Minor Injury),
or C (Possible Injury). Fatalities are counted if a person involved in
the crash is coded K (Fatal Injury). The 2025 edition of the
[Instructions to Police for Reporting
Crashes](https://www.txdot.gov/content/dam/docs/division/trf/crash-records/cr100-v29.pdf)
by TXDOT indicates that code K is used if "a person dies within 30 days,
due to injuries sustained in the crash" (p. 26). In these cases, a
supplement report is submitted to update item 16.

<u>Definition of Intoxicated Driver</u>

Driver and vehicle characteristics on the [Drivers](#drivers) page
represent intoxicated drivers. An intoxicated driver is defined as a
person who's blood alcohol content test result returned a value of 0.08g/dL or greater.

<u>Acronyms</u>

BAC = Blood alcohol content; YTD = Year to date; TXDOT = Texas
Department of Transportation; CRIS = Crash Records Information System
:::

::: {.card title="**License and Use**"}
© 2026 El Paso Crime Lab.

This Alcohol-Related Crash Dashboard is licensed under the Creative Commons Attribution 4.0 International License (CC BY 4.0).
You are free to use, share, and adapt the charts in this dashboard for any purpose, including commercial use, provided that appropriate credit is given to El Paso Crime Lab. Attribution should include the project name, source, a link to the license (https://creativecommons.org/licenses/by/4.0/), and a clear indication of whether any modifications were made.

No additional restrictions apply beyond proper attribution.

Suggested citation: El Paso Crime Lab. (2026). *Alcohol-Related Crash Dashboard*.

<u>Key Notes for Reporting</u>

- Crashes, injuries, and fatalities in this dashboard represent those
  classified as alcohol-related (see definition)
- Crash data is publicly available and provided by the Texas Department of Transportation
- Monthly updates may revise prior counts as data are finalized

<u>Using Figures and Charts</u>

All figures in this dashboard are interactive and downloadable.

- Charts can easily be resized by
  adjusting the size of your internet browsing window. To expand a figure, click the “Expand” button located at the
  bottom-right corner of each chart.
- To download a figure, hover over the chart and click the camera icon
  to save it as an image

<u>Suggested Language</u>

- “Crash data on alcohol-related crashes in El Paso show…”
- “Local data on alcohol-related traffic fatalities indicate…”

<u>Contact</u>

For questions or data clarification, please contact:\
[mduran29\@asu.edu](mduran29@asu.edu)
:::

::: {.card title="**Technical Details**"}
<u>Calculation of Moving Average</u>

The line graphs on the [Trends](#trends) page show 12-month moving
averages for crashes, fatalities, and injuries between 2016 and
*Present*. This technique results in a smooth line, allowing the viewer
to see if a trend is increasing or decreasing.

<u>Holiday Analyses</u>

[Holidays](#holidays) statistics were calculated by aligning crash data
to each holiday. For fixed-date holidays (e.g., New Year’s Day,
Independence Day, Halloween, Christmas, and St. Patrick’s Day), each
crash was assigned an event time based on the number of days before or
after the calendar date of the holiday within each year. For shifting
holidays (e.g., Super Bowl and Memorial Day), the holiday date was first
computed for each year and then used as the reference point for
event-time calculation.

In all cases, crashes were restricted to a ±7-day window around the
holiday and then aggregated across years (2016–2025), excluding 2020 to
avoid COVID-19-related disruptions. Average daily crash counts were then
computed for each relative day to identify typical patterns before and
after each holiday.

<u>Missing Data</u>

A small amount of data for [Drivers](#drivers) characteristics (e.g.,
gender, race/ethnicity, vehicle color) is missing. Observations with
missing or unknown values in these fields were excluded.

<u>Limitations</u>

Not all alcohol-involved crashes are identified in official records. In
many cases, alcohol use may go undetected or unreported, particularly
when testing is not conducted or impairment is not clearly observed. As
a result, these data likely underestimate the true number of
alcohol-related crashes.

Despite this limitation, crash data remain one of the most reliable and
policy-relevant indicators of drunk driving. Unlike arrest data, which
are influenced by enforcement practices, crash records capture
real-world harm—injuries and fatalities—providing a consistent basis for
tracking trends over time and understanding the public safety impact of
impaired driving.
:::

# Trends {#trends}

## Row {height="200px"}

```{r}
#| content: valuebox
#| title: "Total Crashes (YTD)"
ytd_alcohol_crashes <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  summarise(
    ytd_alcohol_crashes = n_distinct(Crash.ID)
  ) %>%
  pull(ytd_alcohol_crashes)

list(
  icon = "car-front-fill",
  color = "primary",
  value = ytd_alcohol_crashes )
```

```{r}
#| content: valuebox
#| title: "Total People Injured (YTD)"

ytd_people_injury <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0
    )
  )

ytd_people_injury <- ytd_people_injury %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    injured == 1) %>%
    nrow()


list(
  icon = "bandaid",
  color = "warning",
  value = ytd_people_injury)

```

```{r}
#| content: valuebox
#| title: "Total Fatalities (YTD)"
ytd_people_fatal <- person_level %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    )
  )
ytd_alcohol_fatalities <- ytd_people_fatal %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    fatal == 1) %>%
    nrow()

list(
  icon = "hospital",
  color = "danger",
  value = ytd_alcohol_fatalities
)
```

## Row

```{r}
#| title: Crashes by Day of Week (YTD)
dow_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  count(Day.of.Week) %>%
  mutate(
    Day.of.Week = factor(
      tools::toTitleCase(tolower(Day.of.Week)),
      levels = c("Sunday", "Monday", "Tuesday", "Wednesday",
                 "Thursday", "Friday", "Saturday")
    )
  )

avg_val <- mean(dow_data$n, na.rm = TRUE)

p <- ggplot(dow_data, aes(
  x = Day.of.Week,
  y = n,
)) +

  geom_col(fill = "steelblue", width = 0.5) +

geom_text(
  aes(label = n, y = n + max(n) * 0.05),
  size = 3
) +

scale_y_continuous(
  limits = c(0, max(dow_data$n, na.rm = TRUE) * 1.3)
) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +

  labs(title = "Alcohol-Related Crashes in El Paso by Day of Week (Year-to-Date)",
       x = "", y = "Total crashes", fill = "") 

ggplotly(p) %>%
  layout(
      yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE
    ),
      margin = list(b = 60),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes in El Paso by Day of Week</b>",
      "<br><sup>Year-to-Date</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
    xaxis = list(
      showgrid = FALSE,
      zeroline = FALSE
    ),
    annotations = list(
      list(
        x = 0, y = -0.1,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )


```

```{r}
#| title: Crashes by Hour (YTD)

time_levels <- c(
  "12am – 3am", "3am – 6am", "6am – 9am", "9am – 12pm",
  "12pm – 3pm", "3pm – 6pm", "6pm – 9pm", "9pm – 12am"
)


hour_data <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  mutate(
    hour = as.numeric(substr(Hour.of.Day, 1, 2)),

    time_block = case_when(
      hour >= 0  & hour <= 2  ~ "12am – 3am",
      hour >= 3  & hour <= 5  ~ "3am – 6am",
      hour >= 6  & hour <= 8  ~ "6am – 9am",
      hour >= 9  & hour <= 11 ~ "9am – 12pm",
      hour >= 12 & hour <= 14 ~ "12pm – 3pm",
      hour >= 15 & hour <= 17 ~ "3pm – 6pm",
      hour >= 18 & hour <= 20 ~ "6pm – 9pm",
      hour >= 21 & hour <= 23 ~ "9pm – 12am",
      TRUE ~ NA_character_
    )
  ) %>%

  # count crashes per block
  count(time_block) %>%

  # force full structure (adds zeros)
  complete(time_block = time_levels, fill = list(n = 0)) %>%

  # enforce correct order
  mutate(
    time_block = factor(time_block, levels = time_levels),
    pct = n / sum(n),
    label = paste0(
      n,
      "<br>(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )


avg_block <- mean(hour_data$n, na.rm = TRUE)


p <- ggplot(hour_data, aes(
  x = time_block,
  y = n,
  text = label)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_text(
    aes(y =  n + max(n) * 0.05, label = round(n, 0)),
    size = 3.5
  ) +
  scale_y_continuous(
    limits = c(0, max(hour_data$n, na.rm = TRUE) * 1.3)) +
    labs(
      x = "",
      y = "Total crashes",
      title = "Alcohol-Related Crashes in El Paso by Hour (Year-to-Date)") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

ggplotly(p) %>%
  layout(
    yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE
    ),
     margin = list(b = 120),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes in El Paso by Hour</b>",
      "<br><sup>Year-to-Date</sup>"),
    x = 0.5,
    xanchor = "center"),
    xaxis = list(
      showgrid = FALSE,
      zeroline = FALSE),
    annotations = list(
      list(
        x = 0, y = -0.3,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )

```

## Row {.tabset height="300px"}

```{r}
#| title: Crash Trends

monthly_counts <- crash_level %>%
  filter(City == "El Paso") %>%  
  mutate(
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    crashes = n_distinct(Crash.ID),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    ma_12 = zoo::rollmean(crashes, k = 12, fill = NA, align = "right"))

p <- ggplot(monthly_counts, aes(x = month)) +
  geom_line(
    aes(y = crashes, color = "Total crashes"),
    linewidth = 0.8) +
  geom_line(
    aes(y = ma_12, color = "Average"),
    linewidth = 1.2) +
  scale_color_manual(
    values = c(
      "Total crashes" = "grey70",
      "Average" = "steelblue")) +
  labs(
    title = "Alcohol-Related Crashes in El Paso (2016 to Present)",
    x = "",
    y = "Total crashes",
    color = "") +
  theme_minimal() +
  theme(
    legend.position = "bottom")

ggplotly(p) %>%
  layout(
      title = list(
      text = paste0(
      "<b>Alcohol-Related Crashes in El Paso</b>",
      "<br><sup>2016–Present</sup>"),
    x = 0.5,
    xanchor = "center"),
    annotations = list(
      list(
        x = 0, y = -0.18,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )

```

```{r}
#| title: Injury Trends

injury_monthly <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0),
    month = floor_date(Crash.Date, "month")) %>%
  filter(
    City == "El Paso",
    injured == 1) %>%
  group_by(month) %>%
  summarise(
    injuries = n(),
    .groups = "drop") %>%
  arrange(month) %>%
  mutate(
    ma_12 = zoo::rollmean(injuries, k = 12, fill = NA, align = "right")
  )

p <- ggplot(injury_monthly, aes(x = month)) +
  geom_line(
    aes(y = injuries, color = "Total people injured"),
    linewidth = 0.8) +
  geom_line(
    aes(y = ma_12, color = "Average"),
    linewidth = 1.2) +
  scale_color_manual(
    values = c(
      "Total people injured" = "grey70", 
      "Average" = "orange")) +
  labs(
    title = "Alcohol-Related Crash Injuries in El Paso (2016 to Present)",
    x = "",
    y = "Total people injured",
    color = "") +
  theme_minimal() +
  theme(
    legend.position = "bottom")

ggplotly(p) %>%
  layout(
    title = list(
    text = paste0(
      "<b>Alcohol-Related Crash Injuries in El Paso</b>",
      "<br><sup>2016–Present</sup>"),
    x = 0.5,
    xanchor = "center"),
    annotations = list(
      list(
        x = 0, y = -0.18,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )

```

```{r}
#| title: Fatality Trends

monthly_counts <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0),
    month = floor_date(Crash.Date, "month")) %>%
  group_by(month) %>%
  summarise(
    alcohol_fatal_crashes = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop") %>%
  arrange(month) %>%
  mutate(
    alcohol_fatal_ma = zoo::rollmean(alcohol_fatal_crashes,
      k = 12,
      fill = NA,
      align = "right"))

p <- ggplot(monthly_counts, aes(x = month)) +

  geom_line(
    aes(y = alcohol_fatal_crashes, color = "Total fatalities"),
    linewidth = 1) +
  geom_line(
    aes(y = alcohol_fatal_ma, color = "Average"),
    linewidth = 1.2) +
  scale_color_manual(
    values = c(
      "Total fatalities" = "grey70",
      "Average" = "maroon")) +
  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, 2)) +
  labs(
    x = "",
    y = "Total fatalities",
    color = "",
    title = "Alcohol-Related Traffic Fatalities in El Paso (2016 to Present)") +
  theme_minimal() +
  theme(
    legend.position = "bottom")

ggplotly(p) %>%
  layout(
  title = list(
    text = paste0(
      "<b>Alcohol-Related Traffic Fatalities in El Paso</b>",
      "<br><sup>2016–Present</sup>"),
    x = 0.5,
    xanchor = "center"),
    annotations = list(
      list(
        x = 0, y = -0.18,
        xref = "paper", yref = "paper",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top",
        font = list(size = 10, color = "gray50")
      )
    )
  )
```

## Row {.tabset height="250px"}

```{r}
#| title: Average Crashes by Month

monthly_year <- crash_level %>%
  filter(City == "El Paso") %>%
  mutate(
    date = as.Date(Crash.Date),
    year = year(date),
    month_num = month(date)
  ) %>%
  group_by(year, month_num) %>%
  summarise(
    monthly_crashes = n_distinct(Crash.ID),
    .groups = "drop"
  )


monthly_avg <- monthly_year %>%
  group_by(month_num) %>%
  summarise(
    avg_crashes = mean(monthly_crashes),
    .groups = "drop"
  ) %>%
  
  mutate(
    pct = avg_crashes / sum(avg_crashes),
    month_name = factor(month.name[month_num], levels = month.name)
  )

overall_avg <- mean(monthly_avg$avg_crashes)

monthly_avg <- monthly_avg %>%
  mutate(
    label = paste0(
      sprintf("%.1f", avg_crashes),
      "\n(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )

p <- ggplot(monthly_avg, aes(
  x = month_name,
  y = avg_crashes,
  text = paste0(
    "Month: ", month_name,
    "<br>Crashes: ", round(avg_crashes, 0)  )
)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(y = avg_crashes + 1, label = round(avg_crashes, 0)),
    size = 3.5
  ) +
  geom_col(fill = "steelblue", width = 0.5) +
     labs(
      x = "",
      y = "Average crashes",
      title = "Average Alcohol-Related Crashes by Month in El Paso (2016 to Present)"
    ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1)
  )

ggplotly(p) %>%
  layout(
    margin = list(b = 60),
    yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE
    ),
    xaxis = list(
      showgrid = FALSE,
      zeroline = FALSE
    ),
  title = list(
    text = paste0(
      "<b>Average Alcohol-Related Crashes by Month in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
    annotations = list(
      list(
        x = 0, y = -0.2,  
        xref = "paper", yref = "paper",
        xanchor = "left",
        yanchor = "bottom",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        font = list(size = 10, color = "gray50")
      )
    )
  )


```

```{r}
#| title: Average Fatalities by Month

monthly_fatal_year <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    date = Crash.Date,
    year = year(date),
    month_num = month(date),
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    ),
  ) %>%
  group_by(year, month_num) %>%
  summarise(
    monthly_fatalities = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  )

monthly_fatal_avg <- monthly_fatal_year %>%
  group_by(month_num) %>%
  summarise(
    avg_fatalities = mean(monthly_fatalities),
    .groups = "drop"
  ) %>%
  mutate(
  month_name = factor(month.name[month_num], levels = month.name),
    pct = avg_fatalities / sum(avg_fatalities),
    label = paste0(
      sprintf("%.1f", avg_fatalities),
      "\n(",
      sprintf("%.1f", pct * 100),
      "%)"
    )
  )

overall_avg <- mean(monthly_fatal_avg$avg_fatalities)

p <- ggplot(monthly_fatal_avg, aes(
  x = month_name,
  y = avg_fatalities,
  text = label
)) +

  geom_col(fill = "maroon", width = 0.5) +

  geom_text(
    aes(y = avg_fatalities + 0.1, label = round(avg_fatalities, 1)),
    size = 3.5
  ) +
  scale_y_continuous(
    limits = c(0, max(monthly_fatal_avg$avg_fatalities, na.rm = TRUE) * 1.3)
  ) +  

  labs(
    x = "",
    y = "Average fatalities",
    title = "Average Alcohol-Related Traffic Fatalities by Month in El Paso (2016-Present)"
  ) +

  theme_minimal(base_size = 12) +

  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 0, hjust = 1)
  )

ggplotly(p) %>%
  layout(
    margin = list(b = 60),
    yaxis = list(
      showticklabels = FALSE,
      showgrid = TRUE,
      zeroline = TRUE
    ),
  title = list(
    text = paste0(
      "<b>Average Alcohol-Related Traffic Fatalities by Month in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
    xaxis = list(
      showgrid = FALSE,
      zeroline = FALSE
    ),
    annotations = list(
      list(
        x = 0, y = -0.2,
        xref = "paper", yref = "paper",
        xanchor = "left",
        yanchor = "bottom",
        text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
        showarrow = FALSE,
        font = list(size = 10, color = "gray50")
      )
    )
  )
```

# Holidays {#holidays}

## Row {height="500px"}

```{r}
#| title: New Year's Day

base <- crash_level %>%
  filter(City == "El Paso", Crash.Year != 2020) %>%
  mutate(
    date = as.Date(Crash.Date),
    year = Crash.Year 
  )


ny_window <- base %>%
  distinct(year) %>%
  mutate(
    holiday = as.Date(paste0(year, "-01-01"))
  ) %>%
  tidyr::expand_grid(event_time = -7:7) %>%
  mutate(date = holiday + event_time)

data <- ny_window %>%
  left_join(
    base %>% select(date, Crash.ID),
    by = "date"
  )

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups = "drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes, na.rm = TRUE), .groups = "drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  labs(
    x = "Days from New Year's Day",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around New Year’s Day in El Paso (All Years)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +
  labs(
    x = "Days from New Year's Day",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around New Year’s Day in El Paso (2016-Present)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),

  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),

  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),

  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around New Year’s Day in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),

  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)


```

```{r}
#| title: Superbowl 

superbowl <- base %>%
  distinct(year) %>%
  mutate(
    feb1 = as.Date(paste0(year, "-02-01")),
    holiday = feb1 + (7 - wday(feb1)) + 7
  )

data <- base %>%
  left_join(superbowl, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +

  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
  scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +

  scale_x_continuous(
    breaks = seq(-7, 7, by = 1) 
  ) +

  labs(
    x = "Days from Superbowl",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Superbowl in El Paso (2016-Present)"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),

  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),

  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),

  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Superbowl in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),

  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)
```

## Row {height="500px"}

```{r}
#| title: St. Patrick's Day

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-03-17")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1) 
  ) +
  labs(
    x = "Days from Saint Patrick's Day",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Saint Patrick's Day in El Paso (2016-Present)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),
  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Saint Patrick's Day in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)
```

```{r}
#| title: Memorial Day

memorial <- base %>%
  distinct(year) %>%
  mutate(
    memorial_day = as.Date(paste0(year, "-05-31")) -
      (wday(as.Date(paste0(year, "-05-31")), week_start = 1) - 1)
  )

data <- base %>%
  left_join(memorial, by="year") %>%
  mutate(event_time = as.numeric(date - memorial_day)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
  scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +

  labs(
    x = "Days from Memorial Day",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Memorial Day in El Paso (2016-Present)"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),

  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Memorial Day in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
) 
```

## Row {height="500px"}

```{r}
#| title: Cinco de Mayo

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-05-05")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
  scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1) 
  ) +

  labs(
    x = "Days from Cinco de Mayo",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Cinco de Mayo in El Paso (2016-Present)"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),

  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),

  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Cinco de Mayo in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),

  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)

```

```{r}
#| title: Independence Day

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-07-04")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +

  labs(
    x = "Days from Independence Day",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Independence Day in El Paso (2016-Present)"
  ) +

  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(
  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),
  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Independence Day in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)
```

## Row {height="500px"}

```{r}
#| title: Labor Day

labor <- base %>%
  distinct(year) %>%
  mutate(
    sept1 = as.Date(paste0(year, "-09-01")),
    holiday = sept1 + (8 - wday(sept1, week_start=1)) %% 7
  )

data <- base %>%
  left_join(labor, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +
  labs(
    x = "Days from Labor Day",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Labor Day in El Paso (2016-Present)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(
  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),
  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),

  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Labor Day in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),

  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
) 
```

```{r}
#| title: Halloween

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-10-31")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +
  labs(
    x = "Days from Halloween",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Halloween in El Paso (2016-Present)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),

  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),

  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Halloween in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),

  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)
```

## Row {height="500px"}

```{r}
#| title: Thanksgiving

thanksgiving <- base %>%
  distinct(year) %>%
  mutate(
    nov1 = as.Date(paste0(year, "-11-01")),
    holiday = nov1 + (4 - wday(nov1) + 7) %% 7 + 21
  )

data <- base %>%
  left_join(thanksgiving, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +
  labs(
    x = "Days from Thanksgiving",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Thanksgiving in El Paso (2016-Present)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplotly(p) %>%
layout(
  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),
  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Thanksgiving in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)
```

```{r}
#| title: Christmas

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-12-25")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +

  geom_col(fill = "steelblue", width = 0.8) +
  geom_text(
    aes(label = sprintf("%.1f", avg_crashes),
        y = avg_crashes + 0.3),
    size = 3.5
  ) +
scale_y_continuous(
  limits = c(0, 6),
  breaks = seq(0, 6, by = 1)
) +
  scale_x_continuous(
    breaks = seq(-7, 7, by = 1)
  ) +
  labs(
    x = "Days from Christmas",
    y = "Average Crashes",
    title = "Alcohol-Related Crashes Around Christmas in El Paso (2016-Present)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

ggplotly(p) %>%
layout(

  yaxis = list(
    showticklabels = FALSE,
    showgrid = TRUE,
    zeroline = TRUE
  ),
  xaxis = list(
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-8, 8),
    dtick = 1
  ),
  margin = list(
    t = 100,
    b = 120,
    l = 40,
    r = 30
  ),
  title = list(
    text = paste0(
      "<b>Alcohol-Related Crashes Around Christmas in El Paso</b>",
      "<br><sup>2016–Present</sup>"
    ),
    x = 0.5,
    xanchor = "center"
  ),
  annotations = list(
    list(
      x = 0, y = -0.5,
      xref = "paper", yref = "paper",
      xanchor = "left",
      yanchor = "bottom",
      text = 'Data source: Texas Department of Transportation | Created by <a href="https://epcrimelab.org" target="_blank">El Paso Crime Lab</a>',
      showarrow = FALSE,
      font = list(size = 10, color = "gray50")
    )
  )
)
```

# Drivers {#drivers}

```{r}
# drivers object with BACs only
drivers <- person_level %>%
  mutate(
    driver_flag = case_when(
      Person.Type %in% c(
        "1 - DRIVER",
        "5 - DRIVER OF MOTORCYCLE TYPE VEHICLE") ~ 1,
      TRUE ~ 0),
    bac_value = as.numeric(Person.Blood.Alcohol.Content.Test.Result),
    bac_result = case_when(
      !is.na(bac_value) & bac_value > 0 ~ 1,
      TRUE ~ 0)) %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    driver_flag == 1,)
```

## Row {height="200px"}

```{r}
#| content: valuebox
#| title: "Total Intoxicated Drivers (YTD)"

total_intox_drivers <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  summarise(n = n()) %>%
  pull(n)

list(
  icon = "person-square",
  color = "primary",
  value = format(total_intox_drivers, big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Average BAC Result (YTD)"
bac_stats <- drivers %>%
  filter(bac_result == 1) %>%
  summarise(
    mean_bac = mean(bac_value, na.rm = TRUE)
  )
list(
  icon = "moisture",
  color = "danger",
  value = paste0(
    round(bac_stats$mean_bac, 2)
  )
)
```

```{r}
#| content: valuebox
#| title: "Average Age (YTD)"

age_stats <- drivers %>%
  filter(bac_value>=0.08 & bac_result==1) %>%
  mutate(
    age_num = as.numeric(Person.Age)
  ) %>%
  summarise(
    mean_age = mean(age_num, na.rm = TRUE),
    sd_age = sd(age_num, na.rm = TRUE)
  )

list(
  icon = "rulers",
  color = "primary",
  value = paste0(
    round(age_stats$mean_age, 0)
  )
)
```

## Row {height="70%"}

### Column {.tabset}

```{r}
#| title: Driver's License Zipcode (YTD)

zip_top10 <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1,
         !is.na(Driver.Zip.Code)) %>%
  group_by(Driver.Zip.Code) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  slice_head(n = 10)

zip_top10 <- zip_top10 %>%
  mutate(
    pct = n / sum(n),
    label_full = paste0(Driver.Zip.Code, "\n", n, " (", scales::percent(pct), ")")
  )

p <- plot_ly(
  data = zip_top10,
  type = "treemap",
  labels = ~label_full,
  parents = rep("", nrow(zip_top10)),
  values = ~n,
  textinfo = "label",
  marker = list(
    colors = rep("#f4b942", nrow(zip_top10))
  ),
  hovertemplate = paste(
    "ZIP: %{label}<br>",
    "Drivers: %{value}<br>",
    "<extra></extra>"
  )
)  

p %>%
  layout(
    title = list(
    text = paste0(
      "<b>Top 10 License ZIP Codes of Intoxicated Drivers Who Crashed in El Paso</b>",
      "<br><sup>Year-to-Date (BAC ≥ 0.08)</sup>"
    ),
      x = 0.5,
      xanchor = "center"
    ),
    margin = list(
      l = 10,
      r = 10,
      t = 70,
      b = 80   # space for caption / footer
    ),
    annotations = list(
      list(
        x = 0, y = -0.08,
        xref = "paper", yref = "paper",
        xanchor = "left",
        yanchor = "top",
        text = 'Data source: Texas Department of Transportation | Created by El Paso Crime Lab',
        showarrow = FALSE,
        font = list(size = 10, color = "gray50")
      )
    )
  )

```

```{r}
#| title: Driver's Demographics (YTD)

drivers_clean <- drivers %>%
  mutate(
    Person.Ethnicity = case_when(
      
      # Missing / unknown categories
      Person.Ethnicity %in% c("No Data", "99 - UNKNOWN", "UNKNOWN", "", NA) ~ NA_character_,
      
      # Special cases
      Person.Ethnicity == "95 - AUTONOMOUS" ~ "Autonomous",
      Person.Ethnicity == "98 - OTHER" ~ "Other",
      
      # Race/Ethnicity categories
      grepl("A - ASIAN", Person.Ethnicity) ~ "Asian",
      grepl("B - BLACK", Person.Ethnicity) ~ "Black",
      grepl("H - HISPANIC", Person.Ethnicity) ~ "Hispanic",
      grepl("I - AMER", Person.Ethnicity) ~ "American Indian/Alaska Native",
      grepl("W - WHITE", Person.Ethnicity) ~ "White",
      
      # Fallback
      TRUE ~ NA_character_
    ),
    
    Person.Gender = case_when(
      Person.Gender %in% c("99 - UNKNOWN", "UNKNOWN", "No Data", "") ~ NA_character_,
      Person.Gender == "1 - MALE" ~ "Male",
      Person.Gender == "2 - FEMALE" ~ "Female",
      TRUE ~ NA_character_
    )
  )

demo_table <- drivers_clean %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  filter(!is.na(Person.Ethnicity), !is.na(Person.Gender)) %>%
    mutate(
    Person.Gender = factor(Person.Gender, levels = c("Female", "Male")),
    Person.Age = as.numeric(Person.Age)
  ) %>%
  group_by(Person.Ethnicity, Person.Gender) %>%
  summarise(
    avg_bac = mean(bac_value, na.rm = TRUE),
    avg_age = mean(Person.Age, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    pct = scales::percent(n / sum(n), accuracy = 0.1),
    avg_bac = round(avg_bac, 2),
    avg_age = round(avg_age, 0)
  ) %>%
  arrange(desc(n)) %>%
  rename(
    "Race/Ethnicity" = Person.Ethnicity,
    "Sex" = Person.Gender,
    "Average Age" = avg_age,
    "Average BAC" = avg_bac,
    "Count" = n,
    "Percent" = pct)

reactable(
  demo_table,

  defaultPageSize = 10,
  searchable = TRUE,
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  columns = list(
    `Average BAC` = colDef(format = colFormat(digits = 2)),
    `Average Age` = colDef(format = colFormat(digits = 0)),
    Count = colDef(
      align = "center"
    ),
    Percent = colDef(
      align = "center")),
    defaultSorted = list(Count = "desc"))

```

### Column {.tabset}

```{r}
#| title: Vehicle Color (YTD)

plot_data <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  mutate(
    color = case_when(
      grepl("BLK", Vehicle.Color) ~ "Black",
      grepl("WHI", Vehicle.Color) ~ "White",
      grepl("GRY", Vehicle.Color) ~ "Gray",
      grepl("RED", Vehicle.Color) ~ "Red",
      grepl("BLU", Vehicle.Color) ~ "Blue",
      grepl("GRN", Vehicle.Color) ~ "Green",
      grepl("SIL", Vehicle.Color) ~ "Silver",
      grepl("TAN", Vehicle.Color) ~ "Tan",
      grepl("MAR", Vehicle.Color) ~ "Maroon",
      grepl("YEL", Vehicle.Color) ~ "Yellow",
      grepl("GLD", Vehicle.Color) ~ "Gold",
      grepl("PLE", Vehicle.Color) ~ "Purple",
      grepl("BGE", Vehicle.Color) ~ "Beige",
      TRUE ~ "Other"
    )
  ) %>%
  count(color) %>%
  arrange(desc(n)) %>%   
  mutate(
    color = factor(color, levels = rev(color)), 
    label = scales::comma(n)
  )

color_map <- c(
  "Black" = "#2C3E50",
  "White" = "#ECF0F1",
  "Gray" = "#95A5A6",
  "Red" = "#E74C3C",
  "Blue" = "#3498DB",
  "Green" = "#2ECC71",
  "Silver" = "#BDC3C7",
  "Tan" = "#D2B48C",
  "Maroon" = "#800000",
  "Yellow" = "#F1C40F",
  "Gold" = "#F39C12",
  "Purple" = "#9B59B6",
  "Beige" = "#F5F5DC",
  "Other" = "#7F8C8D"
)

p <- plot_ly(
  data = plot_data,
  x = ~n,
  y = ~color,
  type = "bar",
  orientation = "h",

  marker = list(
    color = color_map[as.character(plot_data$color)]
  ),

  text = ~label,
  textposition = "outside",
  cliponaxis = FALSE,

  hovertemplate = paste(
    "Color: %{y}<br>",
    "Drivers: %{x}<br>",
    "<extra></extra>"
  )
) %>%

  layout(

    title = list(
      text = paste0(
        "<b>Vehicle Color of Intoxicated Drivers Who Crashed in El Paso</b>",
        "<br><sup>Year-to-Date (BAC ≥ 0.08)</sup>"
      ),
      x = 0.5,
      xanchor = "center"
    ),
    xaxis = list(
      title = "Total intoxicated drivers",
      range = c(0, max(plot_data$n) * 1.2)
    ),
    yaxis = list(
      title = "",
      automargin = TRUE
    ),
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    font = list(
      family = "Arial, sans-serif",
      size = 14,
      color = "#2c3e50"
    ),
    margin = list(
      t = 90,
      r = 140,
      l = 80,
      b = 100
    ),
    annotations = list(
      list(
        x = 0, y = -0.12,
        xref = "paper", yref = "paper",
        xanchor = "left",
        yanchor = "top",
        text = 'Data source: Texas Department of Transportation | Created by El Paso Crime Lab',
        showarrow = FALSE,
        font = list(size = 10, color = "gray50")
      )
    )
  )

p

```

```{r}
#| title: Vehicle Info (YTD)

vehicle_table <- drivers %>%
  
  # keep BAC filter
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  
  # only remove No Data for Make + Model
  filter(
    !Vehicle.Make %in% c("No Data", "No data", "", NA),
    !Vehicle.Model.Name %in% c("No Data", "No data", "", NA),
    !is.na(Vehicle.Make),
    !is.na(Vehicle.Model.Name)
  ) %>%
  
  mutate(
    # Make cleanup
    make = case_when(
      toupper(Vehicle.Make) %in% c("GMC", "BMW") ~ toupper(Vehicle.Make),
      TRUE ~ str_to_title(tolower(Vehicle.Make))
    ),
    # Model cleanup
    model = Vehicle.Model.Name %>%
      str_remove("\\s*\\(.*\\)") %>%
      str_to_sentence()
  ) %>%
  group_by(make, model) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    pct = scales::percent(n / sum(n), accuracy = 0.1)
  ) %>%
  arrange(desc(n))

table <- reactable(
  vehicle_table,
  searchable = TRUE,
  sortable = TRUE,
  pagination = TRUE,
  defaultPageSize = 10,
  striped = TRUE,
  highlight = TRUE,
  compact = TRUE,
  columns = list(
    make = colDef(name = "Make"),
    model = colDef(name = "Model"),
    n = colDef(name = "Count"),
    pct = colDef(name = "Percent")),
  defaultSorted = list(n = "desc"))

table
```

# Maps {#maps}

## Row {height="100%"}

### Column {.tabset width="70%"}

``` {r}
#| title: Peak Risk for Crashes (YTD)
#| echo: false


today <- Sys.Date()
start_2026 <- as.Date("2026-01-01")


crash_2026 <- crash_level %>%
  mutate(
    crash_date = as.Date(Crash.Date),  # <-- adjust if your date field differs
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    hour = as.numeric(substr(Hour.of.Day, 1, 2))
  ) %>%
  filter(
    crash_date >= start_2026,
    crash_date <= today
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

blocks <- st_read(
  "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/OCEPTX/shape files/Census Block Groups/tl_2025_48_bg.shp",
  quiet = TRUE
) %>%
  st_make_valid()

crash_sf <- st_as_sf(crash_2026, coords = c("Longitude", "Latitude"), crs = 4326)
blocks <- st_transform(blocks, st_crs(crash_sf))

joined <- st_join(crash_sf, blocks, join = st_within)

joined <- joined %>%
  mutate(
    time_block = case_when(
      hour >= 0  & hour <= 2  ~ "12am–3am",
      hour >= 3  & hour <= 5  ~ "3am–6am",
      hour >= 6  & hour <= 8  ~ "6am–9am",
      hour >= 9  & hour <= 11 ~ "9am–12pm",
      hour >= 12 & hour <= 14 ~ "12pm–3pm",
      hour >= 15 & hour <= 17 ~ "3pm–6pm",
      hour >= 18 & hour <= 20 ~ "6pm–9pm",
      hour >= 21 & hour <= 23 ~ "9pm–12am"
    )
  )

tbl_time <- joined %>%
  st_drop_geometry() %>%
  group_by(GEOID, time_block) %>%
  summarise(n = n(), .groups = "drop")

peak_time <- tbl_time %>%
  group_by(GEOID) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(GEOID, peak_time = time_block, peak_crashes = n)

geo_totals <- joined %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarise(total_crashes = n(), .groups = "drop")

top_geoids <- geo_totals %>%
  arrange(desc(total_crashes)) %>%
  slice_head(n = 10) %>%
  pull(GEOID)

blocks_map <- blocks %>%
  filter(GEOID %in% top_geoids) %>%
  left_join(geo_totals, by = "GEOID") %>%
  left_join(peak_time, by = "GEOID") %>%
  mutate(
    block_group = substr(GEOID, nchar(GEOID), nchar(GEOID)),
    popup_text = paste0(
      "<b>Block Group:</b> ", block_group, "<br/>",
      "<b>GEOID:</b> ", GEOID, "<br/>",
      "<b>Total Crashes (YTD):</b> ", total_crashes, "<br/>",
      "<b>Peak Risk Time:</b> ", peak_time, "<br/>",
      "<b>Peak Period Crashes:</b> ", peak_crashes
    ),
    hover_text = paste0(
      "BG ", block_group,
      " | Crashes: ", total_crashes,
      " | Peak: ", peak_time
    )
  )

max_val <- max(blocks_map$total_crashes, na.rm = TRUE)

pal <- colorNumeric(
  palette = colorRampPalette(c("#f7fbff", "steelblue"))(100),
  domain = c(0, max_val),
  na.color = "transparent"
)

leaflet(blocks_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(total_crashes),
    fillOpacity = 0.85,
    color = "white",
    weight = 1,
    label = ~lapply(hover_text, htmltools::HTML),
    labelOptions = labelOptions(
      sticky = TRUE,
      direction = "auto"
    ),
    popup = ~popup_text
  ) %>%
  addControl(
    html = paste0(
      "<strong>Top 10 Alcohol-Related Crash Hotspots in El Paso (Year-to-Date)</strong><br/>"),
    position = "topright"
  )
```

``` {r}
#| title: Crashes by Police Region (YTD)
#| echo: false

regions <- st_read(
  "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/OCEPTX/shape files/Police Regions/PoliceRegions.shp",   quiet = TRUE
) %>%
  st_make_valid()

crash_pts <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

crash_sf <- st_as_sf(crash_pts, coords = c("Longitude", "Latitude"), crs = 4326)
regions <- st_transform(regions, st_crs(crash_sf))

joined <- st_join(crash_sf, regions, join = st_within)

counts <- joined %>%
  st_drop_geometry() %>%
  group_by(DISTRICT) %>%
  summarise(Crashes = n(), .groups = "drop")

regions_map <- regions %>%
  left_join(counts, by = "DISTRICT") %>%
  mutate(Crashes = ifelse(is.na(Crashes), 0, Crashes))

max_val <- max(regions_map$Crashes, na.rm = TRUE)

pal <- colorNumeric(
  palette = colorRampPalette(c("#f7fbff", "steelblue4"))(100),
  domain = c(0, max_val),
  na.color = "transparent"
)

leaflet(regions_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addPolygons(
    fillColor = ~pal(Crashes),
    fillOpacity = 0.8,
    color = "white",
    weight = 1,
    popup = ~paste0(
      "<strong>District:</strong> ", DISTRICT,
      "<br><strong>Crashes:</strong> ", Crashes
    )
  ) %>%
  
  addLegend(
    pal = pal,
    values = regions_map$Crashes,
    title = "Total crashes",
    position = "bottomright",
    labFormat = labelFormat(suffix = ""),
    opacity = 1
  ) %>%

  addControl(
    html = paste0(
      "<strong>Alcohol-Related Crashes by Police Region (Year-to-Date)</strong><br/>"),
    position = "topright"
  )

```

``` {r}
#| title: Crashes by City Districts (YTD)
#| echo: false

districts <- st_read(
  "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/OCEPTX/shape files/City Districts/RepresentativeDistricts.shp",   quiet = TRUE) %>%
  st_make_valid()

crash_pts <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

crash_sf <- st_as_sf(crash_pts, coords = c("Longitude", "Latitude"), crs = 4326)
districts <- st_transform(districts, st_crs(crash_sf))

joined <- st_join(crash_sf, districts, join = st_within)

counts <- joined %>%
  st_drop_geometry() %>%
  group_by(DISTRICT) %>%
  summarise(Crashes = n(), .groups = "drop")

districts_map <- districts %>%
  left_join(counts, by = "DISTRICT") %>%
  mutate(Crashes = ifelse(is.na(Crashes), 0, Crashes))

max_val <- max(districts_map$Crashes, na.rm = TRUE)

pal <- colorNumeric(
  palette = colorRampPalette(c("#f7fbff", "steelblue4"))(100),
  domain = c(0, max_val),
  na.color = "transparent"
)

leaflet(districts_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addPolygons(
    fillColor = ~pal(Crashes),
    fillOpacity = 0.8,
    color = "white",
    weight = 1,
    popup = ~paste0(
      "<strong>District:</strong> ", DISTRICT,
      "<br><strong>Crashes:</strong> ", Crashes
    )
  ) %>%
  
  addLegend(
    pal = pal,
    values = districts_map$Crashes,
    title = "Total crashes",
    position = "bottomright",
    labFormat = labelFormat(suffix = ""),
    opacity = 1
  ) %>%
  
  addControl(
    html = paste0(
      "<strong>Alcohol-Related Crashes by City District in El Paso (Year-to-Date)</strong><br/>"),
    position = "topright"
  )

```

```{r}
#| title: Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "black"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  ) %>%
  addControl(
    html = paste0(
      "<strong>Alcohol-Related Crashes in El Paso (Year-to-Date)</strong><br/>"),
    position = "topright"
  )
```

```{r}
#| title: Fatal Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026,
         Fatal.Crash.Flag == 1) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "maroon"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  ) %>%
  addControl(
    html = paste0(
      "<strong>Fatal Alcohol-Related Crashes in El Paso (Year-to-Date)</strong><br/>"),
    position = "topright"
  )
```

### Column {.tabset width="30%"}

```{r}
#| title: Crash Locations (YTD)
top_streets <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    Street.Name = case_when(
      str_detect(Street.Name, " ") ~ str_to_title(str_to_lower(Street.Name)),
      TRUE ~ Street.Name
    ),
    Street.Name = str_replace_all(Street.Name, "\\bDr\\b", "Dr"),
    Street.Name = str_replace_all(Street.Name, "\\bBlvd\\b", "Blvd"),
    Street.Name = str_replace_all(Street.Name, "\\bCt\\b", "Ct"),
    Street.Name = str_replace_all(Street.Name, "\\bRd\\b", "Rd"),
    Street.Name = str_replace_all(Street.Name, "\\bSt\\b", "St"),
    Street.Name = str_replace_all(Street.Name, "\\bAve\\b", "Ave"),
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Street Name" = Street.Name,
    "Crash Count" = n,
    "Percent (%)" = percent
  )

reactable(
  top_streets,
  defaultPageSize = 15,
  striped = TRUE,
  highlight = TRUE,
  searchable = TRUE,
  defaultSorted = list(`Crash Count` = "desc"),
  columns = list(
    `Street Name` = colDef(minWidth = 200),
    `Crash Count` = colDef(align = "center"),
    `Percent (%)` = colDef(
      format = colFormat(digits = 2),
      align = "center"
    )
  ),
  compact = TRUE,
  bordered = FALSE
)
```

```{r}
#| title: Fatal Crash Locations (2016-Present)

top_fatal_streets <- crash_level %>%
  filter(
    Fatal.Crash.Flag == 1
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    Street.Name = case_when(
      str_detect(Street.Name, " ") ~ str_to_title(str_to_lower(Street.Name)),
      TRUE ~ Street.Name
    ),
    
    Street.Name = str_replace_all(Street.Name, "\\b(N|S|E|W|NE|NW|SE|SW)\\b", toupper),
    Street.Name = str_replace_all(Street.Name, "\\bDr\\b", "Dr"),
    Street.Name = str_replace_all(Street.Name, "\\bBlvd\\b", "Blvd"),
    Street.Name = str_replace_all(Street.Name, "\\bCt\\b", "Ct"),
    Street.Name = str_replace_all(Street.Name, "\\bRd\\b", "Rd"),
    Street.Name = str_replace_all(Street.Name, "\\bSt\\b", "St"),
    Street.Name = str_replace_all(Street.Name, "\\bAve\\b", "Ave"),
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Street Name" = Street.Name,
    "Fatal Crashes" = n,
    "Percent (%)" = percent
  )

reactable(
  top_fatal_streets,
  defaultPageSize = 10,
  striped = TRUE,
  highlight = TRUE,
  searchable = TRUE,
  defaultSorted = list(`Fatal Crashes` = "desc"),
  columns = list(
    `Street Name` = colDef(minWidth = 220),
    `Fatal Crashes` = colDef(align = "center"),
    `Percent (%)` = colDef(
      format = colFormat(digits = 2),
      align = "center"
    )
  ),
  compact = TRUE,
  bordered = FALSE
)
```

# Quiz {#quiz}

:::::: {.column width="100%"}
::::: {.card title="Drunk Driving Quiz" style="min-height:780px; margin:0 auto;"}
:::: {#quizApp}
<h4 id="introStats">

Loading stats...

</h4>

<h4 id="avgStats">

</h4>

<button onclick="startQuiz()" id="startBtn">Start Quiz</button>

<p id="tagline" style="font-style: italic;">

</p>

<hr>

::: {#quizBox style="display:none"}
:::

</div>

```{=html}
<script>
const SCRIPT_URL = "https://script.google.com/macros/s/AKfycbw4TekPtUfV08eUUU9TbRu0QdeUl9_mondBZS0c179w-eQQb4YliAQwwUHyODuErsYA/exec";

let qIndex = 0;
let score = 0;

/* ---------------- QUIZ ---------------- */
const quiz = [

  {
    q: "What day(s) of the week do alcohol-related crashes most often occur?",
    a: ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Weekends (Saturday and Sunday)"],
    correct: "Weekends (Saturday and Sunday)",
    explanation: "About 50% of all alcohol-related crashes in El Paso occur on the weekend."
  },

  {
    q: "What time of day do alcohol-related crashes most often occur?",
    a: ["10:00 AM to 5:00 PM", "12:00 PM to 6:00 PM", "6:00 PM to 11:00 PM", "9:00 PM to 3:00 AM"],
    correct: "9:00 PM to 3:00 AM",
    explanation: "More than half of all alcohol-related crashes in El Paso occur during late-night and early morning hours."
  },

  {
    q: "Since 2016, the total number of alcohol-related crashes in El Paso has decreased.",
    a: ["True", "False"],
    correct: "True",
    explanation: "Overall trends show a decline in alcohol-related crashes since 2016. The highest number of alcohol-related crashes recorded was in September 2017."
  },

  {
    q: "What is the legal blood alcohol concentration (BAC) limit for driving in Texas?",
    a: ["0.05%", "0.08%", "0.10%"],
    correct: "0.08%",
    explanation: "In Texas, it is illegal to drive with a BAC of 0.08% or higher. This level indicates significant impairment in reaction time and judgment."
  },

  {
    q: "On average, what blood alcohol concentration (BAC) level do drivers involved in alcohol-related crashes in El Paso have?",
    a: ["Less than the legal limit (<0.08%)", "Around the legal limit (~0.08%)", "More than twice the legal limit (>0.16%)"],
    correct: "More than twice the legal limit (>0.16%)",
    explanation: "The data shows that the average BAC among drivers who crashed and were tested for alcohol is greater than 0.16, which is more than twice the legal limit and associated with severe impairment."
  },

  {
    q: "On which holiday do crashes sharply increase in El Paso?",
    a: ["Memorial Day", "Christmas", "New Years Day"],
    correct: "New Years Day",
    explanation: "The data shows the average number of alcohol-involved crashes on New Years Day doubles compared to the previous day."
  },

  {
    q: "Which roadway has the most alcohol-related traffic fatalities in El Paso?",
    a: ["Mesa Street", "Zaragoza Road", "I-10"],
    correct: "I-10",
    explanation: "The majority of crashes and fatalities occur along I-10 in El Paso, likely due to higher speeds and increased crash severity when alcohol is involved."
  }

];

/* ---------------- STATS ---------------- */
function loadStats() {

  fetch(SCRIPT_URL)
    .then(r => r.text())
    .then(t => {

      const data = JSON.parse(t);
      data.shift();

      const scores = data
        .map(r => Number(r[3]))
        .filter(v => !isNaN(v));

      const count = scores.length;

      const avg = count
        ? (scores.reduce((a,b)=>a+b,0)/count).toFixed(1)
        : 0;

      document.getElementById("introStats").innerHTML =
        `${count} people have taken this quiz.`;

      document.getElementById("avgStats").innerHTML =
        `Average score: ${avg}%`;
    });
}

/* ---------------- START ---------------- */
function startQuiz() {
  document.getElementById("startBtn").style.display = "none";
  document.getElementById("quizBox").style.display = "block";
  loadQuestion();
}

/* ---------------- LOAD QUESTION ---------------- */
function loadQuestion() {

  const q = quiz[qIndex];
  const box = document.getElementById("quizBox");

  box.innerHTML = `
    <h4>Question ${qIndex + 1} of ${quiz.length}</h4>
    <p>${q.q}</p>
    <div id="answers"></div>
    <p id="feedback"></p>
    <button id="nextBtn" style="display:none" onclick="nextQuestion()">Next</button>
  `;

  const answersDiv = document.getElementById("answers");

  q.a.forEach(choice => {

    const btn = document.createElement("button");
    btn.innerText = choice;

    btn.onclick = () => {

      const buttons = answersDiv.querySelectorAll("button");
      buttons.forEach(b => b.disabled = true);

      if (choice === q.correct) {
        score++;
        btn.style.backgroundColor = "#c8f7c5";

        document.getElementById("feedback").innerHTML =
          `<span style="color:green;"><b>Correct</b></span><br><br>${q.explanation}`;
      } else {
        btn.style.backgroundColor = "#f7c5c5";

        buttons.forEach(b => {
          if (b.innerText === q.correct) {
            b.style.backgroundColor = "#c8f7c5";
          }
        });

        document.getElementById("feedback").innerHTML =
          `<b>Correct answer:</b> ${q.correct}<br><br>${q.explanation}`;
      }

      document.getElementById("nextBtn").style.display = "inline-block";
    };

    answersDiv.appendChild(btn);
  });
}

/* ---------------- NEXT ---------------- */
function nextQuestion() {
  qIndex++;

  if (qIndex < quiz.length) {
    loadQuestion();
  } else {
    finishQuiz();
  }
}

/* ---------------- FINISH ---------------- */
function finishQuiz() {

  const percent = Math.round((score / quiz.length) * 100);

  fetch(SCRIPT_URL, {
    method: "POST",
    body: JSON.stringify({
      timestamp: new Date().toISOString(),
      score: percent
    })
  });

  document.getElementById("quizBox").innerHTML =
    `<h3>Finished</h3>
     <p>Your score: <b>${percent}%</b></p>`;
}

/* ---------------- INIT ---------------- */
document.addEventListener("DOMContentLoaded", loadStats);
</script>
```

