---
title: "Alcohol-Related Crash Dashboard"
format:
  dashboard:
    theme: default
    css: styles2.css
    orientation: rows
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
---

Welcome to the Alcohol-Related Crash Dashboard!

This dashboard displays data on alcohol-related vehicle crashes in El Paso, Texas and is updated once a month.

```{r}

rm(list = ls())

suppressMessages(library(gt))
suppressMessages(library(quarto))
suppressMessages(library(rsconnect))
suppressMessages(library(shiny))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
suppressMessages(library(stringr))
suppressMessages(library(readr))
suppressMessages(library(tidyverse))
suppressMessages(library(leaflet))
suppressMessages(library(leaflet.extras))


files <- c(
  austin = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/austin_workingdata.csv",
  brownsville = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/brownsville_workingdata.csv",
  dallas = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/dallas_workingdata.csv",
  elpaso = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/elpaso_workingdata.csv",
  fortworth = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/fortworth_workingdata.csv",
  houston = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/houston_workingdata.csv",
  laredo = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/laredo_workingdata.csv",
  mcallen = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/mcallen_workingdata.csv",
  sanantonio = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/sanantonio_workingdata.csv"
)

all_cities <- bind_rows(lapply(names(files), function(city) {
  df <- read.csv(files[[city]], header = TRUE)
  df$city <- city
  df
}))

all_cities$City <- recode(all_cities$City,
  "AUSTIN" = "Austin",
  "BROWNSVILLE" = "Brownsville",
  "DALLAS" = "Dallas",
  "EL PASO" = "El Paso",
  "FORT WORTH" = "Fort Worth",
  "HOUSTON" = "Houston",
  "MCALLEN" = "McAllen",
  "SAN ANTONIO" = "San Antonio",
  "LAREDO" = "Laredo"
)

# drop current month, since its incomplete
all_cities <- all_cities %>%
  mutate(
    Crash.Date = mdy(Crash.Date)
  ) %>%
  filter(
    !(year(Crash.Date) == year(Sys.Date()) &
      month(Crash.Date) == month(Sys.Date()))
  )

person_level <- all_cities

vehicle_level <- all_cities %>%
  distinct(Crash.ID, Unit.Number, .keep_all = TRUE)

crash_level <- all_cities %>%
  distinct(Crash.ID, .keep_all = TRUE) 
```

# Trends


## Row {height="200px"}

```{r}
#| content: valuebox
#| title: "Total Crashes (YTD)"
ytd_alcohol_crashes <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  summarise(
    ytd_alcohol_crashes = n_distinct(Crash.ID)
  ) %>%
  pull(ytd_alcohol_crashes)

list(
  icon = "car-front-fill",
  color = "primary",
  value = ytd_alcohol_crashes )
```

```{r}
#| content: valuebox
#| title: "Total People Injured (YTD)"

person_level <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0
    )
  )

ytd_people_alcohol <- person_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    injured == 1) %>%
    nrow()


list(
  icon = "bandaid",
  color = "primary",
  value = ytd_people_alcohol)

```

```{r}
#| content: valuebox
#| title: "Total Fatalities (YTD)"
person_level <- person_level %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    )
  )
ytd_alcohol_fatalities <- person_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    fatal == 1) %>%
    nrow()

list(
  icon = "hospital",
  color = "danger",
  value = ytd_alcohol_fatalities
)
```

## Row {.tabset height="400px"}

``` {r}
#| title: Crash Trends
monthly_counts <- crash_level %>%
  filter(City == "El Paso") %>%   # 🔴 THIS IS THE KEY FIX
  mutate(
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    crashes = n_distinct(Crash.ID),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    ma_12 = zoo::rollmean(crashes, k = 12, fill = NA, align = "center")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed
  geom_line(
    aes(y = crashes, color = "Total crashes"),
    linewidth = 0.8
  ) +

  # moving average
  geom_line(
    aes(y = ma_12, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total crashes" = "grey70",
      "Average" = "black"
    )
  ) +

  labs(
    x = "",
    y = "Total crashes",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)
```

``` {r}
#| title: Fatality Trends
monthly_counts <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    alcohol_fatal_crashes = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    alcohol_fatal_ma = zoo::rollmean(alcohol_fatal_crashes,
                                     k = 12,
                                     fill = NA,
                                     align = "right")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed total fatalities (grey)
  geom_line(
    aes(y = alcohol_fatal_crashes, color = "Total fatalities"),
    linewidth = 1
  ) +

  # alcohol fatalities 12-mo moving average (maroon)
  geom_line(
    aes(y = alcohol_fatal_ma, color = "Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total fatalities" = "grey70",
      "Average" = "maroon"
    )
  ) +

  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, 2)
  ) +

  labs(
    x = "",
    y = "Total fatalities",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)
```

``` {r}
#| title: Crashes by Month
monthly_year <- crash_level %>%
  filter(City == "El Paso") %>%
  mutate(
    date = as.Date(Crash.Date),
    year = year(date),
    month_num = month(date)
  ) %>%
  group_by(year, month_num) %>%
  summarise(
    monthly_crashes = n_distinct(Crash.ID),
    .groups = "drop"
  )

# ----------------------------
# Step 2: Average by month-of-year
# ----------------------------
monthly_avg <- monthly_year %>%
  group_by(month_num) %>%
  summarise(
    avg_crashes = mean(monthly_crashes),
    .groups = "drop"
  ) %>%
  mutate(
    month_name = factor(month.abb[month_num], levels = month.abb)
  )

# ----------------------------
# Step 3: overall mean line
# ----------------------------
overall_avg <- mean(monthly_avg$avg_crashes)

# ----------------------------
# Step 4: ggplot base
# ----------------------------
p <- ggplot(monthly_avg, aes(x = month_name, y = avg_crashes)) +

  geom_col(fill = "steelblue") +

  geom_hline(
   aes(yintercept = overall_avg, color = "Average"),
    linetype = "dashed",
    linewidth = 1
  ) +
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  labs(
    x = "",
    y = "Total crashes"
  ) +

  theme_minimal() +

  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ----------------------------
# Step 5: Plotly conversion
# ----------------------------
ggplotly(p)
```

``` {r}
#| title: Fatalities by Month
monthly_fatal_year <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    date = Crash.Date,
    year = year(date),
    month_num = month(date)
  ) %>%
  group_by(year, month_num) %>%
  summarise(
    monthly_fatalities = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------
# Step 2: Average fatalities by month-of-year
# ----------------------------
monthly_fatal_avg <- monthly_fatal_year %>%
  group_by(month_num) %>%
  summarise(
    avg_fatalities = mean(monthly_fatalities),
    .groups = "drop"
  ) %>%
  mutate(
    month_name = factor(month.abb[month_num], levels = month.abb)
  )

# ----------------------------
# Step 3: overall mean line
# ----------------------------
overall_avg <- mean(monthly_fatal_avg$avg_fatalities)

# ----------------------------
# Step 4: ggplot
# ----------------------------
p <- ggplot(monthly_fatal_avg, aes(x = month_name, y = avg_fatalities)) +

  geom_col(fill = "steelblue") +

  geom_hline(
    aes(yintercept = overall_avg, color = "Average"),
    linetype = "dashed",
    linewidth = 1
  ) +
  # fake mapping for legend consistency
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "",
    y = "Total fatalities"
  ) +

  theme_minimal(base_size = 12) +

  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# ----------------------------
# Step 5: Plotly
# ----------------------------
ggplotly(p)
```

## Row {height="400px"}

``` {r}
#| title: Crashes by Day of Week (YTD)
dow_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  count(Day.of.Week) %>%
  mutate(
    Day.of.Week = factor(
      tools::toTitleCase(tolower(Day.of.Week)),
      levels = c("Sunday", "Monday", "Tuesday", "Wednesday",
                 "Thursday", "Friday", "Saturday")
    )
  )

avg_val <- mean(dow_data$n, na.rm = TRUE)

p3 <- ggplot(dow_data, aes(x = Day.of.Week, y = n)) +

  # bars
  geom_col(fill = "steelblue") +

  # average line
  geom_hline(
    aes(yintercept = avg_val, color = "Average"),
    linewidth = 1.2,
    linetype = "dashed"
  ) +

  # fake mapping for legend consistency
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "",
    y = "Total crashes"
  )

ggplotly(p3)

```

``` {r}
#| title: Crashes by Hour (YTD)
hour_data <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  mutate(
    hour = as.numeric(substr(Hour.of.Day, 1, 2)),
    time_block = case_when(
      hour >= 0 & hour <= 5 ~ "00:00–05:59",
      hour >= 6 & hour <= 11 ~ "06:00–11:59",
      hour >= 12 & hour <= 17 ~ "12:00–17:59",
      hour >= 18 & hour <= 23 ~ "18:00–23:59",
      TRUE ~ "Unknown"
    )
  ) %>%
  count(time_block)

avg_block <- mean(hour_data$n, na.rm = TRUE)

p4 <- ggplot(hour_data, aes(x = time_block, y = n)) +

  geom_col(fill = "steelblue") +

  geom_hline(
    aes(yintercept = avg_block, color = "Average"),
    linetype = "dashed",
    linewidth = 1.2
  ) +
  # fake mapping for legend consistency
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "",
    y = "Total crashes"
  )

ggplotly(p4)
```

# Holidays

## Row {height="250px"}

```{r}
#| title: New Year's Day

base <- crash_level %>%
  filter(City == "El Paso", Crash.Year != 2020) %>%
  mutate(
    date = as.Date(Crash.Date),
    year = Crash.Year   # 🔥 FIX: required
  )

# ----------------------------
# 1. Build explicit event window per year
# ----------------------------
ny_window <- base %>%
  distinct(year) %>%
  mutate(
    holiday = as.Date(paste0(year, "-01-01"))
  ) %>%
  tidyr::expand_grid(event_time = -7:7) %>%
  mutate(date = holiday + event_time)

# ----------------------------
# 2. Join crashes to window
# ----------------------------
data <- ny_window %>%
  left_join(
    base %>% select(date, Crash.ID),
    by = "date"
  )

# ----------------------------
# 3. Aggregate
# ----------------------------
plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups = "drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes, na.rm = TRUE), .groups = "drop")

# ----------------------------
# 4. Plot
# ----------------------------
p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.3,
              color = "#2C3E50", size = 1) +
  geom_point(size = 3, color = "#34495E") +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#C0392B", size = 1) +
  scale_x_continuous(breaks = -7:7, limits = c(-7, 7)) +
  labs(x = "Days from New Year's Day", y = "Average Crashes") +
  theme_minimal(base_size = 10)

y_min <- min(plot_data$avg_crashes, na.rm = TRUE)
y_max <- max(plot_data$avg_crashes, na.rm = TRUE)

ggplotly(p) %>%
  layout(xaxis = list(range = c(-7, 7))) %>%
  add_ribbons(
    x = seq(-0.5, 0.5, length.out = 50),
    ymin = y_min, ymax = y_max,
    fillcolor = "rgba(255,0,0,0.08)",
    line = list(width = 0),
    inherit = FALSE,
    showlegend = FALSE
  )
```

``` {r}
#| title: Superbowl 

superbowl <- base %>%
  distinct(year) %>%
  mutate(
    feb1 = as.Date(paste0(year, "-02-01")),
    holiday = feb1 + (7 - wday(feb1)) + 7
  )

data <- base %>%
  left_join(superbowl, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Superbowl", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(0 - 0.5, 0 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(-7 - 0.5, -6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(7 - 0.5, 7 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

## Row {height="250px"}

```{r}
#| title: St. Patrick's Day

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-03-17")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Saint Patrick's Day", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-4 - 0.5, -4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(-1 - 0.5, -1 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(1 - 0.5, 1 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(4 - 0.5, 4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

``` {r}
#| title: Memorial Day

memorial <- base %>%
  distinct(year) %>%
  mutate(
    memorial_day = as.Date(paste0(year, "-05-31")) -
      (wday(as.Date(paste0(year, "-05-31")), week_start = 1) - 1)
  )

data <- base %>%
  left_join(memorial, by="year") %>%
  mutate(event_time = as.numeric(date - memorial_day)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Memorial Day", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-2 - 0.5, -2 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(5 - 0.5, 5 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) 
```

## Row {height="250px"}


``` {r}
#| title: Cinco de Mayo

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-05-05")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from May 5", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-6 - 0.5, -6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(1 - 0.5, 3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))

```

``` {r}
#| title: Independence Day

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-07-04")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from July 4", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-4 - 0.5, -4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(0 - 0.5, 1 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(3 - 0.5, 3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))
```

## Row {height="250px"}

``` {r}
#| title: Labor Day

labor <- base %>%
  distinct(year) %>%
  mutate(
    sept1 = as.Date(paste0(year, "-09-01")),
    holiday = sept1 + (8 - wday(sept1, week_start=1)) %% 7
  )

data <- base %>%
  left_join(labor, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Labor Day", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-3 - 0.5, 0 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(5 - 0.5, 6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))
```

``` {r}
#| title: Halloween

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-10-31")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from October 31", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-4 - 0.5, -4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(-2 - 0.5, -2 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(4 - 0.5, 4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(6 - 0.5, 6 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))
```

## Row {height="250px"}

``` {r}
#| title: Thanksgiving

thanksgiving <- base %>%
  distinct(year) %>%
  mutate(
    nov1 = as.Date(paste0(year, "-11-01")),
    holiday = nov1 + (4 - wday(nov1) + 7) %% 7 + 21
  )

data <- base %>%
  left_join(thanksgiving, by="year") %>%
  mutate(event_time = as.numeric(date - holiday)) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from Thanksgiving", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(-3 - 0.5, -3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0)) %>%
  add_ribbons(x = seq(2 - 0.5, 3 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))
```

``` {r}
#| title: Christmas

data <- base %>%
  mutate(
    holiday = as.Date(paste0(year, "-12-25")),
    event_time = as.numeric(date - holiday)
  ) %>%
  filter(event_time >= -7 & event_time <= 7)

plot_data <- data %>%
  group_by(year, event_time) %>%
  summarise(crashes = n_distinct(Crash.ID), .groups="drop") %>%
  group_by(event_time) %>%
  summarise(avg_crashes = mean(crashes), .groups="drop")

p <- ggplot(plot_data, aes(event_time, avg_crashes)) +
  geom_smooth(se=FALSE, method="loess", span=0.3,
              color="#2C3E50", size=1) +
  geom_point(size=3, color="#34495E") +
  geom_vline(xintercept=0, linetype="dashed", color="#C0392B", size=1) +
  scale_x_continuous(breaks=-7:7) +
  labs(x="Days from December 25", y="Average Crashes") +
  theme_minimal(base_size=10)

ggplotly(p) %>%
  add_ribbons(x = seq(4 - 0.5, 4 + 0.5, length.out = 50),
              ymin = min(plot_data$avg_crashes),
              ymax = max(plot_data$avg_crashes),
              fillcolor = "rgba(255,0,0,0.08)",
              line = list(width = 0))
```

# Drivers

``` {r}
# drivers object with BACs only
drivers <- person_level %>%
  mutate(
    driver_flag = case_when(
      Person.Type %in% c(
        "1 - DRIVER",
        "5 - DRIVER OF MOTORCYCLE TYPE VEHICLE"
      ) ~ 1,
      TRUE ~ 0
    ),

    bac_value = as.numeric(Person.Blood.Alcohol.Content.Test.Result),

    bac_result = case_when(
      !is.na(bac_value) & bac_value > 0 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    driver_flag == 1,
  )
```

## Row {height=30%}

``` {r}
#| content: valuebox
#| title: "Total Intoxicated Drivers (YTD)"

total_intox_drivers <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  summarise(n = n()) %>%
  pull(n)

list(
  icon = "emoji-dizzy",
  color = "primary",
  value = format(total_intox_drivers, big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Average BAC Result (YTD)"
bac_stats <- drivers %>%
  filter(bac_result == 1) %>%
  summarise(
    mean_bac = mean(bac_value, na.rm = TRUE),
    sd_bac = sd(bac_value, na.rm = TRUE)
  )
list(
  icon = "moisture",
  color = "danger",
  value = paste0(
    round(bac_stats$mean_bac, 2)
  )
)
```

```{r}
#| content: valuebox
#| title: "Average Age (YTD)"

age_stats <- drivers %>%
  filter(bac_value>=0.08 & bac_result==1) %>%
  mutate(
    age_num = as.numeric(Person.Age)
  ) %>%
  summarise(
    mean_age = mean(age_num, na.rm = TRUE),
    sd_age = sd(age_num, na.rm = TRUE)
  )

list(
  icon = "person-fill",
  color = "primary",
  value = paste0(
    round(age_stats$mean_age, 0)
  )
)
```

## Row { height=70%}

### Column {.tabset}

``` {r}
#| title: Intoxicated Driver's Zipcode (YTD)
zip_top10 <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1,
         !is.na(Driver.Zip.Code)) %>%
  group_by(Driver.Zip.Code) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  slice_head(n = 10)

zip_top10 <- zip_top10 %>%
  mutate(
    pct = n / sum(n),
    label_full = paste0(Driver.Zip.Code, "\n", n, " (", scales::percent(pct), ")")
  )


plot_ly(
  data = zip_top10,
  type = "treemap",
  labels = ~label_full,
  parents = rep("", nrow(zip_top10)),
  values = ~n,
  textinfo = "label",
  marker = list(
    colors = rep("#f4b942", nrow(zip_top10))
  ),
  hovertemplate = paste(
    "ZIP: %{label}<br>",
    "Drivers: %{value}<br>",
    "<extra></extra>"
  )
)

```

```{r}
#| title: Intoxicated Driver Demographics (YTD)

drivers_clean <- drivers %>%
  mutate(
    Person.Ethnicity = case_when(
      
      # Missing / unknown categories
      Person.Ethnicity %in% c("No Data", "99 - UNKNOWN", "UNKNOWN", "", NA) ~ NA_character_,
      
      # Special cases
      Person.Ethnicity == "95 - AUTONOMOUS" ~ "Autonomous",
      Person.Ethnicity == "98 - OTHER" ~ "Other",
      
      # Race/Ethnicity categories
      grepl("A - ASIAN", Person.Ethnicity) ~ "Asian",
      grepl("B - BLACK", Person.Ethnicity) ~ "Black",
      grepl("H - HISPANIC", Person.Ethnicity) ~ "Hispanic",
      grepl("I - AMER", Person.Ethnicity) ~ "American Indian/Alaska Native",
      grepl("W - WHITE", Person.Ethnicity) ~ "White",
      
      # Fallback
      TRUE ~ NA_character_
    ),
    
    Person.Gender = case_when(
      Person.Gender %in% c("99 - UNKNOWN", "UNKNOWN", "No Data", "") ~ NA_character_,
      Person.Gender == "1 - MALE" ~ "Male",
      Person.Gender == "2 - FEMALE" ~ "Female",
      TRUE ~ NA_character_
    )
  )

demo_table <- drivers_clean %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  filter(!is.na(Person.Ethnicity), !is.na(Person.Gender)) %>%
  
  mutate(
    Person.Gender = factor(Person.Gender, levels = c("Female", "Male")),
    
    Person.Age = as.numeric(Person.Age)  # convert from character
  ) %>%
  
  group_by(Person.Ethnicity, Person.Gender) %>%
  summarise(
    avg_bac = mean(bac_value, na.rm = TRUE),
    avg_age = mean(Person.Age, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    avg_bac = round(avg_bac, 2),
    avg_age = round(avg_age, 0)
  ) %>%
  
  arrange(desc(n)) %>%
  
  rename(
    "Race/Ethnicity" = Person.Ethnicity,
    "Gender" = Person.Gender,
    "Average Age" = avg_age,
    "Average BAC" = avg_bac,
    "Count" = n
  )

demo_table %>%
  gt() %>%
  
  
  # Zebra striping
  opt_row_striping() %>%
  
  # Make header row slightly bolder + cleaner
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) %>%
  
  # Light spacing + clean look
  tab_options(
    table.font.size = "small",
    column_labels.font.weight = "bold",
    column_labels.border.bottom.width = px(2),
    table.border.top.style = "none",
    table.border.bottom.style = "solid"
  )

```


### Column {.tabset}

```{r}
#| title: Vehicle Color
plot_data <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  mutate(
    color = case_when(
      grepl("BLK", Vehicle.Color) ~ "Black",
      grepl("WHI", Vehicle.Color) ~ "White",
      grepl("GRY", Vehicle.Color) ~ "Gray",
      grepl("RED", Vehicle.Color) ~ "Red",
      grepl("BLU", Vehicle.Color) ~ "Blue",
      grepl("GRN", Vehicle.Color) ~ "Green",
      grepl("SIL", Vehicle.Color) ~ "Silver",
      grepl("TAN", Vehicle.Color) ~ "Tan",
      grepl("MAR", Vehicle.Color) ~ "Maroon",
      grepl("YEL", Vehicle.Color) ~ "Yellow",
      grepl("GLD", Vehicle.Color) ~ "Gold",
      grepl("PLE", Vehicle.Color) ~ "Purple",
      grepl("BGE", Vehicle.Color) ~ "Beige",
      TRUE ~ "Other"
    )
  ) %>%
  count(color) %>%
  mutate(
    pct = n / sum(n),
    label = paste0(n, " (", scales::percent(pct), ")")
  )
plot_ly(
  data = plot_data,
  x = ~n,
  y = ~fct_reorder(color, n),
  type = "bar",
  orientation = "h",
  marker = list(color = "steelblue"),
  text = ~label,
  textposition = "outside"
) %>%
  layout(
    xaxis = list(title = "Total intoxicated drivers"),
    yaxis = list(title = ""),
    margin = list(r = 100)
  )

```

``` {r}
#| title: Vehicle Body Style
plot_data <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  mutate(
    body = case_when(
      grepl("MC", Vehicle.Body.Style) ~ "Motorcycle",
      grepl("P2", Vehicle.Body.Style) ~ "Passenger Car (2-door)",
      grepl("P4", Vehicle.Body.Style) ~ "Passenger Car (4-door)",
      grepl("PK", Vehicle.Body.Style) ~ "Pickup",
      grepl("SV", Vehicle.Body.Style) ~ "SUV",
      grepl("TR", Vehicle.Body.Style) ~ "Truck",
      TRUE ~ "Other"
    )
  ) %>%
  filter(!is.na(body)) %>%
  count(body)

plot_data <- plot_data %>%
  mutate(
    pct = n / sum(n),
    label = paste0(n, " (", scales::percent(pct), ")")
  )

plot_ly(
  data = plot_data,
  x = ~n,
  y = ~fct_reorder(body, n),
  type = "bar",
  orientation = "h",
  text = ~label,
  textposition = "outside",
  marker = list(color = "steelblue")
) %>%
  layout(
    xaxis = list(title = "Total intoxicated drivers"),
    yaxis = list(title = "")
  )

```

``` {r}
#| title: Vehicle Make

plot_data <- drivers %>%
  filter(bac_value >= 0.08 & bac_result == 1) %>%
  
  mutate(
    make = Vehicle.Make,
    make = str_to_sentence(str_to_lower(make))  # robust sentence case
  ) %>%
  
  filter(!is.na(make), make != "") %>%
  
  count(make) %>%
  
  mutate(
    pct = n / sum(n),
    label = paste0(n, " (", scales::percent(pct), ")")
  )

plot_ly(
  data = plot_data,
  x = ~n,
  y = ~fct_reorder(make, n),
  type = "bar",
  orientation = "h",
  marker = list(color = "steelblue"),
  text = ~label,
  textposition = "outside"
) %>%
  layout(
    xaxis = list(title = "Total intoxicated drivers"),
    yaxis = list(title = ""),
    margin = list(r = 120)
  )
```

# Map

## Row {height=100%}

### Column {.tabset width=70%}

```{r}
#| title: Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "black"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

```{r}
#| title: Fatal Crashes (Since 2016)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Fatal.Crash.Flag == 1) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "maroon"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

### Column {.tabset width=30%}

``` {r}
#| title: Crash Locations (YTD)
top_10_streets <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  slice_head(n = 15) %>%
  rename(
    "Street Name" = Street.Name,
    "Crash Count" = n,
    "Percent (%)" = percent
  )

top_10_streets %>%
  gt()
```

``` {r}
#| title: Fatal Crash Locations (YTD)
top_fatal_streets <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    Fatal.Crash.Flag == 1
  ) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    "Street Name" = Street.Name,
    "Fatal Crashes" = n,
    "Percent (%)" = percent
  )

top_fatal_streets %>%
  gt()
```

# About

## Row {height=50%}

::: {.card title="About This Dashboard"}

Last Updated: March 2026

Crash data is downloaded from the Texas Department of Transportation (TXDOT) website using the 
[Crash Query Tool](https://cris.dot.state.tx.us/public/Query/app/home). All crash data in this dashboard represent those classified as alcohol-related and that occurred in the city of El Paso.

**Definition of Alcohol-Related Crash**

A crash is defined as alcohol-related if involved drivers or individuals were reported as "Under the Influence of Alcohol", "Had Been Drinking", or were reported to have a blood alcohol content (BAC) greater than 0.

**Definition of Injury/Fatality**

Injuries include people who were classified as having a suspected serious injury, suspected minor injury, or possible injury. Fatalities are defined as any motor vehicle traffic injury crash that results in the death of one or more persons within 30 days of the incident.

**Definition of Intoxicated Driver**

An intoxicated driver is a person who's BAC test returned a value of 0.08g/dL or greater.

:::

::: {.card title="Technical Details"}

**Calculation of Moving Average**

Averages for crashes and fatal crashes between 2016 and *Present* were calculated using a 12-month moving average. This technique results in a smooth line, allowing the viewer to see if a trend is increasing or decreasing.

**Holiday Analyses**

Holiday crash statistics were calculated by aligning crash data to each holiday. For fixed-date holidays (e.g., New Year’s Day, Independence Day, Halloween, Christmas, and St. Patrick’s Day), each crash was assigned an event time based on the number of days before or after the calendar date of the holiday within each year. For shifting holidays (e.g., Super Bowl and Memorial Day), the holiday date was first computed for each year and then used as the reference point for event-time calculation.

In all cases, crashes were restricted to a ±7-day window around the holiday and then aggregated across years (2016–2025), excluding 2020 to avoid COVID-19-related disruptions. Average daily crash counts were then computed for each relative day to identify typical patterns before and after each holiday.

**Missing Data**

Missing data in the driver characteristics (such as gender and race/ethnicity) was minimal. Observations with missing or unknown values in these fields were excluded from analysis to ensure consistency and interpretability of demographic comparisons.
:::

# Quiz

::: {.column width="100%"}

::: {.card title="Drunk Driving Quiz" style="min-height:780px; margin:0 auto;"}

<div id="quizApp">

  <h4 id="introStats">Loading stats...</h4>
  <h4 id="avgStats"></h4>

  <button onclick="startQuiz()" id="startBtn">Start Quiz</button>

  <p id="tagline" style="font-style: italic;"></p>

  <hr>

  <div id="quizBox" style="display:none"></div>

</div>


<script>
const SCRIPT_URL = "https://script.google.com/macros/s/AKfycbw4TekPtUfV08eUUU9TbRu0QdeUl9_mondBZS0c179w-eQQb4YliAQwwUHyODuErsYA/exec";

let qIndex = 0;
let score = 0;

/* ---------------- QUIZ ---------------- */
const quiz = [

  {
    q: "When are alcohol-related crashes most likely to happen in El Paso?",
    a: ["Weekday mornings", "Weekends (Saturday and Sunday)", "Midday during the week"],
    correct: "Weekends (Saturday and Sunday)",
    explanation: "Crash data shows that alcohol-related crashes in El Paso are most common on weekends."
  },

  {
    q: "What time of day do alcohol-related crashes most often occur?",
    a: ["6:00 AM to 12:00 PM", "12:00 PM to 6:00 PM", "6:00 PM to 6:00 AM"],
    correct: "6:00 PM to 6:00 AM",
    explanation: "Most alcohol-related crashes happen at night and early morning hours, when visibility is lower and alcohol consumption is more common."
  },

  {
    q: "Since 2016, what has happened to alcohol-related crashes in El Paso?",
    a: ["They have increased steadily", "They have decreased overall", "They have stayed exactly the same"],
    correct: "They have decreased overall",
    explanation: "Overall trends show a decline in alcohol-related crashes since 2016. There was a sharp drop during COVID-19, followed by a return toward normal levels, and then a continued downward trend."
  },

  {
    q: "What is the legal blood alcohol concentration (BAC) limit for driving in Texas?",
    a: ["0.05%", "0.08%", "0.10%"],
    correct: "0.08%",
    explanation: "In Texas, it is illegal to drive with a BAC of 0.08% or higher. This level indicates significant impairment in reaction time and judgment."
  },

  {
    q: "On average, what BAC level do drivers involved in alcohol-related crashes in El Paso have?",
    a: ["About 0.08% (legal limit)", "About 0.10%", "About 0.16% (more than twice the legal limit)"],
    correct: "About 0.16% (more than twice the legal limit)",
    explanation: "The data shows that many drivers involved in alcohol-related crashes have very high BAC levels—around 0.16%—which is more than twice the legal limit and associated with severe impairment."
  }

];

/* ---------------- STATS ---------------- */
function loadStats() {

  fetch(SCRIPT_URL)
    .then(r => r.text())
    .then(t => {

      const data = JSON.parse(t);
      data.shift();

      const scores = data
        .map(r => Number(r[3]))   // COLUMN D
        .filter(v => !isNaN(v));

      const count = scores.length;

      const avg = count
        ? (scores.reduce((a,b)=>a+b,0)/count).toFixed(1)
        : 0;

      document.getElementById("introStats").innerHTML =
        `${count} people have taken this quiz.`;

      document.getElementById("avgStats").innerHTML =
        `Average score: ${avg}%`;

      document.getElementById("tagline").innerHTML =
        "";
    });
}

/* ---------------- START ---------------- */
function startQuiz() {
  document.getElementById("startBtn").style.display = "none";
  document.getElementById("quizBox").style.display = "block";
  loadQuestion();
}

/* ---------------- LOAD QUESTION ---------------- */
function loadQuestion() {

  const q = quiz[qIndex];
  const box = document.getElementById("quizBox");

  box.innerHTML = `
    <h4>Question ${qIndex + 1} of ${quiz.length}</h4>
    <p>${q.q}</p>
    <div id="answers"></div>
    <p id="feedback"></p>
    <button id="nextBtn" style="display:none" onclick="nextQuestion()">Next</button>
  `;

  const answersDiv = document.getElementById("answers");

  q.a.forEach(choice => {

    const btn = document.createElement("button");
    btn.innerText = choice;

    btn.onclick = () => {

      const buttons = answersDiv.querySelectorAll("button");
      buttons.forEach(b => b.disabled = true);

      if (choice === q.correct) {
        score++;
        btn.style.backgroundColor = "#c8f7c5";

        document.getElementById("feedback").innerHTML =
          `
          <span style="color:green;"><b>Correct</b></span><br><br>
          ${q.explanation}
          `;
      } else {
        btn.style.backgroundColor = "#f7c5c5";

        buttons.forEach(b => {
          if (b.innerText === q.correct) {
            b.style.backgroundColor = "#c8f7c5";
          }
        });

          document.getElementById("feedback").innerHTML =
          `
          <b>Correct answer:</b> ${q.correct}<br><br>
          ${q.explanation}
          `;
      }

      document.getElementById("nextBtn").style.display = "inline-block";
    };

    answersDiv.appendChild(btn);
  });
}

/* ---------------- NEXT ---------------- */
function nextQuestion() {

  qIndex++;

  if (qIndex < quiz.length) {
    loadQuestion();
  } else {
    finishQuiz();
  }
}

/* ---------------- FINISH ---------------- */
function finishQuiz() {

  const percent = Math.round((score / quiz.length) * 100);

  fetch(SCRIPT_URL, {
    method: "POST",
    body: JSON.stringify({
      timestamp: new Date().toISOString(),
      score: percent
    })
  });

  document.getElementById("quizBox").innerHTML =
    `<h3>Finished</h3>
     <p>Your score: <b>${percent}%</b></p>`;
}

/* ---------------- INIT ---------------- */
document.addEventListener("DOMContentLoaded", loadStats);
</script>

