---
title: "The El Paso Drunk Driving Dashboard"
format:
  dashboard:
    theme:
      light: flatly
      dark: darkly
    orientation: rows
execute:
  echo: false
  warning: false
  message: false
---

Welcome to the El Paso, Texas Drunk Driving Dashboard. This dashboard displays vehicle crash data collected by the Texas Department of Public Safety. All calculations are done by the author. Please see the "About Dashboard" page for detailed information about calculations.

```{r}

rm(list = ls())

suppressMessages(library(DT))
suppressMessages(library(quarto))
suppressMessages(library(rsconnect))
suppressMessages(library(shiny))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
suppressMessages(library(stringr))
suppressMessages(library(readr))
suppressMessages(library(tidyverse))
suppressMessages(library(leaflet))
suppressMessages(library(leaflet.extras))


files <- c(
  austin = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/austin_workingdata.csv",
  brownsville = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/brownsville_workingdata.csv",
  dallas = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/dallas_workingdata.csv",
  elpaso = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/elpaso_workingdata.csv",
  fortworth = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/fortworth_workingdata.csv",
  houston = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/houston_workingdata.csv",
  laredo = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/laredo_workingdata.csv",
  mcallen = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/mcallen_workingdata.csv",
  sanantonio = "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/Drunk Driving dashboard/data/sanantonio_workingdata.csv"
)

all_cities <- bind_rows(lapply(names(files), function(city) {
  df <- read.csv(files[[city]], header = TRUE)
  df$city <- city
  df
}))

all_cities$City <- recode(all_cities$City,
  "AUSTIN" = "Austin",
  "BROWNSVILLE" = "Brownsville",
  "DALLAS" = "Dallas",
  "EL PASO" = "El Paso",
  "FORT WORTH" = "Fort Worth",
  "HOUSTON" = "Houston",
  "MCALLEN" = "McAllen",
  "SAN ANTONIO" = "San Antonio",
  "LAREDO" = "Laredo"
)

# drop current month, since its incomplete
all_cities <- all_cities %>%
  mutate(
    Crash.Date = mdy(Crash.Date)
  ) %>%
  filter(
    !(year(Crash.Date) == year(Sys.Date()) &
      month(Crash.Date) == month(Sys.Date()))
  )

person_level <- all_cities

vehicle_level <- all_cities %>%
  distinct(Crash.ID, Unit.Number, .keep_all = TRUE)

crash_level <- all_cities %>%
  distinct(Crash.ID, .keep_all = TRUE) 
```

# Trends


## Row {height=30%}

```{r}
#| content: valuebox
#| title: "Total Alcohol-Related Crashes (YTD)"
ytd_alcohol_crashes <- crash_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026
  ) %>%
  summarise(
    ytd_alcohol_crashes = n_distinct(Crash.ID)
  ) %>%
  pull(ytd_alcohol_crashes)

list(
  icon = "car-front-fill",
  color = "primary",
  value = ytd_alcohol_crashes )
```

```{r}
#| content: valuebox
#| title: "Total People Injured in Alcohol-Related Crashes (YTD)"

person_level <- person_level %>%
  mutate(
    injured = if_else(
      str_detect(Person.Injury.Severity, "A - SUSPECTED SERIOUS INJURY|B - SUSPECTED MINOR INJURY|C - POSSIBLE INJURY"),
      1, 0,
      missing = 0
    )
  )

ytd_people_alcohol <- person_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    injured == 1) %>%
    nrow()


list(
  icon = "bandaid",
  color = "primary",
  value = ytd_people_alcohol)

```

```{r}
#| content: valuebox
#| title: "Total Alcohol-Related Traffic Fatalities (YTD)"
person_level <- person_level %>%
  mutate(
    fatal = if_else(
      str_detect(Person.Injury.Severity, "K - FATAL INJURY"),
      1, 0,
      missing = 0
    )
  )
ytd_alcohol_fatalities <- person_level %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    fatal == 1) %>%
    nrow()

list(
  icon = "hospital",
  color = "danger",
  value = ytd_alcohol_fatalities
)
```


## Row {height=70%}

### Column {.tabset}

``` {r}
#| title: Crashes
monthly_counts <- crash_level %>%
  mutate(
    month = floor_date(Crash.Date, "month"),
  ) %>%
  group_by(month) %>%
  summarise(
    alcohol_crashes = n_distinct(Crash.ID),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    alcohol_ma = zoo::rollmean(alcohol_crashes, k = 12, fill = NA, align = "center")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed
  geom_line(
    aes(y = alcohol_crashes, color = "Total Alcohol-Related Crashes"),
    linewidth = 0.8
  ) +

  # moving average
  geom_line(
    aes(y = alcohol_ma, color = "Moving Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total Alcohol-Related Crashes" = "grey70",
      "Moving Average" = "black"
    )
  ) +

  labs(
    x = "",
    y = "Total crashes",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)
```

``` {r}
#| title: Fatalities
monthly_counts <- person_level %>%
  filter(City == "El Paso") %>%
  mutate(
    month = floor_date(Crash.Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    alcohol_fatal_crashes = sum(fatal == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    alcohol_fatal_ma = zoo::rollmean(alcohol_fatal_crashes,
                                     k = 12,
                                     fill = NA,
                                     align = "right")
  )

p <- ggplot(monthly_counts, aes(x = month)) +

  # observed total fatalities (grey)
  geom_line(
    aes(y = alcohol_fatal_crashes, color = "Total Alcohol-Related Traffic Fatalities"),
    linewidth = 1
  ) +

  # alcohol fatalities 12-mo moving average (maroon)
  geom_line(
    aes(y = alcohol_fatal_ma, color = "Moving Average"),
    linewidth = 1.2
  ) +

  scale_color_manual(
    values = c(
      "Total Alcohol-Related Traffic Fatalities" = "grey70",
      "Moving Average" = "maroon"
    )
  ) +

  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, 2)
  ) +

  labs(
    x = "",
    y = "Total fatalities",
    color = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  )

ggplotly(p)
```

### Column {.tabset}

``` {r}
#| title: Day of Crash
dow_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  count(Day.of.Week) %>%
  mutate(
    Day.of.Week = factor(
      tools::toTitleCase(tolower(Day.of.Week)),
      levels = c("Sunday", "Monday", "Tuesday", "Wednesday",
                 "Thursday", "Friday", "Saturday")
    )
  )

avg_val <- mean(dow_data$n, na.rm = TRUE)

p3 <- ggplot(dow_data, aes(x = Day.of.Week, y = n)) +

  # bars
  geom_col(fill = "steelblue") +

  # average line
  geom_hline(
    aes(yintercept = avg_val, color = "Average"),
    linewidth = 1.2,
    linetype = "dashed"
  ) +

  # fake mapping for legend consistency
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "",
    y = "Alcohol-Related Crashes"
  )

ggplotly(p3)

```

``` {r}
#| title: Hour of Crash
hour_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(hour = as.numeric(substr(Hour.of.Day, 1, 2))) %>%
  count(hour)

avg_hour <- mean(hour_data$n, na.rm = TRUE)

p4 <- ggplot(hour_data, aes(x = hour, y = n)) +

  # bars
  geom_col(fill = "steelblue") +

  # average line
  geom_hline(
    aes(yintercept = avg_hour, color = "Average"),
    linewidth = 1.2,
    linetype = "dashed"
  ) +

  # legend mapping
  scale_color_manual(
    values = c("Average" = "orange"),
    name = ""
  ) +

  scale_x_continuous(breaks = 0:23) +

  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Hour (24-Hour Format)",
    y = "Alcohol-Related Crashes"
  )

ggplotly(p4)
```


# Drivers

``` {r}
# drivers object with BACs
drivers <- person_level %>%
  mutate(
    driver_flag = case_when(
      Person.Type %in% c(
        "1 - DRIVER",
        "5 - DRIVER OF MOTORCYCLE TYPE VEHICLE"
      ) ~ 1,
      TRUE ~ 0
    ),

    bac_value = as.numeric(Person.Blood.Alcohol.Content.Test.Result),

    bac_result = case_when(
      !is.na(bac_value) & bac_value > 0 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  filter(
    City == "El Paso",
    Crash.Year == 2026,
    driver_flag == 1,
    bac_result == 1
  )
```

## Row {height=30%}

```{r}
#| content: valuebox
#| title: "Average BAC (YTD)"
bac_stats <- drivers %>%
  summarise(
    mean_bac = mean(bac_value, na.rm = TRUE),
    sd_bac = sd(bac_value, na.rm = TRUE)
  )
list(
  icon = "moisture",
  color = "danger",
  value = paste0(
    round(bac_stats$mean_bac, 2),
    " ± ",
    round(bac_stats$sd_bac, 2)
  )
)
```

```{r}
#| content: valuebox
#| title: "Male Drivers (YTD)"
male_pct <- drivers %>%
  mutate(
    male = if_else(Person.Gender == "1 - MALE", 1, 0)
  ) %>%
  summarise(
    pct_male = mean(male, na.rm = TRUE) * 100
  ) %>%
  pull(pct_male)

list(
  icon = "gender-male",
  color = "primary",
  value = paste0(round(male_pct, 1), "%")
)
```

```{r}
#| content: valuebox
#| title: "Average Driver Age (YTD)"

age_stats <- drivers %>%
  mutate(
    age_num = as.numeric(Person.Age)
  ) %>%
  summarise(
    mean_age = mean(age_num, na.rm = TRUE),
    sd_age = sd(age_num, na.rm = TRUE)
  )

list(
  icon = "person-fill",
  color = "primary",
  value = paste0(
    round(age_stats$mean_age, 1),
    " ± ",
    round(age_stats$sd_age, 1)
  )
)
```

## Row { height=70%}

### Column

```{r}
#| title: Drivers Over the Legal Limit (YTD)


drivers %>%
  mutate(
    bac_group = if_else(
      bac_value >= 0.08,
      "0.08+ (Over legal limit)",
      "<0.08"
    )
  ) %>%
  count(bac_group) %>%
  plot_ly(
    labels = ~bac_group,
    values = ~n,
    type = "pie",
    textinfo = "label+percent",
    marker = list(
      colors = c("maroon", "lightgray")
    )
  )

```


### Column {.tabset}

```{r}
#| title: Vehicle Color
drivers %>%
  mutate(
    color = case_when(
      grepl("BLK", Vehicle.Color) ~ "Black",
      grepl("WHI", Vehicle.Color) ~ "White",
      grepl("GRY", Vehicle.Color) ~ "Gray",
      grepl("RED", Vehicle.Color) ~ "Red",
      grepl("BLU", Vehicle.Color) ~ "Blue",
      grepl("GRN", Vehicle.Color) ~ "Green",
      grepl("SIL", Vehicle.Color) ~ "Silver",
      grepl("TAN", Vehicle.Color) ~ "Tan",
      grepl("MAR", Vehicle.Color) ~ "Maroon",
      grepl("YEL", Vehicle.Color) ~ "Yellow",
      grepl("GLD", Vehicle.Color) ~ "Gold",
      grepl("PLE", Vehicle.Color) ~ "Purple",
      grepl("BGE", Vehicle.Color) ~ "Beige",
      TRUE ~ "Other"
    )
  ) %>%
  count(color) %>%
  plot_ly(
    x = ~n,
    y = ~fct_reorder(color, n),
    type = "bar",
    orientation = "h",
    marker = list(color = "steelblue")
  ) %>%
  layout(xaxis = list(title = ""), yaxis = list(title = ""))

```

``` {r}
#| title: Vehicle Body Style
drivers %>%
  mutate(
    body = case_when(
      grepl("MC", Vehicle.Body.Style) ~ "Motorcycle",
      grepl("P2", Vehicle.Body.Style) ~ "Passenger Car (2-door)",
      grepl("P4", Vehicle.Body.Style) ~ "Passenger Car (4-door)",
      grepl("PK", Vehicle.Body.Style) ~ "Pickup",
      grepl("SV", Vehicle.Body.Style) ~ "SUV",
      grepl("TR", Vehicle.Body.Style) ~ "Truck",
      TRUE ~ "Other"
    )
  ) %>%
  count(body) %>%
  plot_ly(
    x = ~n,
    y = ~fct_reorder(body, n),
    type = "bar",
    orientation = "h",
    marker = list(color = "steelblue")
  ) %>%
  layout(xaxis = list(title = ""), yaxis = list(title = ""))
```

``` {r}
#| title: Vehicle Make
drivers %>%
  mutate(
    make = case_when(
      Vehicle.Make %in% c("FORD","CHEVROLET","DODGE","GMC") ~ Vehicle.Make,
      Vehicle.Make %in% c("TOYOTA","HONDA","NISSAN","HYUNDAI","KIA","MAZDA","SUBARU","VOLKSWAGEN","LEXUS") ~ Vehicle.Make,
      TRUE ~ NA_character_
    ),
    make = str_to_title(make)
  ) %>%
  filter(!is.na(make)) %>%
  count(make) %>%
  plot_ly(
    x = ~n,
    y = ~fct_reorder(make, n),
    type = "bar",
    orientation = "h",
    marker = list(color = "steelblue")
  ) %>%
  layout(
    xaxis = list(title = ""),
    yaxis = list(title = "")
  )
```

# Map

## Row {height=100%}

### Column {.tabset width=75%}

```{r}
#| title: Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "black"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

```{r}
#| title: Fatal Crashes (YTD)
map_data <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026,
         Fatal.Crash.Flag == 1) %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.4,
    color = "maroon"
  ) %>%
  
  setView(
    lng = mean(map_data$Longitude, na.rm = TRUE),
    lat = mean(map_data$Latitude, na.rm = TRUE),
    zoom = 11
  )
```

### Column {width=25%}

``` {r}
#| title: Streets with most alcohol-related crashes (YTD)
top_10_streets <- crash_level %>%
  filter(City == "El Paso",
         Crash.Year == 2026) %>%
  count(Street.Name, sort = TRUE) %>%
  mutate(
    percent = paste0(round((n / sum(n) * 100), 2), "%")
  ) %>%
  slice_head(n = 15)

datatable(
  top_10_streets,
  colnames = c("Name", "Count", "Percent"),
  options = list(
    pageLength = 15,
    searching = TRUE,
    dom = "tip"   # keeps search box, table info, pagination
  )
)
```





