---
title: "Latest Local News on Drunk Driving"
format: html
execute:
  echo: false
  warning: false
  message: false
---
<style>
.quarto-title-block {
  display: none;
} </style>

## Latest Local News on Drunk Driving

```{r results='asis'}
el_paso_time <- format(Sys.time(), tz = "America/Denver", "%B %d, %Y • %I:%M %p")

cat(paste0("Last updated (MST): ", el_paso_time))
```

<div style="display:flex; gap:12px; margin-bottom:8px; width:100%; align-items:center;">

<input id="searchBox" type="text"
placeholder="Search news stories"
style="flex:1; padding:12px; border-radius:8px; border:1px solid #ddd; font-size:14px;">

<select id="categoryFilter"
style="width:220px; padding:12px; border-radius:8px; border:1px solid #ddd;">
  <option value="all">All Tags</option>
  <option value="DWI">DWI</option>
  <option value="Crash">Crash</option>
  <option value="Policy">Policy</option>
  <option value="Arrest">Arrest</option>
  <option value="Court">Court</option>
  <option value="Legislation">Legislation</option>
  <option value="Enforcement">Enforcement</option>
  <option value="Alcohol">Alcohol</option>
  <option value="Traffic">Traffic</option>
  <option value="Injury">Injury</option>
  <option value="Fatality">Fatality</option>
  <option value="Police">Police</option>
  <option value="Texas">Texas</option>
  <option value="ElPaso">El Paso</option>
</select>

</div>

<div style="margin-bottom:10px; color:#888; font-size:0.9em;">
  <span id="resultCount"></span>
</div>

<div id="news-grid" style="
  display:grid;
  grid-template-columns: repeat(3, 1fr);
  gap:16px;
">
</div>

<div style="margin-top:20px; text-align:center;">
  <button onclick="prevPage()">Previous</button>
  <span id="pageInfo" style="margin:0 10px;"></span>
  <button onclick="nextPage()">Next</button>
</div>

```{r, results='asis'}
library(tidyRSS)
library(dplyr)
library(stringr)

rss_url <- "https://news.google.com/rss/search?q=(\"DWI\"%20OR%20\"drunk%20driving\"%20OR%20\"impaired%20driving\")%20\"El%20Paso\"&hl=en-US&gl=US&ceid=US:en"

rss_items <- tidyfeed(rss_url) %>%
  select(item_title, item_link, item_pub_date) %>%
  slice_head(n = 100)

clean_title <- function(full_title) {
  parts <- str_split(full_title, " - ", simplify = TRUE)

  if (ncol(parts) > 1) {
    title <- parts[,1]
    source <- parts[,ncol(parts)]
  } else {
    title <- full_title
    source <- "Unknown"
  }

  list(title = title, source = toupper(source))
}

format_date <- function(x) {
  format(as.Date(x), "%B %d, %Y")
}

get_tags <- function(text) {
  tags <- c()

  if (str_detect(text, regex("dwi|dui|drunk|impaired", TRUE))) tags <- c(tags, "DWI")
  if (str_detect(text, regex("crash|accident|collision", TRUE))) tags <- c(tags, "Crash")
  if (str_detect(text, regex("policy|law|bill", TRUE))) tags <- c(tags, "Policy")
  if (str_detect(text, regex("arrest|charged", TRUE))) tags <- c(tags, "Arrest")
  if (str_detect(text, regex("court|trial", TRUE))) tags <- c(tags, "Court")
  if (str_detect(text, regex("legislation", TRUE))) tags <- c(tags, "Legislation")
  if (str_detect(text, regex("enforcement", TRUE))) tags <- c(tags, "Enforcement")
  if (str_detect(text, regex("alcohol", TRUE))) tags <- c(tags, "Alcohol")
  if (str_detect(text, regex("traffic", TRUE))) tags <- c(tags, "Traffic")
  if (str_detect(text, regex("injury", TRUE))) tags <- c(tags, "Injury")
  if (str_detect(text, regex("fatal", TRUE))) tags <- c(tags, "Fatality")
  if (str_detect(text, regex("police", TRUE))) tags <- c(tags, "Police")
  if (str_detect(text, regex("texas", TRUE))) tags <- c(tags, "Texas")
  if (str_detect(text, regex("el paso", TRUE))) tags <- c(tags, "ElPaso")

  if (length(tags) == 0) tags <- "General"
  paste(tags, collapse = ",")
}

for (i in 1:nrow(rss_items)) {

  parsed <- clean_title(rss_items$item_title[i])
  tags <- get_tags(rss_items$item_title[i])

  cat(
    paste0(
      "<div class='news-card' data-tags='", tags, "' style='
        padding:16px;
        border-radius:12px;
        background:#fff3d6;
        border:1px solid #e5e7eb;
        position:relative;
        transition: all 0.25s ease;
      '
      onmouseover=\"this.style.transform='translateY(-3px)'; this.style.boxShadow='0 8px 20px rgba(0,0,0,0.12)'\"
      onmouseout=\"this.style.transform='translateY(0)'; this.style.boxShadow='none'\">",

      "<div style='position:absolute; top:10px; right:12px; font-size:0.75em; color:#888;'>",
      format_date(rss_items$item_pub_date[i]),
      "</div>",

      "<div style='font-size:0.75em; font-weight:700; color:#7a2e1d; margin-bottom:6px;'>",
      parsed$source,
      "</div>",

      "<a href='", rss_items$item_link[i], "' target='_blank' style='
        font-size:1.05em;
        font-weight:600;
        color:#5c1f14;
        text-decoration:none;
      '>",
      parsed$title,
      "</a>",

      "<div style='margin-top:10px;'>",
      paste(
        sapply(strsplit(tags, ",")[[1]], function(tag) {
          paste0(
            "<span style='
              font-size:0.7em;
              background:#7a2e1d;
              color:#fff3d6;
              padding:4px 8px;
              border-radius:6px;
              margin-right:6px;
            '>", tag, "</span>"
          )
        }),
        collapse = ""
      ),
      "</div></div>"
    )
  )
}
```

<script>
let currentPage = 1;
const pageSize = 12;

function getCards() {
  return Array.from(document.querySelectorAll(".news-card"));
}

function updateResultCount(visibleCards) {
  document.getElementById("resultCount").innerText =
    visibleCards + " results found";
}

function renderPage() {
  const cards = getCards();
  const start = (currentPage - 1) * pageSize;
  const end = start + pageSize;

  let visible = 0;

  cards.forEach((card, i) => {
    const show = (i >= start && i < end);
    if (show) visible++;
    card.style.display = show ? "block" : "none";
  });

  updateResultCount(cards.length);

  document.getElementById("pageInfo").innerText =
    "Page " + currentPage + " of " + Math.ceil(cards.length / pageSize);
}

function nextPage() {
  const cards = getCards();
  if (currentPage < Math.ceil(cards.length / pageSize)) {
    currentPage++;
    renderPage();
  }
}

function prevPage() {
  if (currentPage > 1) {
    currentPage--;
    renderPage();
  }
}

function filterCards() {
  const search = document.getElementById("searchBox").value.toLowerCase();
  const category = document.getElementById("categoryFilter").value;

  let visible = 0;

  getCards().forEach(card => {
    const text = card.innerText.toLowerCase();
    const tags = card.dataset.tags;

    const matchesSearch = text.includes(search);
    const matchesCategory = category === "all" || tags.includes(category);

    const show = matchesSearch && matchesCategory;

    card.style.display = show ? "block" : "none";
    if (show) visible++;
  });

  updateResultCount(visible);
}

document.getElementById("searchBox").addEventListener("input", filterCards);
document.getElementById("categoryFilter").addEventListener("change", filterCards);

renderPage();
</script>
