---
title: "Service Provider Directory"
subtitle: "El Paso, Texas"
format:
  html:
    embed-resources: true
    page-layout: full
    toc: true
    toc-depth: 2
    toc-title: "On this page"
execute:
  echo: false
  warning: false
  message: false
---

<style>

.quarto-title-block {
  display: none;
}

</style>



<style>
/* ============================================================
   SECTION CONTAINERS
   ============================================================ */

.section-container {
  background: #faf8f3;
  border-radius: 16px;
  padding: 2rem;
  margin: 2rem 0;
  border: 1px solid #e4ddd2;
  box-shadow: 0 4px 14px rgba(60, 45, 30, 0.07);
}

/* Directory */
.directory-container {
  background: #faf8f3;
}

/* Map */
.map-container {
  background: #eee8de;
  padding: 1.25rem;
}

/* Section headings */
.section-container h2 {
  margin-top: 0;
  margin-bottom: 1.25rem;
  font-weight: 700;
  color: #292929;
}

/* Make leaflet map fit nicely inside its container */
.map-container .leaflet-container {
  width: 100% !important;
  border-radius: 12px;
  overflow: hidden;
  min-height: 600px;
}

/* ============================================================
   PROVIDER CARDS
   ============================================================ */

.provider-card {
  padding: 20px;
  border-radius: 12px;
  background: #ffffff;
  border: 1px solid #e5dfd5;
  margin-bottom: 14px;
  transition: all 0.25s ease;
}

.provider-card:hover {
  transform: translateY(-3px);
  box-shadow: 0 8px 20px rgba(60,45,30,0.12);
}

.provider-name {
  font-size: 1.1rem;
  margin-bottom: 6px;
}

.provider-blurb {
  font-size: 13px;
  color: #555;
  margin-top: 8px;
  margin-bottom: 12px;
}

.provider-info {
  font-size: 13px;
  line-height: 1.7;
  margin-top: 12px;
}

.provider-info a {
  color: #7a2e1d;
  text-decoration: none;
}

.provider-info a:hover {
  text-decoration: underline;
}

.family-message {
  margin-top: 16px;
  padding: 14px 16px;
  background: #f8f9fa;
  border-left: 4px solid #7a2e1d;
  border-radius: 6px;
  font-size: 13px;
  line-height: 1.6;
}

.family-message-label {
  font-weight: 600;
  margin-bottom: 4px;
  color: #7a2e1d;
}

.tag {
  display: inline-block;
  font-size: 0.7em;
  background: #7a2e1d;
  color: #fff3d6;
  padding: 4px 8px;
  border-radius: 6px;
  margin-right: 6px;
  margin-top: 8px;
}

.location-badge {
  display: inline-block;
  font-size: 0.75em;
  padding: 4px 8px;
  border-radius: 6px;
  background: #e9ecef;
  color: #495057;
  margin-left: 6px;
}

.hidden {
  display: none !important;
}

/* ============================================================
   PAGE HEADER
   ============================================================ */

.header {
  text-align: center;
  padding: 3rem 0 3rem 0;
}

.header h1 {
  font-size: 3.2rem;
  font-weight: 700;
  margin-bottom: .6rem;
}

.header p {
  font-size: 1.15rem;
  color: var(--text-light);
  max-width: 850px;
  margin: auto;
}

/* ============================================================
   SEARCH AND FILTER CONTROLS
   ============================================================ */

.controls {
  display: flex;
  gap: 12px;
  margin-bottom: 10px;
  width: 100%;
  align-items: center;
}

#searchBox {
  flex: 7 1 auto;
  min-width: 0;
  width: auto;
  padding: 12px;
  border-radius: 8px;
  border: 1px solid #ddd;
  font-size: 14px;
  box-sizing: border-box;
}

#categoryFilter {
  flex: 0 0 220px;
  width: 220px;
  padding: 12px;
  border-radius: 8px;
  border: 1px solid #ddd;
  box-sizing: border-box;
}
</style>


<div class="header">

<h1>Service Provider Directory</h1>

</div>

This directory shows contact information for support services offered in the El Paso community. To be added or have information updated, please send a request using the [contact form](https://www.epcrimelab.org/contact.html). 


```{r}
library(leaflet)
library(dplyr)
library(readxl)
library(htmltools)

# ============================================================
# LOAD PROVIDER DATA
# ============================================================

providers <- read_excel(
  "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/OCEPTX/service_providers/service_providers_list_ep.xlsx"
)

# Create an ID automatically
providers <- providers %>%
  mutate(id = row_number())

# Clean character fields
providers <- providers %>%
  mutate(
    across(
      c(
        name,
        el_paso,
        services,
        contact_preference,
        phone,
        email,
        website,
        address,
        family_message
      ),
      ~ ifelse(is.na(.), "", as.character(.))
    )
  )

# ============================================================
# HTML ESCAPE FUNCTION
# ============================================================

esc <- function(x) {
  htmlEscape(as.character(x))
}

# ============================================================
# PROVIDER SETTINGS
# ============================================================

providers$tags <- providers$services

providers$website_display <- providers$website

providers$website_url <- ifelse(
  providers$website == "",
  "",
  ifelse(
    grepl("^https?://", providers$website, ignore.case = TRUE),
    providers$website,
    paste0("https://", providers$website)
  )
)

# ============================================================
# MAP POPUPS
# ============================================================

providers$popup <- paste0(
  "<b>", esc(providers$name), "</b><br><br>",
  "<b>Services:</b> ", esc(providers$services), "<br>",
  
  ifelse(
    providers$phone != "",
    paste0(
      "<b>Phone:</b> <a href='tel:",
      gsub("[^0-9]", "", providers$phone),
      "'>",
      esc(providers$phone),
      "</a><br>"
    ),
    ""
  ),
  
  ifelse(
    providers$email != "",
    paste0(
      "<b>Email:</b> <a href='mailto:",
      esc(providers$email),
      "'>",
      esc(providers$email),
      "</a><br>"
    ),
    ""
  ),
  
  ifelse(
    providers$website_url != "",
    paste0(
      "<b>Website:</b> <a href='",
      esc(providers$website_url),
      "' target='_blank'>Visit website</a><br>"
    ),
    ""
  ),
  
  ifelse(
    providers$address != "",
    paste0(
      "<b>Address:</b> ",
      esc(providers$address),
      "<br>"
    ),
    ""
  )
)

# ============================================================
# MAP
# ============================================================

map <- leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-106.485, 31.76, zoom = 11)

mapped_providers <- providers %>%
  filter(
    !is.na(lon),
    !is.na(lat)
  )

for (i in seq_len(nrow(mapped_providers))) {

  map <- map %>%
    addCircleMarkers(
      lng = mapped_providers$lon[i],
      lat = mapped_providers$lat[i],
      layerId = mapped_providers$id[i],
      radius = 8,
      fillOpacity = 0.9,
      popup = mapped_providers$popup[i]
    )
}
```


<div class="section-container directory-container">

<h2>Directory</h2>

<div class="controls">
  <input 
    id="searchBox" 
    type="text" 
    placeholder="Search providers, services, or support..."
  >

  <select id="categoryFilter">
    <option value="all">All Services</option>
    `r paste0(
      "<option value='",
      trimws(unique(providers$services)),
      "'>",
      trimws(unique(providers$services)),
      "</option>",
      collapse = "\n"
    )`
  </select>
</div>

<div id="resultCount" style="margin-bottom:10px; color:#888; font-size:0.9em;"></div>

<div id="directory">

```{r results='asis'}

for (i in seq_len(nrow(providers))) {

  # Contact information
  phone_html <- if (providers$phone[i] != "") {
    paste0(
      "<div><strong>Phone:</strong> ",
      "<a href='tel:",
      gsub("[^0-9]", "", providers$phone[i]),
      "'>",
      esc(providers$phone[i]),
      "</a></div>"
    )
  } else {
    ""
  }

  email_html <- if (providers$email[i] != "") {
    paste0(
      "<div><strong>Email:</strong> ",
      "<a href='mailto:",
      esc(providers$email[i]),
      "'>",
      esc(providers$email[i]),
      "</a></div>"
    )
  } else {
    ""
  }

  website_html <- if (providers$website_url[i] != "") {
    paste0(
      "<div><strong>Website:</strong> ",
      "<a href='",
      esc(providers$website_url[i]),
      "' target='_blank'>",
      esc(providers$website_display[i]),
      "</a></div>"
    )
  } else {
    ""
  }

  # Location
  location_html <- if (
    tolower(trimws(providers$el_paso[i])) == "yes"
  ) {

    address_text <- if (providers$address[i] != "") {
      paste0(
        "<div><strong>Address:</strong> ",
        esc(providers$address[i]),
        "</div>"
      )
    } else {
      ""
    }

    paste0(
      "<div><strong>Location:</strong> El Paso, Texas ",
      "<span class='location-badge'>El Paso-based</span></div>",
      address_text
    )

  } else {

    paste0(
      "<div><strong>Location:</strong> ",
      "Virtual",
      "</div>"
    )
  }

  # Service tags
  service_tags <- paste0(
    "<span class='tag'>",
    trimws(unlist(strsplit(providers$services[i], ","))),
    "</span>",
    collapse = ""
  )

  # Family message
  family_message_html <- if (
    providers$family_message[i] != ""
  ) {

    paste0(
      "<div class='family-message'>",
      "<div class='family-message-label'>",
      "A message to families",
      "</div>",
      esc(providers$family_message[i]),
      "</div>"
    )

  } else {
    ""
  }

  # Searchable text
  searchable <- paste(
    providers$name[i],
    providers$services[i],
    providers$phone[i],
    providers$email[i],
    providers$website[i],
    providers$address[i],
    providers$family_message[i]
  )

  cat(
    "<div class='provider-card' ",
    "data-name='", esc(providers$name[i]), "' ",
    "data-category='", esc(providers$services[i]), "' ",
    "data-search='", esc(searchable), "'>",

    "<div class='provider-name'>",
    "<a href='#' class='provider-link' data-id='",
    providers$id[i],
    "'>",
    "<b>",
    esc(providers$name[i]),
    "</b>",
    "</a>",
    "</div>",

    "<div>",
    service_tags,
    "</div>",

    ifelse(
      providers$blurb[i] != "",
      paste0(
        "<div class='provider-blurb'>",
        esc(providers$blurb[i]),
        "</div>"
      ),
      ""
    ),

    "<div class='provider-info'>",

    location_html,

    "<div><strong>Preferred contact:</strong> ",
    esc(providers$contact_preference[i]),
    "</div>",

    phone_html,
    email_html,
    website_html,

    "</div>",

    family_message_html,

    "</div>\n"
  )
}


```

</div>

</div>

<div class="section-container map-container">

<h2>Map</h2>

``` {r}
map

```

</div>

<script>
document.addEventListener("DOMContentLoaded", function() {

  setTimeout(function() {

    const mapWidget = document.querySelector(".leaflet-container");

    let map = null;

    if (mapWidget) {
      const widget = HTMLWidgets.find("#" + mapWidget.id);

      if (widget) {
        map = widget.getMap();
      }
    }

    const searchBox = document.getElementById("searchBox");
    const categoryFilter = document.getElementById("categoryFilter");
    const resultCount = document.getElementById("resultCount");

    const cards = Array.from(
      document.querySelectorAll(".provider-card")
    );

    function filterCards() {

      const search = (
        searchBox.value || ""
      ).toLowerCase().trim();

      const category = (
        categoryFilter.value || "all"
      ).trim();

      let visible = 0;

      cards.forEach(function(card) {

        const searchable = (
          card.dataset.search || ""
        ).toLowerCase();

        const categories = (
          card.dataset.category || ""
        ).toLowerCase();

        const matchesSearch =
          !search ||
          searchable.includes(search);

        const matchesCategory =
          category === "all" ||
          categories.includes(category.toLowerCase());

        if (matchesSearch && matchesCategory) {

          card.classList.remove("hidden");
          visible++;

        } else {

          card.classList.add("hidden");

        }

      });

      resultCount.textContent =
        visible + (visible === 1 ? " result" : " results");
    }

    searchBox.addEventListener(
      "input",
      filterCards
    );

    categoryFilter.addEventListener(
      "change",
      filterCards
    );

    filterCards();


    /*
      Directory → Map
    */

    document.querySelectorAll(".provider-link").forEach(
      function(link) {

        link.addEventListener(
          "click",
          function(e) {

            e.preventDefault();

            const id = Number(
              this.dataset.id
            );

            if (!map) return;

            map.eachLayer(function(layer) {

              if (
                layer.options &&
                Number(layer.options.layerId) === id &&
                typeof layer.getLatLng === "function"
              ) {

                map.setView(
                  layer.getLatLng(),
                  15
                );

                layer.openPopup();

              }

            });

          }
        );

      }
    );

  }, 800);

});
</script>

