---
title: "Service Provider Directory"
subtitle: "El Paso, Texas"
format:
  html:
    embed-resources: true
    page-layout: full
    toc: true
    toc-depth: 2
    toc-title: "On this page"
execute:
  echo: false
  warning: false
  message: false
---

<style>

/* ============================================================
   EL PASO CRIME LAB — DESERT COLOR PALETTE
   ============================================================ */

:root {
  --epl-rust: #7a2f1d;
  --epl-rust-dark: #562014;
  --epl-terracotta: #a7462c;
  --epl-sand: #c8ad87;
  --epl-sand-light: #e5d3b9;
  --epl-cream: #fffaf2;
  --epl-cream-dark: #ead7bb;
  --epl-brown: #2d1c16;
  --epl-brown-light: #4e3d33;
  --epl-desert: #a97847;
  --epl-sage: #4e563d;
}


/* ============================================================
   GLOBAL
   ============================================================ */

body {
  background:
    linear-gradient(
      180deg,
      #e5d3b9 0%,
      #f5e9d8 32%,
      #fffaf2 100%
    );

  color: var(--epl-brown);
}

.quarto-title-block {
  display: none;
}

main.content {
  max-width: 1500px;
  margin: 0 auto;
  padding: 2.5rem 3rem 4rem 3rem;
}


/* ============================================================
   PAGE HEADER
   ============================================================ */

.header {
  padding: 2.5rem 0 2rem 0;
  margin-bottom: 1rem;
}

.header h1 {
  font-size: 3rem;
  font-weight: 750;
  letter-spacing: -0.04em;
  line-height: 1.1;
  margin: 0 0 0.8rem 0;
  color: var(--epl-brown);
}

.header h1::before {
  content: "EL PASO CRIME LAB";
  display: block;

  font-size: 0.78rem;
  font-weight: 700;

  letter-spacing: 0.14em;

  color: var(--epl-rust);

  margin-bottom: 0.6rem;
}

.header h1::after {
  content: "";
  display: block;

  width: 55px;
  height: 4px;

  background: var(--epl-terracotta);

  border-radius: 10px;

  margin-top: 1rem;
}


/* ============================================================
   INTRODUCTION
   ============================================================ */

.header + p {
  max-width: 850px;

  font-size: 1.05rem;
  line-height: 1.7;

  color: var(--epl-brown-light);

  margin: 0 0 2rem 0;
}

.header + p a {
  color: var(--epl-rust);
  font-weight: 600;
}


/* ============================================================
   SECTION CONTAINERS
   ============================================================ */

.section-container {
  background: rgba(255, 250, 242, 0.96);

  border: 1px solid var(--epl-sand);

  border-radius: 18px;

  padding: 1.5rem;

  margin: 1.5rem 0;

  box-shadow:
    0 8px 25px rgba(59, 42, 34, 0.08);
}

.section-container h2 {
  margin: 0 0 1.2rem 0;

  font-size: 1.25rem;
  font-weight: 700;

  letter-spacing: -0.02em;

  color: var(--epl-brown);
}


/* ============================================================
   DIRECTORY CONTAINER
   ============================================================ */

.directory-container {
  background:
    linear-gradient(
      135deg,
      #fffaf2 0%,
      #f0ddc3 100%
    );

  border-color: #d8c3a5;
}


/* ============================================================
   SEARCH + FILTER
   ============================================================ */

.controls {
  display: grid;

  grid-template-columns: 1fr 240px;

  gap: 12px;

  margin-bottom: 12px;

  width: 100%;

  align-items: center;
}

#searchBox,
#categoryFilter {
  height: 46px;

  padding: 0 14px;

  border-radius: 10px;

  border: 1px solid #af8d65;

  background: #fffaf2;

  color: var(--epl-brown);

  font-size: 14px;

  box-sizing: border-box;

  outline: none;

  transition: all 0.2s ease;
}

#searchBox::placeholder {
  color: #765f4e;
}

#searchBox:focus,
#categoryFilter:focus {
  border-color: var(--epl-rust);

  box-shadow:
    0 0 0 3px rgba(139, 58, 36, 0.12);

  background: #fffdf9;
}

#categoryFilter {
  cursor: pointer;
}

#resultCount {
  margin: 0 0 14px 2px !important;

  color: #5c4638 !important;

  font-size: 0.82rem !important;
}


/* ============================================================
   PROVIDER CARDS
   ============================================================ */

.provider-card {
  position: relative;

  padding: 20px 22px;

  margin-bottom: 12px;

  background: #fffaf2;

  border: 1px solid #c6a47c;

  border-radius: 13px;

  transition:
    transform 0.2s ease,
    box-shadow 0.2s ease,
    border-color 0.2s ease;
}

.provider-card:hover {
  transform: translateY(-3px);

  background: #fffdf9;

  box-shadow:
    0 10px 25px rgba(59, 42, 34, 0.12);

  border-color: #c99a6b;
}


/* ============================================================
   PROVIDER NAME
   ============================================================ */

.provider-name {
  font-size: 1.1rem;

  line-height: 1.3;

  margin-bottom: 4px;
}

.provider-name a {
  color: var(--epl-brown);

  text-decoration: none;

  transition: color 0.2s ease;
}

.provider-name a:hover {
  color: var(--epl-rust);
}


/* ============================================================
   SERVICE TAGS
   ============================================================ */

.tag {
  display: inline-block;

  font-size: 0.68rem;

  font-weight: 650;

  line-height: 1;

  background: var(--epl-rust);

  color: #fff5e8;

  padding: 6px 9px;

  border-radius: 6px;

  margin-right: 5px;

  margin-top: 7px;

  box-shadow:
    0 2px 5px rgba(139, 58, 36, 0.12);
}


/* ============================================================
   PROVIDER INFORMATION
   ============================================================ */

.provider-info {
  font-size: 13px;

  line-height: 1.7;

  margin-top: 13px;

  color: #47352b;
}

.provider-info > div {
  margin-bottom: 2px;
}

.provider-info strong {
  color: var(--epl-brown);

  font-weight: 650;
}

.provider-info a {
  color: var(--epl-rust);

  text-decoration: none;

  font-weight: 500;
}

.provider-info a:hover {
  color: var(--epl-rust-dark);

  text-decoration: underline;
}


/* ============================================================
   LOCATION
   ============================================================ */

.location-badge {
  display: inline-block;

  font-size: 0.72em;

  font-weight: 650;

  padding: 4px 8px;

  border-radius: 6px;

  background: #d7dbc3;

  color: #37412a;

  margin-left: 5px;
}


/* ============================================================
   FAMILY MESSAGE
   ============================================================ */

.family-message {
  margin-top: 15px;

  padding: 14px 16px;

  background:
    linear-gradient(
      135deg,
      #f7e7cf,
      #efdbbd
    );

  border-left: 4px solid var(--epl-terracotta);

  border-radius: 7px;

  font-size: 13px;

  line-height: 1.6;

  color: var(--epl-brown-light);
}

.family-message-label {
  font-weight: 700;

  margin-bottom: 4px;

  color: var(--epl-rust);
}


/* ============================================================
   MAP CONTAINER
   ============================================================ */

.map-container {
  background:
    linear-gradient(
      135deg,
      #d1ad82 0%,
      #af7c4d 100%
    );

  border: 1px solid #b9946b;

  padding: 1.5rem;

  box-shadow:
    0 10px 30px rgba(59, 42, 34, 0.12);
}

.map-container h2 {
  color: var(--epl-brown);

  margin-bottom: 1rem;
}


/* ============================================================
   MAP
   ============================================================ */

.map-container .leaflet-container {
  width: 100% !important;

  height: 600px !important;

  border-radius: 13px;

  overflow: hidden;

  border: 2px solid rgba(255, 250, 242, 0.8);

  box-shadow:
    0 6px 20px rgba(59, 42, 34, 0.18);
}


/* ============================================================
   LEAFLET CONTROLS
   ============================================================ */

.map-container .leaflet-control-layers {
  border: none !important;

  border-radius: 9px !important;

  box-shadow:
    0 5px 18px rgba(59, 42, 34, 0.18) !important;

  background: #fffaf2 !important;
  color: var(--epl-brown) !important;
}

.map-container .leaflet-control-layers-expanded {
  padding: 8px 10px !important;
}

.map-container .leaflet-control-layers label {
  color: var(--epl-brown) !important;
  font-size: 0.82rem;
  font-weight: 600;
}

.map-container .leaflet-control-zoom {
  border: none !important;

  box-shadow:
    0 5px 18px rgba(59, 42, 34, 0.18) !important;

  border-radius: 9px !important;

  overflow: hidden;
}

.map-container .leaflet-control-zoom a {
  color: var(--epl-brown) !important;
}


/* ============================================================
   MAP POPUPS
   ============================================================ */

.leaflet-popup-content-wrapper {
  border-radius: 10px !important;
}

.leaflet-popup-content {
  color: var(--epl-brown);

  font-size: 13px;

  line-height: 1.6;
}

.leaflet-popup-content a {
  color: var(--epl-rust);

  text-decoration: none;
}


/* ============================================================
   HIDDEN
   ============================================================ */

.hidden {
  display: none !important;
}


/* ============================================================
   RESPONSIVE
   ============================================================ */

@media (max-width: 900px) {

  main.content {
    padding: 1.5rem;
  }

  .header h1 {
    font-size: 2.4rem;
  }

  .controls {
    grid-template-columns: 1fr;
  }

  #categoryFilter {
    width: 100%;
  }

  .map-container .leaflet-container {
    height: 500px !important;
  }

}


@media (max-width: 600px) {

  main.content {
    padding: 1rem;
  }

  .header {
    padding-top: 1rem;
  }

  .header h1 {
    font-size: 2rem;
  }

  .section-container {
    padding: 1rem;

    border-radius: 13px;
  }

  .provider-card {
    padding: 17px;
  }

  .map-container .leaflet-container {
    height: 450px !important;
  }

}

</style>



<div class="header">

<h1>Service Provider Directory</h1>

</div>

Find local services here. To be added, removed, or have information changed, please send a request using the [contact form](https://www.epcrimelab.org/contact.html). 


```{r}
library(leaflet)
library(dplyr)
library(readxl)
library(htmltools)

# ============================================================
# LOAD PROVIDER DATA
# ============================================================

providers <- read_excel(
  "C:/Users/Matt/ASU Dropbox/Matthew Duran/Professional Development/Website/OCEPTX/service_providers/service_providers_list_ep.xlsx"
)

# Create an ID automatically
providers <- providers %>%
  mutate(id = row_number())

# Clean character fields
providers <- providers %>%
  mutate(
    across(
      c(
        name,
        el_paso,
        services,
        contact_preference,
        phone,
        email_1,
        email_1_description,
        email_2,
        email_2_description,
        website,
        address,
        family_message
      ),
      ~ ifelse(is.na(.), "", as.character(.))
    )
  )

# ============================================================
# HTML ESCAPE FUNCTION
# ============================================================

esc <- function(x) {
  htmlEscape(as.character(x))
}

# ============================================================
# PROVIDER SETTINGS
# ============================================================

providers$tags <- providers$services

providers$website_display <- providers$website

providers$website_url <- ifelse(
  providers$website == "",
  "",
  ifelse(
    grepl("^https?://", providers$website, ignore.case = TRUE),
    providers$website,
    paste0("https://", providers$website)
  )
)

# ============================================================
# MAP POPUPS
# ============================================================

providers$popup <- paste0(
  "<b>", esc(providers$name), "</b><br><br>",
  "<b>Services:</b> ", esc(providers$services), "<br>",
  
  ifelse(
    providers$phone != "",
    paste0(
      "<b>Phone:</b> <a href='tel:",
      gsub("[^0-9]", "", providers$phone),
      "'>",
      esc(providers$phone),
      "</a><br>"
    ),
    ""
  ),
  
  ifelse(
    providers$email != "",
    paste0(
      "<b>Email:</b> <a href='mailto:",
      esc(providers$email),
      "'>",
      esc(providers$email),
      "</a><br>"
    ),
    ""
  ),
  
  ifelse(
    providers$website_url != "",
    paste0(
      "<b>Website:</b> <a href='",
      esc(providers$website_url),
      "' target='_blank'>Visit website</a><br>"
    ),
    ""
  ),
  
  ifelse(
    providers$address != "",
    paste0(
      "<b>Address:</b> ",
      esc(providers$address),
      "<br>"
    ),
    ""
  )
)

# ============================================================
# MAP
# ============================================================

map <- leaflet() %>%

  # Street map with street names — default
  addTiles(
    group = "Street Map"
  ) %>%

  # Clean/light background
  addProviderTiles(
    "CartoDB.Positron",
    group = "Light"
  ) %>%

  # Satellite imagery
  addProviderTiles(
    "Esri.WorldImagery",
    group = "Satellite"
  ) %>%

  # Basemap toggle; Street Map remains visible when the page opens.
  addLayersControl(
    baseGroups = c("Street Map", "Light", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%

  hideGroup(c("Light", "Satellite")) %>%

  setView(
    -106.485,
    31.76,
    zoom = 11
  )

mapped_providers <- providers %>%
  filter(
    !is.na(lon),
    !is.na(lat)
  )

for (i in seq_len(nrow(mapped_providers))) {

  map <- map %>%
    addCircleMarkers(
      lng = mapped_providers$lon[i],
      lat = mapped_providers$lat[i],
      layerId = mapped_providers$id[i],
      radius = 8,
      fillOpacity = 0.9,
      popup = mapped_providers$popup[i]
    )
}
```


<div class="section-container directory-container">

<div class="controls">
  <input 
    id="searchBox" 
    type="text" 
    placeholder="Search providers, services, or support..."
  >

<select id="categoryFilter">
  <option value="all">All Services</option>
  `r paste0(
    "<option value='",
    trimws(
      unique(
        unlist(
          strsplit(
            providers$services,
            ","
          )
        )
      )
    ),
    "'>",
    trimws(
      unique(
        unlist(
          strsplit(
            providers$services,
            ","
          )
        )
      )
    ),
    "</option>",
    collapse = "\n"
  )`
</select>
</div>

<div id="resultCount" style="margin-bottom:10px; color:#888; font-size:0.9em;"></div>

<div id="directory">

```{r results='asis'}

for (i in seq_len(nrow(providers))) {

  # Contact information
  phone_html <- if (providers$phone[i] != "") {
    paste0(
      "<div><strong>Phone:</strong> ",
      "<a href='tel:",
      gsub("[^0-9]", "", providers$phone[i]),
      "'>",
      esc(providers$phone[i]),
      "</a></div>"
    )
  } else {
    ""
  }

# ============================================================
# EMAIL
# ============================================================

email_items <- character(0)

if (providers$email_1[i] != "") {

  email_items <- c(
    email_items,
    paste0(
      "<span style='margin-right:12px;'>",
      "<a href='mailto:",
      esc(providers$email_1[i]),
      "'>",
      esc(providers$email_1[i]),
      "</a>",
      if (providers$email_1_description[i] != "") {
        paste0(
          " <span style='color:#777;'>— ",
          esc(providers$email_1_description[i]),
          "</span>"
        )
      } else {
        ""
      },
      "</span>"
    )
  )
}

if (providers$email_2[i] != "") {

  email_items <- c(
    email_items,
    paste0(
      "<span style='margin-right:12px;'>",
      "<a href='mailto:",
      esc(providers$email_2[i]),
      "'>",
      esc(providers$email_2[i]),
      "</a>",
      if (providers$email_2_description[i] != "") {
        paste0(
          " <span style='color:#777;'>— ",
          esc(providers$email_2_description[i]),
          "</span>"
        )
      } else {
        ""
      },
      "</span>"
    )
  )
}

email_html <- if (length(email_items) > 0) {
  paste0(
    "<div><strong>Email:</strong> ",
    paste0(email_items, collapse = ""),
    "</div>"
  )
} else {
  ""
}

  website_html <- if (providers$website_url[i] != "") {
    paste0(
      "<div><strong>Website:</strong> ",
      "<a href='",
      esc(providers$website_url[i]),
      "' target='_blank'>",
      esc(providers$website_display[i]),
      "</a></div>"
    )
  } else {
    ""
  }

  # Location
  location_html <- if (
    tolower(trimws(providers$el_paso[i])) == "yes"
  ) {

    address_text <- if (providers$address[i] != "") {
      paste0(
        "<div><strong>Address:</strong> ",
        esc(providers$address[i]),
        "</div>"
      )
    } else {
      ""
    }

    paste0(
      "<div><strong>Location:</strong> El Paso, Texas ",
      "<span class='location-badge'>El Paso-based</span></div>",
      address_text
    )

  } else {

    paste0(
      "<div><strong>Location:</strong> ",
      "Virtual",
      "</div>"
    )
  }

  # Service tags
  service_tags <- paste0(
    "<span class='tag'>",
    trimws(unlist(strsplit(providers$services[i], ","))),
    "</span>",
    collapse = ""
  )

  # Family message
  family_message_html <- if (
    providers$family_message[i] != ""
  ) {

    paste0(
      "<div class='family-message'>",
      "<div class='family-message-label'>",
      "A message to families",
      "</div>",
      esc(providers$family_message[i]),
      "</div>"
    )

  } else {
    ""
  }

  # Searchable text
searchable <- paste(
  providers$name[i],
  providers$services[i],
  providers$phone[i],
  providers$email_1[i],
  providers$email_1_description[i],
  providers$email_2[i],
  providers$email_2_description[i],
  providers$website[i],
  providers$address[i],
  providers$family_message[i]
)

  cat(
    "<div class='provider-card' ",
    "data-name='", esc(providers$name[i]), "' ",
    "data-category='", esc(providers$services[i]), "' ",
    "data-search='", esc(searchable), "'>",

    "<div class='provider-name'>",
    "<a href='#' class='provider-link' data-id='",
    providers$id[i],
    "'>",
    "<b>",
    esc(providers$name[i]),
    "</b>",
    "</a>",
    "</div>",

    "<div>",
    service_tags,
    "</div>",

    "<div class='provider-info'>",

    location_html,

    "<div><strong>Preferred contact:</strong> ",
    esc(providers$contact_preference[i]),
    "</div>",

    phone_html,
    email_html,
    website_html,

    "</div>",

    family_message_html,

    "</div>\n"
  )
}


```

</div>

</div>

<div class="section-container map-container">

<h2>Map</h2>

``` {r}
map

```

</div>

<script>
document.addEventListener("DOMContentLoaded", function() {

  setTimeout(function() {

    const mapWidget = document.querySelector(".leaflet-container");

    let map = null;

    if (mapWidget) {
      const widget = HTMLWidgets.find("#" + mapWidget.id);

      if (widget) {
        map = widget.getMap();
      }
    }

    const searchBox = document.getElementById("searchBox");
    const categoryFilter = document.getElementById("categoryFilter");
    const resultCount = document.getElementById("resultCount");

    const cards = Array.from(
      document.querySelectorAll(".provider-card")
    );

    function filterCards() {

      const search = (
        searchBox.value || ""
      ).toLowerCase().trim();

      const category = (
        categoryFilter.value || "all"
      ).trim();

      let visible = 0;

      cards.forEach(function(card) {

        const searchable = (
          card.dataset.search || ""
        ).toLowerCase();

        const categories = (
          card.dataset.category || ""
        ).toLowerCase();

        const matchesSearch =
          !search ||
          searchable.includes(search);

        const serviceList = categories
          .split(",")
          .map(function(service) {
            return service.trim();
          });
        
        const matchesCategory =
          category === "all" ||
          serviceList.includes(category.toLowerCase());

        if (matchesSearch && matchesCategory) {

          card.classList.remove("hidden");
          visible++;

        } else {

          card.classList.add("hidden");

        }

      });

      resultCount.textContent =
        visible + (visible === 1 ? " result" : " results");
    }

    searchBox.addEventListener(
      "input",
      filterCards
    );

    categoryFilter.addEventListener(
      "change",
      filterCards
    );

    filterCards();


    /*
      Directory → Map
    */

    document.querySelectorAll(".provider-link").forEach(
      function(link) {

        link.addEventListener(
          "click",
          function(e) {

            e.preventDefault();

            const id = Number(
              this.dataset.id
            );

            if (!map) return;

            map.eachLayer(function(layer) {

              if (
                layer.options &&
                Number(layer.options.layerId) === id &&
                typeof layer.getLatLng === "function"
              ) {

                map.setView(
                  layer.getLatLng(),
                  15
                );

                layer.openPopup();

              }

            });

          }
        );

      }
    );

  }, 800);

});
</script>

